---
title: "Decomposition of Concentration-Index Trees"
author: "Moses Mburu"
format:
  pdf:
    documentclass: article
    papersize: a4
    fontsize: 11pt
    geometry:
      - margin=25mm
    pdf-engine: xelatex
    keep-tex: true
    toc: false
    number-sections: true
    number-depth: 3
    fig-cap-location: bottom
    tbl-cap-location: top
    colorlinks: true
    fig-pos: "H"
    tbl-pos: "H"
    fig-dpi: 500
    include-in-header:
      text: |
        \renewcommand{\baselinestretch}{1.1}
        \usepackage{microtype}
        \frenchspacing
        \usepackage{graphicx}
        \usepackage{xcolor}
        \usepackage{amsmath,amssymb}
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{multirow}
        \usepackage{makecell}
        \usepackage{pdflscape}
        \usepackage{float}
        \usepackage{fancyhdr}
execute:
  echo: false
  warning: false
  message: false
  fig-dpi: 700
params:
  render_appendix_tree_plots: true
bibliography: ref/thesis_ref.bib
csl: ref/harvard.csl
citeproc: true
cite-method: citeproc
link-citations: true
reference-location: block
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 10,
  fig.height = 6,
  dpi = 500,
  dev = "pdf",
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(gridExtra)
library(data.table)
library(ggplot2)
library(grid)
library(knitr)
library(ineqTrees)
library(partykit)
library(here)
library(kableExtra)
use_latex_packages()
source(here("R", "report_helpers.R"))
source(here("R", "ci_rpart_comparison_methods.R"))

if (!exists("params", inherits = FALSE)) {
  params <- list(render_appendix_tree_plots = TRUE)
}


```

```{r load-results, include=FALSE}
results_object_file <- here("data", "drc_report_results_objects.rda")
report_results <- load_drc_report_results(results_object_file)
list2env(report_results, envir = environment())
var_labels <- str_wrap(congo_var_labels, width = 25)

simulation_plot_file <- here("simulation", "simulation_tree_plots.rda")
if (file.exists(simulation_plot_file)) {
  load(simulation_plot_file)
} else {
  simulation_tree_plots <- list()
  simulation_tree_plot_metadata <- list()
}

```


# Results

## Exploratory Data Analysis

In the DRC Data Set[@dhs_drc_2007_dataset], 10.32% of children died before reaching age five (@fig-results-wealth-distribution). Deaths were more commonly observed among children from disadvantaged or rural backgrounds. The children who died differed descriptively from those who survived. Low-wealth households accounted for a larger share of deaths than of survivors (51.3% versus 42.8%). The same pattern was present for rural residence (69.3% versus 60.7%), no skilled attendance at birth (65.6% versus 56.8%), and mothers with no education (67.5% versus 55.2%) (@tbl-appendix-weighted-baseline). Mortality also differed across provinces, ranging from 6.2% in Kinshasa to 14.2% in Katanga (@fig-appendix-province-mortality-distribution).  Next we use the direct and indirect methods for decomposition.

```{r results-outcome-distribution, fig.width=6.5, fig.height=3}
#| label: fig-results-outcome-distribution
#| fig-cap: "Distribution of under-five mortality in the DRC analysis data."
outcome_gg <- ggplot(
    drc_results_objects$outcome_distribution,
    aes(x = outcome_label, y = weighted_percent)
) +
    geom_col(fill = "#3268A8", width = 0.58) +
    geom_text(aes(label = sprintf("%.1f%%", weighted_percent)),
        vjust = -0.35, size = 3.3
    ) +
    labs(x = NULL, y = "Weighted percent") +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank())


```

```{r results-wealth-distribution, fig.width=6, fig.height=3}
#| label: fig-results-wealth-distribution
#| fig-cap: "Under-five mortality distribution and DHS wealth index in the DRC analysis data."
wealth_by_outcome_dt <- copy(congo_model_dt)
wealth_by_outcome_dt[, outcome_label := fifelse(
    deadu5_num == 1,
    outcome_labels[["died"]],
    outcome_labels[["alive"]]
)]

wealth_distr <- ggplot(
    wealth_by_outcome_dt,
    aes(x = wealth, weight = sample_weight)
) +
    geom_histogram(
        bins = 35, fill = "#3268A8",
        color = "white", linewidth = 0.2
    ) +
    # facet_wrap(~ outcome_label, ncol = 1, scales = "free_y") +
    labs(x = "Wealth index", y = "Weighted count") +
    theme_minimal(base_size = 11) +
    theme(
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold")
    )
grid.arrange(outcome_gg, wealth_distr, ncol = 2)
```



## Decomposition of determinants of under-five mortality

### Indirect methods comparison (regression and predictive-forest SHAP)

From the indirect decomposition, we see that the largest regression contribution came from mother’s education: 23.1% under CI and 28.7% under CIg/CIc (@tbl-results-method-comparison). Father’s occupation in agriculture was second, contributing 17.9% under CI and 19.9% under CIg/CIc. In the predictive-forest SHAP decomposition, the contributions were less concentrated in one or two variables. Residence and household wealth each contributed about 12% under CI, CIg and CIc, followed by mother’s occupation, father’s occupation, mother’s education and skilled birth attendance. The fitted regression model is reported in @tbl-appendix-regression-model-summary. Details of the fitted predictive forest are provided in @tbl-appendix-ranger-sampling-summary and @tbl-appendix-ranger-cross-validation, and its SHAP contributions are shown in @fig-appendix-shap-summary.

```{r results-method-comparison}
#| label: tbl-results-method-comparison
#| tbl-cap: "Regression and predictive-forest SHAP percentage contributions by criterion."
method_comparison_report <- copy(drc_results_objects$method_comparison[
    ,
    .(
        variable = variable_label,
        regression_ci_pct_contribution,
        regression_cig_pct_contribution,
        regression_cic_pct_contribution,
        forest_shap_ci_pct_contribution,
        forest_shap_cig_pct_contribution,
        forest_shap_cic_pct_contribution,
        forest_shap_l_pct_contribution
    )
])

setnames(
    method_comparison_report,
    c(
        "variable",
        "regression_ci_pct_contribution",
        "regression_cig_pct_contribution",
        "regression_cic_pct_contribution",
        "forest_shap_ci_pct_contribution",
        "forest_shap_cig_pct_contribution",
        "forest_shap_cic_pct_contribution",
        "forest_shap_l_pct_contribution"
    ),
    c(
        "Variable",
        "CI (%)",
        "CIg (%)",
        "CIc (%)",
        "CI (%)",
        "CIg (%)",
        "CIc (%)",
        "L (%)"
    )
)

method_comparison_table <- report_compact_table(
    method_comparison_report,
    digits = 3,
    align = c("l", rep("r", 7L)),
    column_widths = c("4.4cm", rep("1.35cm", 7L))
)


method_comparison_table <- kableExtra::add_header_above(
    method_comparison_table,
    c(" " = 1, "Regression" = 3, "Predictive-forest SHAP" = 4),
    bold = TRUE
)


method_comparison_table
```

## Direct methods (concentration-index trees and forests)

### Rank Dependent Concentration-Index Tree 

#### The Corrected Concentration-Index Tree $CIc$


We start with a tree based on the corrected concentration index, appropriate for bounded outcomes[@erreygers2009correcting]. In this method, the tree shows that inequality in under-five mortality is structured primarily by maternal education, father’s agricultural occupation, household wealth, skilled birth attendance, and residence in Katanga [@fig-appendix-ci-tree-cic]. The lowest-mortality terminal group consists of children whose mothers had some education and whose fathers were not in agriculture, with an under-five mortality rate of about 6.5%. At the opposite end, children whose mothers had no education, who lived in low-wealth households, and who resided in Katanga had the highest mortality rate, about 17.8%. Thus, the corrected rank-dependent tree identifies a clear socioeconomic gradient, but one that is expressed through combinations of education, occupation, household wealth, delivery care, and region rather than through wealth alone.



```{r fig.width=9, fig.height=4}
#| label: fig-appendix-ci-tree-cic
#| fig-cap: "Concentration-index tree fitted with the CIc criterion."
#| fig-pos: "H"


  ci_report_tree_plot(
    fit = tree_models_by_type[["CIc"]],
    data = congo_model_dt,
    outcome_name = "deadu5_num",
    outcome_label = "U5",
    ci_type = "CIc",
    var_labels = congo_var_labels,
    fontsize = 11
  )

```

#### Generalised and standard concentration-index trees $CIg$ and $CI$


The appendix also shows the trees fitted using the generalised and standard concentration indices. The generalised concentration-index tree is almost identical to the corrected concentration-index tree [@fig-appendix-ci-tree-cig]. It again separates children first by maternal education and father’s agricultural occupation, then by skilled birth attendance, household wealth, and residence in Katanga. The mortality pattern is therefore substantively unchanged: the lowest-risk group consists of children whose mothers had some education and whose fathers were not in agriculture, while the highest-risk group consists of children whose mothers had no education, who lived in low-wealth households, and who resided in Katanga, with mortality around 17.8% [@fig-appendix-ci-tree-cig]. The standard concentration-index tree gives a similar structure. However, it adds an extra distinction among children whose mothers had some education and whose fathers were not in agriculture, splitting them by short preceding birth interval [@fig-appendix-ci-tree-ci]. Children in this otherwise advantaged group without a short birth interval have the lowest mortality, about 6.0%. In contrast, those with a short birth interval have higher mortality, about 11.7% [@fig-appendix-ci-tree-ci]. Apart from this additional demographic split, the standard CI tree points to the same main sources of inequality as the corrected and generalised trees: maternal education, father’s agricultural occupation, delivery care, household wealth, and residence in Katanga.



### Level dependent Concentration-Index Tree $L$

We then fit a tree based on the level-dependent concentration index [@erreygers2017socioeconomic] criterion, $L$. In this method, the selected tree shows that inequality in under-five mortality is structured primarily by rural residence, father’s agricultural occupation, household wealth, and province, especially Kinshasa, Equateur, and Katanga [@fig-results-selected-ci-tree]. The lowest-mortality terminal group consists of urban children in Kinshasa whose fathers were not in agriculture, with an under-five mortality rate of about 5.1%. The highest-mortality terminal group consists of rural children living in low-wealth households in Katanga, with an under-five mortality rate of about 18.5%. Thus, the level-dependent tree tells a more spatial story of inequality: mortality risk is first organised by the rural-urban divide, then by occupation, wealth, and region. This shows that the level-dependent method found a different structure compared to the rank-dependent methods


```{r results-selected-tree, fig.width=9, fig.height=4.5, out.width="100%"}
#| label: fig-results-selected-ci-tree
#| fig-cap: "Level dependent concentration-index tree for under-five mortality in the DRC."
#| fig-pos: "H"

ci_report_tree_plot(
  fit = selected_tree_fit,
  data = congo_model_dt,
  outcome_name = "deadu5_num",
  outcome_label = "U5",
  ci_type = selected_tree_type,
  var_labels = congo_var_labels
)
```



### Tree validation plots

The rank-dependent trees performed well on the training data but poorly on the validation data. The CI, CIc, and CIg trees removed approximately 87 to 88% of root-node inequality during training, but had negative validation gains at the selected settings: -75.5% for CI and -73.1% for both CIc and CIg [@fig-appendix-tree-validation-path; @tbl-appendix-tree-selected-setting-summary]. In this criterion, a negative validation gain means that the sum of the terminal-node concentration-index values in the validation fold exceeded the concentration-index value at the root node. Thus, although these trees identified substantial rank-based structure in the fitted sample, the same partitions did not reduce inequality when applied to held-out folds. Unlike rank-dependent trees, the level-dependent tree maintained a positive validation gain throughout the tuning path. At the selected setting, it removed 70.5% of root-node inequality in held-out folds while retaining an average of 6.8 terminal nodes [@fig-appendix-tree-validation-path; @tbl-appendix-tree-selected-setting-summary]. The rural-urban, occupational, wealth, and regional partition identified by the L tree was therefore more stable under the validation criterion used in this thesis.

```{r appendix-tree-validation-path-plots, fig.width= 6.3, fig.height=3}
#| label: fig-appendix-tree-validation-path
#| fig-cap: "Tree validation gain and complexity by minimum relative split gain."
#| fig-pos: "H"
tree_path_plot_dt <- copy(drc_results_objects$tree_min_relative_gain_path %||% data.table())

plot_validation_percent_gain_path(
    tree_path_plot_dt,
    title = "Tree Validation Gain (% Root)"
)
```


#### Tree complexity plots

Tree complexity decreased as the minimum relative split gain increased, since larger thresholds allowed only splits that removed a larger share of parent-node impurity [@fig-appendix-tree-complexity-path]. The selected trees remained small, with an average of 7.5 terminal nodes for CI, 7.3 for CIc and CIg, and 6.8 for L [@tbl-appendix-tree-selection; @fig-appendix-tree-complexity-path].


```{r}
#| label: fig-appendix-tree-complexity-path
#| fig-cap: "Tree complexity by number of terminal nodes."
#| fig-pos: "H"
#| fig-width: 6.3
#| fig-height: 3
#| 
plot_terminal_nodes_path(
    tree_path_plot_dt,
    title = "Tree Complexity"
)
```

#### Rpart comparison

The rpart comparison tree recovered the same upper-level structure as the main CI tree (@fig-appendix-ci-tree-ci; @fig-appendix-lecturer-rpart-robust). Both trees first split on maternal education, then separated children by father’s agricultural occupation among those whose mothers had some education, and by household wealth among those whose mothers had no education. The high-mortality branch was also similar, identifying children whose mothers had no education, who lived in low-wealth households, and who resided in Katanga. The main difference was in the lower branches: the rpart comparison introduced a split by long preceding birth interval, while the main CI tree used a short preceding birth-interval split. This suggests that the main rank-dependent structure was not driven only by the custom implementation, although some lower-level refinements depended on the tree-growing and stopping rules.

```{r appendix-lecturer-rpart-robust, fig.width=9, fig.height=5}
#| label: fig-appendix-lecturer-rpart-robust
#| fig-cap: "Rpart comparison tree using CI impurity."
#| fig-pos: "H"
if (isTRUE(params$render_appendix_tree_plots) &&
    "robust_abs" %in% names(lecturer_rpart_models)) {
    plot_lecturer_rpart("robust_abs",
        fontsize = 11,
        outcome_label = "U5",
        var_labels = var_labels
    )
}
```



### Concentration index forest and surrogate tree

#### Forest validation plots

We evaluated concentration-index forests using the same validation criterion as the single trees. For the forests, the minimum relative split gain was tuned to 0.35, since larger values produced near-root-only forests with mean terminal-node counts close to 1. At that point, further validation failed to reveal a meaningful structure.
For the rank-dependent criteria, the forests reduced overfitting but did not stabilise the partitions. The CI, CIc, and CIg forests removed 29.6%, 38.5%, and 38.7% of root-node inequality in training, compared with about 87 to 88% for the corresponding single trees [@tbl-appendix-forest-selected-setting-summary; @tbl-appendix-tree-selected-setting-summary]. Their validation gains remained negative: -25.0% for CI, -28.1% for CIc, and -34.7% for CIg [@fig-appendix-forest-validation-path; @tbl-appendix-forest-selected-setting-summary]. Thus, averaging across trees reduced the validation loss, but did not produce a reproducible rank-dependent subgroup structure.
The tuning path shows the trade-off. Validation gain improved as the minimum relative split gain increased, but the component trees also became shallower [@fig-appendix-forest-validation-path; @fig-appendix-forest-complexity-path]. At the selected setting, the mean terminal-node counts were 2.4 for CI, 2.9 for CIc, and 2.9 for CIg [@tbl-appendix-forest-selection]. 
The level-dependent forest remained more stable. It removed 95.1% of root-node inequality in training and 69.7% in validation [@tbl-appendix-forest-selected-setting-summary]. The selected L forest had an average of 4.1 terminal nodes per tree, compared with 6.8 for the selected L tree [@tbl-appendix-forest-selection; @tbl-appendix-tree-selection]. 

```{r appendix-forest-validation-path-plots, fig.width=6.3, fig.height=3.8}
#| label: fig-appendix-forest-validation-path
#| fig-cap: "Forest validation gain and complexity by minimum relative split gain."
#| fig-pos: "H"

forest_path_plot_dt <- copy(drc_results_objects$forest_min_relative_gain_path)

plot_validation_percent_gain_path(
  forest_path_plot_dt,
  title = "Forest Validation Gain (% Root)"
)
```


####  Surrogate trees from Forest Predictions


To interpret the selected forest, we fitted a surrogate tree to summarise its predictions (@fig-results-surrogate-tree). The surrogate tree retained the main structure seen in the level-dependent tree. 


```{r  fig.width=9, fig.height=5, out.width="100%"}
#| label: fig-results-surrogate-tree
#| fig-cap: "Surrogate tree summarising the selected concentration-index forest predictions."
#| fig-pos: "H"
selected_forest_surrogate <- drc_results_objects$selected_forest_surrogate


ci_report_tree_plot(
    fit = selected_forest_surrogate$fit,
    data = selected_forest_surrogate$data,
    outcome_name = selected_forest_surrogate$prediction_name,
    outcome_label = "U5",
    ci_type = selected_tree_type,
    var_labels = var_labels,
    fontsize = 11
)


```

The first split was type of residence, separating urban and rural children. Among urban children, the main divisions were father’s agricultural occupation and household wealth. Among rural children, the main divisions were household wealth, maternal education, and residence in Katanga. The lowest terminal-node mortality was observed among urban children in the high-wealth branch, while the highest terminal-node mortality was observed among rural, low-wealth children in Katanga. Compared with the single L tree, the surrogate gives a smoother version of the same rural-urban, socioeconomic, and regional pattern, with less extreme terminal-node mortality differences. The criterion-specific forest surrogate trees are shown in the appendix for comparison (@fig-appendix-forest-surrogate-cig; @fig-appendix-forest-surrogate-cic; @fig-appendix-forest-surrogate-ci).

### Variable importance

The variable-importance plots support the interpretation of the selected level-dependent models. In both the $L$ tree and $L$ forest, the type of residence gives the largest reduction in concentration-index impurity [@fig-results-selected-model-variable-importance]. In the single tree, residence is followed by father’s agricultural occupation, Kinshasa, household wealth, Equateur, and Katanga, reflecting a structure organised around the rural-urban divide, occupational position, household wealth, and regional differences.

```{r results-selected-model-variable-importance, fig.width=8, fig.height=4}
#| label: fig-results-selected-model-variable-importance
#| fig-cap: "Variable importance for the selected concentration-index tree and forest."
#| fig-pos: "H"
selected_tree_importance <- copy(drc_results_objects$tree_variable_importance[
    criterion == selected_tree_type
])
selected_forest_importance <- copy(drc_results_objects$forest_variable_importance[
    criterion == selected_tree_type
])

tree_importance_plot <- make_importance_plot(
    selected_tree_importance,
    title = paste("Selected tree:", selected_tree_type),
    fill = "#2F6F9F"
)
forest_importance_plot <- make_importance_plot(
    selected_forest_importance,
    title = paste("Selected forest:", selected_tree_type),
    fill = "#4B7F52"
)

gridExtra::grid.arrange(tree_importance_plot,
    forest_importance_plot,
    ncol = 2
)
```

The $L$ forest gives a less concentrated ranking than the single tree. Residence is still the first variable, but the forest also gives weight to mother’s agricultural occupation, maternal education, household wealth, father’s agricultural occupation, and regional indicators. In the appendix plots, the same groups of variables appear again under the other criteria: residence, region, occupation, education, and wealth [@fig-appendix-tree-variable-importance; @fig-appendix-forest-variable-importance].


## Simulation Study

### Null case

From (@fig-results-simulation-null-trees) we see that the algorithm developed here did not find a meaningful structure and behaved as expected. This shows that the tree-growing rule does not create subgroup structure when the outcome and socioeconomic ranking are independent, since the tree uses concentraion index as an impurity measure, which measures how sociaeconomic ranking and outcome covary. 


```{r results-simulation-null-trees, fig.width=1.5, fig.height=1.5}
#| label: fig-results-simulation-null-trees
#| fig-cap: "Null simulation concentration-index trees by criterion."
#| fig-subcap:
#|   - "CI"
#|   - "CIg"
#|   - "CIc"
#|   - "L"
#| layout-ncol: 2

null_plot_types <- c("CI", "CIg", "CIc", "L")
if (
  "Null simulation" %in% names(simulation_tree_plots) &&
    all(null_plot_types %in% names(simulation_tree_plots[["Null simulation"]]))
) {
  for (type in null_plot_types) {
    grid::grid.newpage()
    grid::grid.draw(simulation_tree_plots[["Null simulation"]][[type]])
  }
}
```

### Poor subgroup case


In the positive-control simulation, the method(the concentration index tree with $CIc$ as an impurity measure) recovered the planted subgroup structure. The tree first separated urban from rural observations, then split the rural group by province and maternal education (@fig-results-simulation-tree). This structure matched the way the simulated outcome was generated: poorer observations in the planted rural, low-education, selected-province subgroup had higher outcome prevalence. The terminal nodes also followed the expected gradient, with the highest prevalence among observations in the most disadvantaged planted subgroup.


```{r results-simulation-tree, fig.width=9, fig.height=4}
#| label: fig-results-simulation-tree
#| fig-cap: "CIc concentration-index tree recovered from simulated data with a known subgroup-generating mechanism."

if (
  "Poor subgroup simulation" %in% names(simulation_tree_plots) &&
    "CIc" %in% names(simulation_tree_plots[["Poor subgroup simulation"]])
) {
  grid::grid.newpage()
  grid::grid.draw(simulation_tree_plots[["Poor subgroup simulation"]][["CIc"]])
}
```

We see the same pattern across the sensitivity plots in the appendix the trees developed using $CI$, $CIg$, and $CIc$, and $L$ as impurity measures(@fig-appendix-simulation-positive-control-ci; @fig-appendix-simulation-positive-control-cic; @fig-appendix-simulation-positive-control-l).


# Discussion

## Main findings

This thesis developed and applied a tree-based model for analysing socioeconomic inequality in under-five mortality using the concentration index impurity. The main finding is that inequality in the DRC data could not be explained by household wealth alone. Under-five mortality was also related to residence, maternal education, skilled birth attendance, parental occupation, and province. Overall, 10.32% of children died before age five. The children who died were more often from low-wealth households, rural households, mothers with no education, births without skilled attendance, and some provinces, especially Katanga. This is in line with earlier DRC work on geographic variation in under-five mortality, and with decomposition studies showing that wealth, education, residence, and occupation contribute to mortality inequalities in African settings [@kandala2014drc; @vanmalderen2013under5; @vanmalderen2019socioeconomic].

The indirect methods provided a variable view of this inequality. In the regression decomposition, mother’s education made the largest contribution, accounting for 23.1% under $CI$ and 28.7% under $CIg$ and $CIc$. Father’s agricultural occupation was the second-largest contributor, accounting for 17.9% under $CI$ and 19.9% under $CIg$ and $CIc$. These results are consistent with earlier decomposition work, which found that household wealth, parental education, occupation, and residence were important contributors to under-five mortality inequality [@vanmalderen2013under5; @vanmalderen2019socioeconomic]. The SHAP analysis (from the fitted random forest) gave a less concentrated picture: residence and household wealth each contributed about 12%, followed by mother’s occupation, father’s occupation, mother’s education, and skilled birth attendance. Shap-based decomposition provides an addition to the regression methods by allowing the use of more flexible models such as Random Forests or XGBoost. 

The direct concentration-index trees added the subgroup interpretation that is less visible in the regression-based decomposition. While the indirect methods ranked variables by their contribution to inequality, the trees showed how these variables combined within the data to define groups with different inequality patterns. The rank-dependent trees based on $CI$, $CIg$, and $CIc$ identified a structure involving maternal education, father’s agricultural occupation, skilled birth attendance, household wealth, and Katanga. Mortality was lower among children whose mothers had some education and whose fathers were not in agriculture, and higher among children whose mothers had no education, who lived in low-wealth households, and who resided in Katanga. However, these partitions did not validate well. Although the rank-based methods removed about 87% to 88% of root-node inequality in training, their validation gains were negative, suggesting that the strong training structure did not transfer well to held-out folds.


## Rank-dependent and level-dependent views of inequality


From these analyses, we see differences in the results (@fig-appendix-ci-tree-ci; @fig-appendix-ci-tree-cig; @fig-appendix-ci-tree-cic) in subgroups formed by the two ways of defining the impurity measure when developing the concentration index tree. In the rank-dependent criteria $CI$, $CIg$, and $CIc$, a child's contribution depends on the child's position in the current node's socioeconomic ranking. After each split, the children in the new subgroup are ranked again, so the same child can receive a different rank because the comparison group has changed. In the level-dependent criterion, re-ranking is still done but in $L$, the child's observed socioeconomic level is carried into the subgroup, so a child's rank also reflects their wealth level.

The differences also showed up in cross-validation; the level-dependent method outperformed the rank-dependent methods. The rank-dependent trees had strong training performance but negative validation gains, while the level-dependent tree kept a positive validation gain (@fig-appendix-tree-validation-path; @tbl-appendix-tree-selected-setting-summary). The same pattern was seen in the forests: averaging trees reduced the validation loss for the rank-dependent criteria, but their validation gains remained below zero, whereas the $L$ forest retained a positive validation gain (@fig-appendix-forest-validation-path; @tbl-appendix-forest-selected-setting-summary). The validation results show that $CI$, $CIg$, and $CIc$ identified detailed in-sample subgroup rules, whereas $L$ produced a more reproducible subgroup structure in this application.

These results are also mirrored by the subgroup-decomposability argument in [@erreygers2017subgroup]. In their comparison of rank-dependent and level-dependent measures, they show that the level-dependent index $L$ is subgroup-decomposable and rank-dependent measures are more difficult to decompose cleanly across population groups. Decomposability of the level-dependent concentration index shown by [@erreygers2017subgroup] can also explain why $L$ performed better as a tree impurity criterion in this application: children inherit their socioeconomic level within each subgroup, rather than being re-weighted only by a new within-node rank. Although the DHS wealth index is an asset-based score[@dhsprogram_wealth_index_construction] and [@erreygers2017socioeconomic] recommends using Level depedent concentration index($L$) with a wealth-ranking variable that has ratio properties. The cross-validation results show that the $L$ concentration index impurity measure yielded a structure transferable to held-out samples using the DHS wealth score.


## Methodological and data limitations

The analysis is descriptive and should not be interpreted as a causal design. The DHS data are cross-sectional, and the fitted trees identify subgroup patterns rather than intervention effects. A terminal node with high mortality does not show that the splitting variables caused the outcome; it shows that the observed covariates define a group with a different mortality and inequality profile.

The choice of socioeconomic ranking variable is also a limitation. The DHS wealth index is appropriate for ordering households from poorer to richer, but it is not equivalent to measured income, consumption, or expenditure. This matters especially for the level-dependent index, because level-dependent weights are easier to justify when socioeconomic levels have a clear ratio-scale interpretation. The validation results favour the $L$ criterion in this application, but that empirical stability should be read alongside this measurement caveat.

The SHAP analysis is also approximate. It is based on a fitted predictive forest and a sampled SHAP calculation, so the reported percentages describe how the fitted model allocates prediction differences across variables. They do not provide causal effects, and they are not the same object as the concentration-index tree impurity. Their value in this thesis is comparative: they show whether a flexible predictive model points to the same determinant groups as the direct inequality trees.

## Ethical, societal, and stakeholder relevance



This thesis uses child-level data from the Democratic Republic of the Congo DHS phase V survey [@dhs_drc_2007_dataset]. These data contain sensitive information, including child survival status, birth histories, household wealth, residence, province, maternal education, occupation, and skilled birth attendance. Even without names, the data describe mothers, children, households, and family deaths. The DHS Program states that participation is voluntary and that respondents' identities and information are kept confidential [@dhs_privacy_respondents]. In this thesis, we used secondary survey data and should remain within the conditions under which we were granted access to the DHS files.

The concentration index trees developed here can identify subgroups with higher mortality and greater socioeconomic disadvantage. For example, the fitted trees point to rural children from low-wealth households in Katanga, or children whose mothers had no education and who lived in low-wealth households. These subgroups are useful for public health analysis and planning, but the results should be used carefully. Groups we identified through concentration index trees should not be described as if the children, mothers, or provinces are the problem. The tree is showing where mortality and disadvantage are concentrated in the data. The interpretation should therefore focus on unequal conditions, access to services, and policy response, not labelling groups.

Compared with other statistical methods used in decomposition, such as generalised linear models and tree-based models, tree-based models are easier to interpret. The variable percentage contribution from regression or Shap-based decomposition can show which variables contribute more. However, it cannot show how disadvantage is organised across residence, household wealth, occupation, maternal background, and province.   A tree is easier to discuss because it works like a flow chart. As shown in this thesis, the tree partitions the data graphically into high- and low-risk groups. Concentration index trees make it easier to follow the path to these high- and low-risk groups and identify them for public health planning. 


# Future research

Future research should examine why the rank-dependent criteria overfit in this application and whether alternative validation rules, bootstrap stability checks, or constraints on early splits improve their held-out performance. A second direction is to repeat the level-dependent analysis with socioeconomic variables that have clearer level meaning, such as consumption or income, where available. The method should also be applied to another DHS country sample, such as Kenya, to assess whether the same rural-urban, occupational, educational, and regional structure appears outside the DRC data.

# Conclusion

The thesis shows that concentration-index trees can turn socioeconomic inequality in under-five mortality into interpretable subgroup rules. In the DRC application, the indirect regression decomposition highlighted mother's education and father's agricultural occupation, while the SHAP, tree, forest, and surrogate results repeatedly pointed to residence, household wealth, education, occupation, skilled birth attendance, and province. The rank-dependent trees recovered a clear in-sample socioeconomic gradient, but they did not validate well. The level-dependent tree and forest produced a more stable subgroup structure centred on rural-urban residence, occupation, wealth, and region. The method therefore adds a useful descriptive tool to classical decomposition: it does not only rank determinants, but shows how determinants combine to define inequality-relevant groups.


# Appendix

## Results


### Appendix R1: Exploratory Data Outputs

```{r appendix-weighted-baseline-table}
#| label: tbl-appendix-weighted-baseline
#| tbl-cap: "Weighted distribution of analysis variables by under-five mortality status."

weighted_baseline_table(drc_results_objects, outcome_labels)
```

```{r appendix-province-mortality-distribution, fig.width=9.2, fig.height=6.2}
#| label: fig-appendix-province-mortality-distribution
#| fig-cap: "Weighted province distribution by under-five mortality status."

province_plot_dt <- copy(as.data.table(drc_results_objects$raw_congo_model_dt))
province_plot_dt[, u5_mortality := fifelse(
  deadu5_num == 1,
  outcome_labels[["died"]],
  outcome_labels[["alive"]]
)]
province_plot_dt <- province_plot_dt[
  ,
  .(weighted_n = sum(sample_weight, na.rm = TRUE)),
  by = .(reg, u5_mortality)
]
province_plot_dt[
  ,
  weighted_percent := 100 * weighted_n / sum(weighted_n),
  by = reg
]
province_order <- province_plot_dt[
  u5_mortality == outcome_labels[["died"]]
][
  order(-weighted_percent)
]$reg
province_plot_dt[, reg := factor(reg, levels = province_order)]

ggplot(
  province_plot_dt,
  aes(x = reg, y = weighted_percent, fill = u5_mortality)
) +
  geom_col(width = 0.72) +
  coord_flip() +
  scale_fill_manual(
    values = setNames(c("#3268A8", "#B94E48"), c(outcome_labels[["alive"]], outcome_labels[["died"]]))
  ) +
  labs(x = NULL, y = "Weighted percent within province", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

### Appendix R2: Regression and Predictive-Forest Details

#### Regression model summary

```{r appendix-regression-model-summary}
#| label: tbl-appendix-regression-model-summary
#| tbl-cap: "Survey-weighted regression model summary."

regression_summary_out <- copy(drc_results_objects$regression_model_summary)
if (!nrow(regression_summary_out) || "note" %in% names(regression_summary_out)) {
  regression_summary_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd after installing broom."
  )
} else {
  regression_summary_out <- regression_summary_out[
    ,
    .(
      term = term_label,
      estimate,
      std_error = std.error,
      statistic,
      p_value = format.pval(p.value, digits = 3, eps = 0.001),
      conf_low = conf.low,
      conf_high = conf.high
    )
  ]
  setnames(
    regression_summary_out,
    c("term", "estimate", "std_error", "statistic", "p_value", "conf_low", "conf_high"),
    c("Term", "Estimate", "Std. error", "Statistic", "p-value", "Conf. low", "Conf. high")
  )
}

report_compact_table(
  regression_summary_out,
  digits = 3,
  caption = "Survey-weighted regression model summary",
  column_widths = c("4.5cm", rep(NA_character_, 6L))
)
```

#### Predictive ranger tuning results

```{r appendix-ranger-sampling-summary}
#| label: tbl-appendix-ranger-sampling-summary
#| tbl-cap: "Weighted class distribution before and after ranger undersampling."

ranger_sampling_out <- copy(drc_results_objects$predictive_forest_sampling %||% data.table())
if (!nrow(ranger_sampling_out)) {
  ranger_sampling_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with run_shap: true to build the predictive ranger sample."
  )
} else {
  ranger_sampling_out <- ranger_sampling_out[
    ,
    .(
      Sample = sample,
      Outcome = outcome,
      Rows = rows,
      `Weighted N` = weighted_n,
      `Weighted percent` = weighted_percent
    )
  ]
}

report_compact_table(
  ranger_sampling_out,
  digits = 2,
  caption = "Weighted class distribution before and after ranger undersampling"
)
```

```{r appendix-ranger-cross-validation}
#| label: tbl-appendix-ranger-cross-validation
#| tbl-cap: "Cross-validation results for the predictive ranger model."

ranger_cv_out <- copy(drc_results_objects$predictive_forest_tuning)
if (!nrow(ranger_cv_out)) {
  ranger_cv_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with run_shap: true to tune the predictive ranger."
  )
} else {
  ranger_cv_out[
    ,
    rank := frank(-mean, ties.method = "first"),
    by = .metric
  ]
  ranger_cv_out <- ranger_cv_out[
    order(.metric, rank),
    .(
      metric = fifelse(.metric == "sens", "sensitivity", .metric),
      mtry,
      min_n,
      mean,
      std_error = std_err,
      folds = n,
      rank
    )
  ]
  setnames(
    ranger_cv_out,
    c("metric", "mtry", "min_n", "mean", "std_error", "folds", "rank"),
    c("Metric", "mtry", "min_n", "Mean", "Std. error", "Folds", "Rank")
  )
}

report_compact_table(
  ranger_cv_out,
  digits = 4,
  caption = "Cross-validation results for the predictive ranger model",
  column_widths = c("2.4cm", rep(NA_character_, 6L))
)
```

#### Predictive ranger SHAP summary plot
```{r appendix-shap-summary-plot, fig.width=6.2, fig.height=5}
#| label: fig-appendix-shap-summary
#| fig-cap: "SHAP summary plot for the predictive ranger model."

shap_plot_dt <- copy(as.data.table(drc_results_objects$shap_summary_long))

  top_shap_features <- shap_plot_dt[
    ,
    .(mean_abs_shap = mean(abs(shap_value), na.rm = TRUE)),
    by = .(feature, feature_label)
  ][
    order(-mean_abs_shap)
  ][
    seq_len(min(.N, 16L))
  ]
  shap_plot_dt <- shap_plot_dt[feature %in% top_shap_features$feature]
  shap_plot_dt <- merge(
    shap_plot_dt,
    top_shap_features[, .(feature, feature_label, mean_abs_shap)],
    by = c("feature", "feature_label"),
    all.x = TRUE,
    sort = FALSE
  )
  top_shap_features[
    ,
    feature_axis_label := sprintf("%s", feature_label)
  ]
  shap_plot_dt <- merge(
    shap_plot_dt,
    top_shap_features[, .(feature, feature_axis_label)],
    by = "feature",
    all.x = TRUE,
    sort = FALSE
  )
  shap_plot_dt[
    ,
    feature_axis_label := factor(
      feature_axis_label,
      levels = rev(top_shap_features$feature_axis_label)
    )
  ]

  ggplot(
    shap_plot_dt,
    aes(x = shap_value, y = feature_axis_label, color = feature_value_scaled)
  ) +
    geom_vline(xintercept = 0, color = "black", linewidth = 0.35) +
    geom_point(
      position = position_jitter(width = 0, height = 0.18),
      size = 1.25,
      alpha = 0.75
    ) +
    scale_color_gradient(
      name = "Feature value",
      low = "#FDBE3B",
      high = "#6A00C2",
      breaks = c(0, 1),
      labels = c("Low", "High")
    ) +
    labs(x = "SHAP value (impact on model output)", y = NULL) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      legend.title = element_text(size = 11),
      legend.text = element_text(size = 11)
    )

```

### Appendix R3: Concentration-Index Tree Details

#### Tree tuning and selection details

```{r appendix-tree-best-controls}
#| label: tbl-appendix-tree-best-controls
#| tbl-cap: "Best tree hyperparameters selected by validation gain within each impurity criterion."

tree_best_controls_out <- copy(drc_results_objects$tree_best_controls %||% data.table())
if (!nrow(tree_best_controls_out)) {
  tree_best_controls_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with RUN_CV=true"
  )
}

report_compact_table(
  tree_best_controls_out,
  digits = 4,
  caption = "Best tree hyperparameters selected by validation gain within each impurity criterion"
)
```

#### Tree fit and selection summaries

```{r appendix-tree-selected-setting-summary}
#| label: tbl-appendix-tree-selected-setting-summary
#| tbl-cap: "Tree summary for the selected settings."

tree_selection_for_summary <- copy(drc_results_objects$tree_selection %||% data.table())
if (!nrow(tree_selection_for_summary)) {
  tree_fit_table_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with RUN_CV=true"
  )
} else {
  tree_fit_table_out <- fit_summary_long(tree_selection_for_summary)
}

report_compact_table(
  tree_fit_table_out,
  digits = 4,
  caption = "Tree summary for the selected settings"
)
```

#### Tree selection metrics
```{r appendix-tree-selection}
#| label: tbl-appendix-tree-selection
#| tbl-cap: "Selection metrics for fitted concentration-index trees."

tree_selection_out <- copy(drc_results_objects$tree_selection)
tree_selection_out <- tree_selection_out[
  ,
  .(
    criterion = type,
    root_objective = mean_root_objective,
    train_gain = mean_train_gain,
    train_relative_gain = mean_train_relative_gain,
    validation_gain = mean_validation_gain,
    validation_relative_gain = mean_validation_relative_gain,
    terminal_nodes
  )
]
setnames(
  tree_selection_out,
  c(
    "criterion",
    "root_objective",
    "train_gain",
    "train_relative_gain",
    "validation_gain",
    "validation_relative_gain",
    "terminal_nodes"
  ),
  c(
    "Criterion",
    "Root objective",
    "Train gain",
    "Train gain/root",
    "Validation gain",
    "Validation gain/root",
    "Terminal nodes"
  )
)

report_compact_table(
  tree_selection_out,
  digits = 4,
  caption = "Selection metrics for fitted concentration-index trees",
  column_widths = c("1.4cm", rep("1.8cm", 6L))
)
```


```{r appendix-tree-variable-importance-plot, fig.width=8, fig.height=6}
#| label: fig-appendix-tree-variable-importance
#| fig-cap: "Top tree variables by concentration-index criterion."
#| fig-pos: "H"

plot_importance_by_criterion(
  drc_results_objects$tree_variable_importance,
  "Top Tree Variables by Concentration-Index Criterion",
  fill = "#3268A8"
)
```



### Additional Concentration-Index Tree Plots

```{r appendix-ci-tree-cig, fig.width=9, fig.height=5.0}
#| label: fig-appendix-ci-tree-cig
#| fig-cap: "Concentration-index tree fitted with the CIg criterion."
#| fig-pos: "H"

if (isTRUE(params$render_appendix_tree_plots) && "CIg" %in% names(tree_models_by_type)) {
  ci_report_tree_plot(
    fit = tree_models_by_type[["CIg"]],
    data = congo_model_dt,
    outcome_name = "deadu5_num",
    outcome_label = "U5",
    ci_type = "CIg",
    var_labels = congo_var_labels,
    fontsize = 11
  )
}
```

```{r appendix-ci-tree-ci, fig.width=9, fig.height=5.0}
#| label: fig-appendix-ci-tree-ci
#| fig-cap: "Concentration-index tree fitted with the CI criterion."
#| fig-pos: "H"

if (isTRUE(params$render_appendix_tree_plots) && "CI" %in% names(tree_models_by_type)) {
  ci_report_tree_plot(
    fit = tree_models_by_type[["CI"]],
    data = congo_model_dt,
    outcome_name = "deadu5_num",
    outcome_label = "U5",
    ci_type = "CI",
    var_labels = congo_var_labels,
    fontsize = 11
  )
}
```

### Appendix R4: Rpart Comparison Tree Details



```{r appendix-lecturer-rpart-simple, fig.width=14, fig.height=8}
#| label: fig-appendix-lecturer-rpart-simple
#| fig-cap: "Lecturer rpart comparison tree using absolute CI impurity."
#| fig-pos: "H"

if (isTRUE(params$render_appendix_tree_plots) && "simple_abs" %in% names(lecturer_rpart_models)) {
  plot_lecturer_rpart("simple_abs", fontsize = 11)
}
```


### Appendix 5: Concentration-Index Forest Details

#### Forest tuning and selection details

```{r appendix-forest-best-controls}
#| label: tbl-appendix-forest-best-controls
#| tbl-cap: "Best forest hyperparameters selected by validation gain within each impurity criterion."

forest_best_controls_out <- copy(drc_results_objects$forest_best_controls %||% data.table())
if (!nrow(forest_best_controls_out)) {
  forest_best_controls_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with RUN_FOREST=true"
  )
}

report_compact_table(
  forest_best_controls_out,
  digits = 4,
  caption = "Best forest hyperparameters selected by validation gain within each impurity criterion"
)
```

#### Forest fit and selection summaries

```{r appendix-forest-selected-setting-summary}
#| label: tbl-appendix-forest-selected-setting-summary
#| tbl-cap: "Forest summary for the selected settings."
forest_selection_for_summary <- copy(drc_results_objects$forest_selection %||% data.table())
if (!nrow(forest_selection_for_summary)) {
    forest_fit_table_out <- data.table(
        note = "Run Generate_DRC_Results_Objects.qmd with RUN_FOREST=true"
    )
} else {
    forest_fit_table_out <- fit_summary_long(forest_selection_for_summary)
}

report_compact_table(
    forest_fit_table_out,
    digits = 4,
    caption = "Forest summary for the selected settings"
)
```

#### Forest selection metrics

```{r appendix-forest-selection}
#| label: tbl-appendix-forest-selection
#| tbl-cap: "Selection metrics for fitted concentration-index forests."
forest_selection_out <- copy(drc_results_objects$forest_selection %||% data.table())
if (!nrow(forest_selection_out)) {
    forest_selection_out <- data.table(
        note = "Run Generate_DRC_Results_Objects.qmd with RUN_FOREST=true"
    )
} else {
    forest_selection_out <- forest_selection_out[
        ,
        .(
            criterion = type,
            root_objective = mean_root_objective,
            train_gain = mean_train_gain,
            train_relative_gain = mean_train_relative_gain,
            validation_gain = mean_validation_gain,
            validation_relative_gain = mean_validation_relative_gain,
            mean_terminal_nodes
        )
    ]
    setnames(
        forest_selection_out,
        c(
            "criterion",
            "root_objective",
            "train_gain",
            "train_relative_gain",
            "validation_gain",
            "validation_relative_gain",
            "mean_terminal_nodes"
        ),
        c(
            "Criterion",
            "Root objective",
            "Train gain",
            "Train gain/root",
            "Validation gain",
            "Validation gain/root",
            "Mean terminal nodes"
        )
    )
}

report_compact_table(
    forest_selection_out,
    digits = 4,
    caption = "Selection metrics for fitted concentration-index forests",
    column_widths = c("1.4cm", rep("1.8cm", 6L))
)
```


#### Forest complexity plots

```{r}
#| label: fig-appendix-forest-complexity-path
#| fig-cap: "Forest complexity by number of terminal nodes."
#| fig-pos: "H"
#| fig-width: 6.3
#| fig-height: 3.8

  plot_terminal_nodes_path(
    forest_path_plot_dt,
    title = "Forest Complexity"
  )
```

```{r appendix-forest-variable-importance-plot, fig.width=10, fig.height=7}
#| label: fig-appendix-forest-variable-importance
#| fig-cap: "Top forest variables by concentration-index criterion."
#| fig-pos: "H"

plot_importance_by_criterion(
  drc_results_objects$forest_variable_importance,
  "Top Forest Variables by Concentration-Index Criterion",
  fill = "#4B7F52"
)
```

#### Forest Surrogate Trees plots

The selected forest surrogate appears in the results section. The remaining
criterion-specific forest surrogates are retained here for comparison.

```{r appendix-forest-surrogate-objects, include=FALSE}
forest_surrogates <- drc_results_objects$forest_surrogates_by_type %||% list()
```

```{r appendix-forest-surrogate-ci, fig.width=9, fig.height=5}
#| label: fig-appendix-forest-surrogate-ci
#| fig-cap: "Surrogate tree for the CI concentration-index forest."
#| fig-pos: "H"

if ("CI" %in% names(forest_surrogates)) {
  forest_surrogate_ci <- forest_surrogates[["CI"]]
  ci_report_tree_plot(
    fit = forest_surrogate_ci$fit,
    data = forest_surrogate_ci$data,
    outcome_name = forest_surrogate_ci$prediction_name,
    outcome_label = "U5",
    ci_type = "CI",
    var_labels = var_labels,
    fontsize = 11
  )
}
```

```{r appendix-forest-surrogate-cig, fig.width=9, fig.height=5}
#| label: fig-appendix-forest-surrogate-cig
#| fig-cap: "Surrogate tree for the CIg concentration-index forest."
#| fig-pos: "H"

if ("CIg" %in% names(forest_surrogates)) {
  forest_surrogate_cig <- forest_surrogates[["CIg"]]
  ci_report_tree_plot(
    fit = forest_surrogate_cig$fit,
    data = forest_surrogate_cig$data,
    outcome_name = forest_surrogate_cig$prediction_name,
    outcome_label = "U5",
    ci_type = "CIg",
    var_labels = var_labels,
    fontsize = 11
  )
}



```

```{r appendix-forest-surrogate-cic, fig.width=9, fig.height=5}
#| label: fig-appendix-forest-surrogate-cic
#| fig-cap: "Surrogate tree for the CIc concentration-index forest."
#| fig-pos: "H"

if ("CIc" %in% names(forest_surrogates)) {
  forest_surrogate_cic <- forest_surrogates[["CIc"]]
  ci_report_tree_plot(
    fit = forest_surrogate_cic$fit,
    data = forest_surrogate_cic$data,
    outcome_name = forest_surrogate_cic$prediction_name,
    outcome_label = "U5",
    ci_type = "CIc",
    var_labels = var_labels,
    fontsize = 11
  )
}
```

```{r appendix-forest-surrogate-l, fig.width=9, fig.height=5}
#| label: fig-appendix-forest-surrogate-l
#| fig-cap: "Surrogate tree for the L concentration-index forest."
#| fig-pos: "H"

if ("L" %in% names(forest_surrogates)) {
  forest_surrogate_l <- forest_surrogates[["L"]]
  ci_report_tree_plot(
    fit = forest_surrogate_l$fit,
    data = forest_surrogate_l$data,
    outcome_name = forest_surrogate_l$prediction_name,
    outcome_label = "U5",
    ci_type = "L",
    var_labels = var_labels,
    fontsize = 11
  )
}
```



### Appendix S: Simulation Sensitivity Plots

The CIg tree is shown in the main results section. The remaining
positive-control simulation trees are shown here to check whether the recovered
subgroup pattern is stable across concentration-index criteria.

```{r appendix-simulation-positive-control-ci, fig.width=9, fig.height=6}
#| label: fig-appendix-simulation-positive-control-ci
#| fig-cap: "Positive-control simulation tree fitted with the CI criterion."
#| fig-pos: "H"

if (
  "Poor subgroup simulation" %in% names(simulation_tree_plots) &&
    "CI" %in% names(simulation_tree_plots[["Poor subgroup simulation"]])
) {
  grid::grid.newpage()
  grid::grid.draw(simulation_tree_plots[["Poor subgroup simulation"]][["CI"]])
}
```

```{r appendix-simulation-positive-control-cic, fig.width=9, fig.height=6}
#| label: fig-appendix-simulation-positive-control-cic
#| fig-cap: "Positive-control simulation tree fitted with the CIc criterion."
#| fig-pos: "H"

if (
  "Poor subgroup simulation" %in% names(simulation_tree_plots) &&
    "CIc" %in% names(simulation_tree_plots[["Poor subgroup simulation"]])
) {
  grid::grid.newpage()
  grid::grid.draw(simulation_tree_plots[["Poor subgroup simulation"]][["CIc"]])
}
```

```{r appendix-simulation-positive-control-l, fig.width=9, fig.height=6}
#| label: fig-appendix-simulation-positive-control-l
#| fig-cap: "Positive-control simulation tree fitted with the L criterion."
#| fig-pos: "H"

if (
  "Poor subgroup simulation" %in% names(simulation_tree_plots) &&
    "L" %in% names(simulation_tree_plots[["Poor subgroup simulation"]])
) {
  grid::grid.newpage()
  grid::grid.draw(simulation_tree_plots[["Poor subgroup simulation"]][["L"]])
}
```
