---
title: "Decomposition of Concentration-Index Trees"
author: "Moses Mburu"
format:
  pdf:
    documentclass: article
    papersize: a4
    fontsize: 11pt
    geometry:
      - margin=25mm
    pdf-engine: xelatex
    toc: false
    number-sections: true
    number-depth: 3
    fig-cap-location: bottom
    tbl-cap-location: top
    colorlinks: false
    fig-pos: "H"
    tbl-pos: "H"
    fig-dpi: 500
execute:
  echo: false
  warning: false
  message: false
  fig-dpi: 500
params:
  render_appendix_tree_plots: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 10,
  fig.height = 6,
  dpi = 500,
  dev = "pdf",
  warning = FALSE,
  message = FALSE
)

library(gridExtra)
library(data.table)
library(ggplot2)
library(grid)
library(knitr)
library(ineqTrees)
library(partykit)
library(here)
library(kableExtra)
source(here("R", "report_helpers.R"))
source(here("R", "ci_known_subgroup_demo.R"))
source(here("R", "ci_rpart_comparison_methods.R"))

if (!exists("params", inherits = FALSE)) {
  params <- list(render_appendix_tree_plots = TRUE)
}


```

```{r load-results, include=FALSE}
results_object_file <- here("data", "drc_report_results_objects.rda")
if (!file.exists(results_object_file)) {
  stop(
    "Missing ", results_object_file, ". Render Generate_DRC_Results_Objects.qmd first.",
    call. = FALSE
  )
}

load(results_object_file)

congo_model_dt <- drc_results_objects$congo_model_dt
congo_var_labels <- drc_results_objects$congo_var_labels
congo_predictors <- drc_results_objects$congo_predictors
congo_ci_formula <- drc_results_objects$congo_ci_formula
outcome_labels <- drc_results_objects$outcome_labels %||% c(
  alive = "Alive at age 5",
  died = "Died before age 5"
)
selected_tree_type <- drc_results_objects$selected_tree_type
selected_tree_fit <- drc_results_objects$selected_tree_fit
tree_models_by_type <- drc_results_objects$tree_models_by_type
lecturer_rpart_models <- drc_results_objects$lecturer_rpart_models
known_subgroup_sim_dt <- drc_results_objects$known_subgroup_sim_dt
known_subgroup_fit <- drc_results_objects$known_subgroup_fit

format_variable_label <- function(x) {
  predictor_names <- names(congo_var_labels)
  vapply(x, function(value) {
    if (is.na(value) || !nzchar(value)) {
      return(NA_character_)
    }
    if (value %in% predictor_names) {
      return(unname(congo_var_labels[value]))
    }

    hits <- predictor_names[value == predictor_names | startsWith(value, predictor_names)]
    if (length(hits)) {
      parent <- hits[which.max(nchar(hits))]
      suffix <- trimws(sub(paste0("^", parent), "", value))
      suffix <- gsub("^[:._]+", "", suffix)
      if (nzchar(suffix)) {
        return(paste0(unname(congo_var_labels[parent]), ": ", suffix))
      }
      return(unname(congo_var_labels[parent]))
    }

    value
  }, character(1L))
}

make_importance_plot <- function(importance_dt, title, fill, top_n = 10L) {
  importance_dt <- as.data.table(importance_dt)
  if (!nrow(importance_dt) || !"importance" %in% names(importance_dt)) {
    return(
      ggplot() +
        annotate("text", x = 0, y = 0, label = "Not available") +
        labs(title = title) +
        theme_void(base_size = 11)
    )
  }

  importance_dt <- importance_dt[is.finite(importance) & importance != 0]
  if (!nrow(importance_dt)) {
    return(
      ggplot() +
        annotate("text", x = 0, y = 0, label = "No non-zero importance") +
        labs(title = title) +
        theme_void(base_size = 11)
    )
  }

  importance_dt[, variable_label := format_variable_label(variable)]
  importance_dt[
    ,
    variable_label := vapply(
      variable_label,
      function(value) paste(strwrap(value, width = 34), collapse = "\n"),
      character(1L)
    )
  ]
  setorder(importance_dt, -importance)
  importance_dt <- head(importance_dt, top_n)
  importance_dt[, variable_label := make.unique(variable_label)]
  importance_dt[, variable_label := factor(variable_label, levels = rev(variable_label))]

  ggplot(importance_dt, aes(x = importance, y = variable_label)) +
    geom_col(fill = fill, width = 0.68) +
    labs(title = title, x = "Importance", y = NULL) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 11),
      axis.text.y = element_text(size = 10),
      panel.grid.minor = element_blank()
    )
}

extract_surrogate_importance <- function(surrogate_object) {
  if (is.null(surrogate_object) ||
      !length(surrogate_object) ||
      is.null(surrogate_object$fit$variable.importance)) {
    return(data.table(variable = character(), importance = numeric()))
  }

  importance <- surrogate_object$fit$variable.importance
  data.table(
    variable = names(importance),
    importance = as.numeric(importance)
  )[order(-importance)]
}

plot_forest_surrogate <- function(type, fontsize = 6.2) {
  surrogate <- drc_results_objects$forest_surrogates_by_type[[type]]
  if (is.null(surrogate) || !length(surrogate)) {
    return(
      ggplot() +
        annotate("text", x = 0, y = 0, label = paste("No", type, "forest surrogate available")) +
        theme_void(base_size = 11)
    )
  }

  ci_report_tree_plot(
    fit = surrogate$fit,
    data = surrogate$data,
    outcome_name = surrogate$prediction_name,
    outcome_label = "Forest predicted U5 death",
    ci_type = type,
    var_labels = congo_var_labels,
    fontsize = fontsize
  )
}

plot_importance_by_criterion <- function(importance_dt, title, fill = "#3268A8") {
  plot_dt <- as.data.table(importance_dt)
  if (!nrow(plot_dt)) {
    return(
      ggplot() +
        annotate("text", x = 0, y = 0, label = "No variable importance available") +
        labs(title = title, x = NULL, y = NULL) +
        theme_void(base_size = 11)
    )
  }

  plot_dt[, variable_label := format_variable_label(variable)]
  plot_dt[
    ,
    variable_label := vapply(
      variable_label,
      function(value) paste(strwrap(value, width = 32), collapse = "\n"),
      character(1L)
    )
  ]
  plot_dt[
    ,
    facet_label := factor(
      paste(criterion, variable_label, sep = " | "),
      levels = rev(paste(criterion, variable_label, sep = " | "))
    )
  ]

  ggplot(plot_dt, aes(x = facet_label, y = importance)) +
    geom_col(width = 0.7, fill = fill) +
    coord_flip() +
    facet_wrap(vars(criterion), scales = "free_y", ncol = 2) +
    scale_x_discrete(labels = function(x) sub("^.* \\| ", "", x)) +
    labs(x = NULL, y = "Total split gain", title = title) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 11),
      axis.text.y = element_text(size = 9),
      panel.grid.minor = element_blank()
    )
}
```

# 4. Results

The results section reports only the objects needed to answer the main technical
question: whether inequality-oriented recursive partitioning identifies
subgroup structure that is visible, stable, and comparable with classical
decomposition output. Full generated outputs are moved to the appendix.

## 4.1 Exploratory Data Analysis

```{r results-outcome-distribution, fig.width=6.5, fig.height=3}
#| label: fig-results-outcome-distribution
#| fig-cap: "Distribution of under-five mortality in the DRC analysis data."
outcome_gg <- ggplot(
    drc_results_objects$outcome_distribution,
    aes(x = outcome_label, y = weighted_percent)
) +
    geom_col(fill = "#3268A8", width = 0.58) +
    geom_text(aes(label = sprintf("%.1f%%", weighted_percent)),
        vjust = -0.35, size = 3.3
    ) +
    labs(x = NULL, y = "Weighted percent") +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank())


```

```{r results-wealth-distribution, fig.width=6, fig.height=3}
#| label: fig-results-wealth-distribution
#| fig-cap: "Under-five mortality distribution and DHS wealth index in the DRC analysis data."
wealth_by_outcome_dt <- copy(congo_model_dt)
wealth_by_outcome_dt[, outcome_label := fifelse(
    deadu5_num == 1,
    outcome_labels[["died"]],
    outcome_labels[["alive"]]
)]

wealth_distr <- ggplot(
    wealth_by_outcome_dt,
    aes(x = wealth, weight = sample_weight)
) +
    geom_histogram(
        bins = 35, fill = "#3268A8",
        color = "white", linewidth = 0.2
    ) +
    # facet_wrap(~ outcome_label, ncol = 1, scales = "free_y") +
    labs(x = "Wealth index", y = "Weighted count") +
    theme_minimal(base_size = 11) +
    theme(
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold")
    )
grid.arrange(outcome_gg, wealth_distr, ncol = 2)
```



## 4.2 Application of developed tree models to the DRC DHS Data

### 4.2.1 Comparison of Indirect Decomposition Results

Categorical predictors enter the regression decomposition as separate one-hot
terms. The predictive ranger is fitted on the same design-matrix terms, so the
comparison table reports categorical levels at the same granularity instead of
collapsing them back to their parent variable.

```{r results-method-comparison}
#| label: tbl-results-method-comparison
#| tbl-cap: "Regression and predictive-forest SHAP percentage contributions by criterion."
method_comparison_report <- copy(drc_results_objects$method_comparison[
    ,
    .(
        variable = variable_label,
        regression_ci_pct_contribution,
        regression_cig_pct_contribution,
        regression_cic_pct_contribution,
        forest_shap_ci_pct_contribution,
        forest_shap_cig_pct_contribution,
        forest_shap_cic_pct_contribution,
        forest_shap_l_pct_contribution
    )
])

setnames(
    method_comparison_report,
    c(
        "variable",
        "regression_ci_pct_contribution",
        "regression_cig_pct_contribution",
        "regression_cic_pct_contribution",
        "forest_shap_ci_pct_contribution",
        "forest_shap_cig_pct_contribution",
        "forest_shap_cic_pct_contribution",
        "forest_shap_l_pct_contribution"
    ),
    c(
        "Variable",
        "CI (%)",
        "CIg (%)",
        "CIc (%)",
        "CI (%)",
        "CIg (%)",
        "CIc (%)",
        "L (%)"
    )
)

method_comparison_table <- report_compact_table(
    method_comparison_report,
    digits = 3,
    align = c("l", rep("r", 7L)),
    caption = "Regression and predictive-forest SHAP percentage contributions by criterion",
    column_widths = c("3.4cm", rep("1.35cm", 7L))
)


method_comparison_table <- kableExtra::add_header_above(
    method_comparison_table,
    c(" " = 1, "Regression" = 3, "Predictive-forest SHAP" = 4),
    bold = TRUE
)


method_comparison_table
```

### 4.2.2 Root-Node Inequality

```{r results-root-ci}
#| label: tbl-results-root-ci
#| tbl-cap: "Root-node inequality in under-five mortality by concentration-index criterion."

root_ci_report <- copy(drc_results_objects$root_ci)
root_ci_report[, root_node := "Full sample"]
root_ci_report <- dcast(
  root_ci_report,
  root_node ~ criterion,
  value.var = "estimate"
)
setcolorder(root_ci_report, intersect(c("root_node", "CI", "CIg", "CIc", "L"), names(root_ci_report)))
setnames(root_ci_report, "root_node", "Root node")

report_compact_table(
  root_ci_report,
  digits = 4,
  align = c("l", rep("r", ncol(root_ci_report) - 1L)),
  caption = "Root-node inequality in under-five mortality by concentration-index criterion",
  column_widths = c("3.2cm", rep("2cm", ncol(root_ci_report) - 1L))
)
```

### 4.2.3 Selected Concentration-Index Tree

```{r results-selected-tree, fig.width=12, fig.height=9}
#| label: fig-results-selected-ci-tree
#| fig-cap: "Selected concentration-index tree for under-five mortality in the DRC."
#| fig-pos: "H"

ci_report_tree_plot(
  fit = selected_tree_fit,
  data = congo_model_dt,
  outcome_name = "deadu5_num",
  outcome_label = "U5 death",
  ci_type = selected_tree_type,
  var_labels = congo_var_labels,
  fontsize = 11
)
```

### Surrogate Tree Comparison

```{r results-surrogate-tree, fig.width=14, fig.height=9}
#| label: fig-results-surrogate-tree
#| fig-cap: "Surrogate tree summarising the selected concentration-index forest predictions."
#| fig-pos: "H"

selected_forest_surrogate <- drc_results_objects$selected_forest_surrogate

if (is.null(selected_forest_surrogate) || !length(selected_forest_surrogate)) {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "Run Generate_DRC_Results_Objects.qmd with run_forest: true") +
    theme_void(base_size = 11)
} else {
  ci_report_tree_plot(
    fit = selected_forest_surrogate$fit,
    data = selected_forest_surrogate$data,
    outcome_name = selected_forest_surrogate$prediction_name,
    outcome_label = "U5 death",
    ci_type = selected_tree_type,
    var_labels = congo_var_labels,
    fontsize = 10
  )
}
```



```{r results-selected-model-variable-importance, fig.width=9, fig.height=4.4}
#| label: fig-results-selected-model-variable-importance
#| fig-cap: "Variable importance for the selected concentration-index tree and forest."
#| fig-pos: "H"
selected_tree_importance <- copy(drc_results_objects$tree_variable_importance[
    criterion == selected_tree_type
])
selected_forest_importance <- copy(drc_results_objects$forest_variable_importance[
    criterion == selected_tree_type
])

tree_importance_plot <- make_importance_plot(
    selected_tree_importance,
    title = paste("Selected tree:", selected_tree_type),
    fill = "#2F6F9F"
)
forest_importance_plot <- make_importance_plot(
    selected_forest_importance,
    title = paste("Selected forest:", selected_tree_type),
    fill = "#4B7F52"
)

gridExtra::grid.arrange(tree_importance_plot,
    forest_importance_plot,
    ncol = 2
)
```


## 4.3 Simulation Study

### 4.3.1 Recovery of the Simulated Subgroup Structure


```{r results-simulation-tree, fig.width=13, fig.height=8}
#| label: fig-results-simulation-tree
#| fig-cap: "Tree recovered from simulated data with a known subgroup-generating mechanism."

plot_known_subgroup_tree(
  fit = known_subgroup_fit,
  data = known_subgroup_sim_dt,
  type = selected_tree_type
)
```

# Appendix

## Results

### Appendix R1: Exploratory Data Outputs

```{r appendix-weighted-baseline-table}
#| label: tbl-appendix-weighted-baseline
#| tbl-cap: "Weighted distribution of analysis variables by under-five mortality status."

if (!requireNamespace("table1", quietly = TRUE)) {
  stop("Install table1 to render the weighted appendix baseline table.", call. = FALSE)
}

baseline_dt <- copy(as.data.table(drc_results_objects$raw_congo_model_dt))
baseline_labels <- drc_results_objects$raw_congo_var_labels
baseline_vars <- setdiff(drc_results_objects$raw_congo_predictors, "reg")
baseline_dt[, u5_mortality := factor(
  fifelse(deadu5_num == 1, outcome_labels[["died"]], outcome_labels[["alive"]]),
  levels = c(outcome_labels[["alive"]], outcome_labels[["died"]])
)]

baseline_table_df <- as.data.frame(baseline_dt[, c(baseline_vars, "u5_mortality"), with = FALSE])
for (variable in baseline_vars) {
  table1::label(baseline_table_df[[variable]]) <- unname(baseline_labels[variable])
}

baseline_weights <- baseline_dt$sample_weight
render_weighted_categorical <- function(x, ...) {
  row_index <- table1:::indices.indexed(x)
  if (is.null(row_index)) {
    row_index <- seq_along(x)
  }

  keep <- !is.na(x)
  x_keep <- x[keep]
  w_keep <- baseline_weights[row_index][keep]
  levels_keep <- if (is.factor(x)) levels(x) else sort(unique(as.character(x_keep)))

  out <- vapply(levels_keep, function(level) {
    weighted_n <- sum(w_keep[as.character(x_keep) == level], na.rm = TRUE)
    weighted_denominator <- sum(w_keep, na.rm = TRUE)
    weighted_percent <- if (weighted_denominator > 0) {
      100 * weighted_n / weighted_denominator
    } else {
      NA_real_
    }
    sprintf(
      "%s (%.1f%%)",
      format(round(weighted_n, 1), big.mark = ",", scientific = FALSE),
      weighted_percent
    )
  }, character(1L))
  names(out) <- levels_keep
  c(" " = "", out)
}

render_weighted_strata <- function(strata, ...) {
  out <- vapply(seq_along(strata), function(i) {
    row_index <- table1:::indices.indexed(strata[[i]])
    weighted_n <- sum(baseline_weights[row_index], na.rm = TRUE)
    sprintf(
      "%s\n(weighted N=%s)",
      names(strata)[i],
      format(round(weighted_n), big.mark = ",", scientific = FALSE)
    )
  }, character(1L))
  names(out) <- out
  out
}

baseline_formula <- as.formula(
  paste("~", paste(baseline_vars, collapse = " + "), "| u5_mortality")
)
baseline_table1 <- table1::table1(
  baseline_formula,
  data = table1::indexed(baseline_table_df),
  overall = "Overall",
  rowlabelhead = "Variable",
  render.categorical = render_weighted_categorical,
  render.strat = render_weighted_strata
)
baseline_table_out <- as.data.frame(baseline_table1)
names(baseline_table_out) <- c(
  "Variable",
  outcome_labels[["alive"]],
  outcome_labels[["died"]],
  "Overall"
)

baseline_table_kable <- report_compact_table(
  baseline_table_out,
  caption = "Weighted distribution of analysis variables by under-five mortality status",
  align = c("l", "r", "r", "r"),
  column_widths = c("4.5cm", rep("3.1cm", 3L)),
  escape = TRUE,
  bold_header = FALSE
)

if (requireNamespace("kableExtra", quietly = TRUE)) {
  baseline_table_kable <- kableExtra::add_header_above(
    baseline_table_kable,
    c(" " = 1, "Under-five mortality" = 2, "Overall" = 1),
    bold = TRUE
  )
  baseline_table_kable <- kableExtra::row_spec(baseline_table_kable, 0, bold = TRUE)
}

baseline_table_kable
```

```{r appendix-province-mortality-distribution, fig.width=9.2, fig.height=6.2}
#| label: fig-appendix-province-mortality-distribution
#| fig-cap: "Weighted province distribution by under-five mortality status."

province_plot_dt <- copy(as.data.table(drc_results_objects$raw_congo_model_dt))
province_plot_dt[, u5_mortality := fifelse(
  deadu5_num == 1,
  outcome_labels[["died"]],
  outcome_labels[["alive"]]
)]
province_plot_dt <- province_plot_dt[
  ,
  .(weighted_n = sum(sample_weight, na.rm = TRUE)),
  by = .(reg, u5_mortality)
]
province_plot_dt[
  ,
  weighted_percent := 100 * weighted_n / sum(weighted_n),
  by = reg
]
province_order <- province_plot_dt[
  u5_mortality == outcome_labels[["died"]]
][
  order(-weighted_percent)
]$reg
province_plot_dt[, reg := factor(reg, levels = province_order)]

ggplot(
  province_plot_dt,
  aes(x = reg, y = weighted_percent, fill = u5_mortality)
) +
  geom_col(width = 0.72) +
  coord_flip() +
  scale_fill_manual(
    values = setNames(c("#3268A8", "#B94E48"), c(outcome_labels[["alive"]], outcome_labels[["died"]]))
  ) +
  labs(x = NULL, y = "Weighted percent within province", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

### Appendix R2: Regression and Predictive-Forest Details

```{r appendix-regression-model-summary}
#| label: tbl-appendix-regression-model-summary
#| tbl-cap: "Survey-weighted regression model summary."

regression_summary_out <- copy(drc_results_objects$regression_model_summary)
if (!nrow(regression_summary_out) || "note" %in% names(regression_summary_out)) {
  regression_summary_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd after installing broom."
  )
} else {
  regression_summary_out <- regression_summary_out[
    ,
    .(
      term = term_label,
      estimate,
      std_error = std.error,
      statistic,
      p_value = format.pval(p.value, digits = 3, eps = 0.001),
      conf_low = conf.low,
      conf_high = conf.high
    )
  ]
  setnames(
    regression_summary_out,
    c("term", "estimate", "std_error", "statistic", "p_value", "conf_low", "conf_high"),
    c("Term", "Estimate", "Std. error", "Statistic", "p-value", "Conf. low", "Conf. high")
  )
}

report_compact_table(
  regression_summary_out,
  digits = 3,
  caption = "Survey-weighted regression model summary",
  column_widths = c("4.5cm", rep(NA_character_, 6L))
)
```

```{r appendix-ranger-cross-validation}
#| label: tbl-appendix-ranger-cross-validation
#| tbl-cap: "Cross-validation results for the predictive ranger model."

ranger_cv_out <- copy(drc_results_objects$predictive_forest_tuning)
if (!nrow(ranger_cv_out)) {
  ranger_cv_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with run_shap: true to tune the predictive ranger."
  )
} else {
  ranger_cv_out[
    ,
    rank := if (.BY$.metric == "mn_log_loss") {
      frank(mean, ties.method = "first")
    } else {
      frank(-mean, ties.method = "first")
    },
    by = .metric
  ]
  ranger_cv_out <- ranger_cv_out[
    order(.metric, rank),
    .(
      metric = .metric,
      mtry,
      min_n,
      mean,
      std_error = std_err,
      folds = n,
      rank
    )
  ]
  setnames(
    ranger_cv_out,
    c("metric", "mtry", "min_n", "mean", "std_error", "folds", "rank"),
    c("Metric", "mtry", "min_n", "Mean", "Std. error", "Folds", "Rank")
  )
}

report_compact_table(
  ranger_cv_out,
  digits = 4,
  caption = "Cross-validation results for the predictive ranger model",
  column_widths = c("2.4cm", rep(NA_character_, 6L))
)
```

```{r appendix-shap-summary-plot, fig.width=8.2, fig.height=5.8}
#| label: fig-appendix-shap-summary
#| fig-cap: "SHAP summary plot for the predictive ranger model."

shap_plot_dt <- copy(as.data.table(drc_results_objects$shap_summary_long))
if (!nrow(shap_plot_dt)) {
  ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = "Run Generate_DRC_Results_Objects.qmd with run_shap: true to compute SHAP values."
    ) +
    theme_void(base_size = 11)
} else {
  top_shap_features <- shap_plot_dt[
    ,
    .(mean_abs_shap = mean(abs(shap_value), na.rm = TRUE)),
    by = .(feature, feature_label)
  ][
    order(-mean_abs_shap)
  ][
    seq_len(min(.N, 12L))
  ]
  shap_plot_dt <- shap_plot_dt[feature %in% top_shap_features$feature]
  shap_plot_dt <- merge(
    shap_plot_dt,
    top_shap_features[, .(feature, feature_label, mean_abs_shap)],
    by = c("feature", "feature_label"),
    all.x = TRUE,
    sort = FALSE
  )
  top_shap_features[
    ,
    feature_axis_label := sprintf("%s", feature_label)
  ]
  shap_plot_dt <- merge(
    shap_plot_dt,
    top_shap_features[, .(feature, feature_axis_label)],
    by = "feature",
    all.x = TRUE,
    sort = FALSE
  )
  shap_plot_dt[
    ,
    feature_axis_label := factor(
      feature_axis_label,
      levels = rev(top_shap_features$feature_axis_label)
    )
  ]

  ggplot(
    shap_plot_dt,
    aes(x = shap_value, y = feature_axis_label, color = feature_value_scaled)
  ) +
    geom_vline(xintercept = 0, color = "black", linewidth = 0.35) +
    geom_point(
      position = position_jitter(width = 0, height = 0.18),
      size = 1.25,
      alpha = 0.75
    ) +
    scale_color_gradient(
      name = "Feature value",
      low = "#FDBE3B",
      high = "#6A00C2",
      breaks = c(0, 1),
      labels = c("Low", "High")
    ) +
    labs(x = "SHAP value (impact on model output)", y = NULL) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    )
}
```

### Appendix R3: Concentration-Index Tree Details

```{r appendix-tree-best-controls}
#| label: tbl-appendix-tree-best-controls
#| tbl-cap: "Best tree hyperparameters selected by validation gain within each impurity criterion."

tree_best_controls_out <- copy(drc_results_objects$tree_best_controls %||% data.table())
if (!nrow(tree_best_controls_out)) {
  tree_best_controls_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with RUN_CV=true"
  )
}

report_compact_table(
  tree_best_controls_out,
  digits = 4,
  caption = "Best tree hyperparameters selected by validation gain within each impurity criterion"
)
```

```{r appendix-tree-selected-setting-summary}
#| label: tbl-appendix-tree-selected-setting-summary
#| tbl-cap: "Tree summary for the selected settings."

tree_fit_table_out <- copy(drc_results_objects$tree_fit_table %||% data.table())
if (!nrow(tree_fit_table_out)) {
  tree_fit_table_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with RUN_CV=true"
  )
}

report_compact_table(
  tree_fit_table_out,
  digits = 4,
  caption = "Tree summary for the selected settings"
)
```

```{r appendix-tree-selection}
#| label: tbl-appendix-tree-selection
#| tbl-cap: "Selection metrics for fitted concentration-index trees."

tree_selection_out <- copy(drc_results_objects$tree_selection)
tree_selection_out <- tree_selection_out[
  ,
  .(
    criterion = type,
    root_objective = mean_root_objective,
    train_gain = mean_train_gain,
    train_relative_gain = mean_train_relative_gain,
    validation_gain = mean_validation_gain,
    validation_relative_gain = mean_validation_relative_gain,
    terminal_nodes
  )
]
setnames(
  tree_selection_out,
  c(
    "criterion",
    "root_objective",
    "train_gain",
    "train_relative_gain",
    "validation_gain",
    "validation_relative_gain",
    "terminal_nodes"
  ),
  c(
    "Criterion",
    "Root objective",
    "Train gain",
    "Train gain/root",
    "Validation gain",
    "Validation gain/root",
    "Terminal nodes"
  )
)

report_compact_table(
  tree_selection_out,
  digits = 4,
  caption = "Selection metrics for fitted concentration-index trees",
  column_widths = c("1.4cm", rep("1.8cm", 6L))
)
```

```{r appendix-tree-complexity-metrics}
#| label: tbl-appendix-tree-complexity-metrics
#| tbl-cap: "Collected tree complexity metrics from cross-validation."

tree_complexity_out <- copy(drc_results_objects$tree_complexity_metrics %||% data.table())
if (!nrow(tree_complexity_out)) {
  tree_complexity_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with RUN_CV=true"
  )
} else {
  tree_complexity_out <- tree_complexity_out[
    ,
    .(
      type,
      maxdepth,
      min_relative_gain,
      mean_terminal_nodes,
      mean_train_relative_gain,
      mean_validation_relative_gain,
      std_err_validation_relative_gain
    )
  ]
}

report_compact_table(
  tree_complexity_out,
  digits = 4,
  caption = "Collected tree complexity metrics from cross-validation"
)
```

```{r appendix-tree-validation-path-table}
#| label: tbl-appendix-tree-validation-path
#| tbl-cap: "Average tree validation percentage and complexity by minimum relative split gain."

tree_path_out <- copy(drc_results_objects$tree_min_relative_gain_path %||% data.table())
if (!nrow(tree_path_out)) {
  tree_path_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with RUN_CV=true"
  )
} else {
  tree_path_out <- tree_path_out[
    ,
    .(
      type,
      min_relative_gain,
      n_settings,
      mean_root_objective,
      mean_validation_gain,
      validation_percent_from_mean_root,
      mean_terminal_nodes,
      validation_percent_gain,
      validation_percent_se,
      selected
    )
  ]
}

report_compact_table(
  tree_path_out,
  digits = 4,
  caption = "Average tree validation percentage and complexity by minimum relative split gain"
)
```

```{r appendix-tree-validation-path-plots, fig.width=12, fig.height=4.8}
#| label: fig-appendix-tree-validation-path
#| fig-cap: "Tree validation gain and complexity by minimum relative split gain."
#| fig-pos: "H"

tree_path_plot_dt <- copy(drc_results_objects$tree_min_relative_gain_path %||% data.table())
gridExtra::grid.arrange(
  plot_validation_percent_gain_path(
    tree_path_plot_dt,
    title = "Tree Validation Gain (% Root)"
  ),
  plot_terminal_nodes_path(
    tree_path_plot_dt,
    title = "Tree Complexity"
  ),
  ncol = 2
)
```

```{r appendix-tree-variable-importance-plot, fig.width=10, fig.height=7}
#| label: fig-appendix-tree-variable-importance
#| fig-cap: "Top tree variables by concentration-index criterion."
#| fig-pos: "H"

plot_importance_by_criterion(
  drc_results_objects$tree_variable_importance,
  "Top Tree Variables by Concentration-Index Criterion",
  fill = "#3268A8"
)
```

```{r appendix-ci-tree-ci, fig.width=7.2, fig.height=5.0}
#| label: fig-appendix-ci-tree-ci
#| fig-cap: "Concentration-index tree fitted with the CI criterion."
#| fig-pos: "H"

if (isTRUE(params$render_appendix_tree_plots) && "CI" %in% names(tree_models_by_type)) {
  ci_report_tree_plot(
    fit = tree_models_by_type[["CI"]],
    data = congo_model_dt,
    outcome_name = "deadu5_num",
    outcome_label = "U5 death",
    ci_type = "CI",
    var_labels = congo_var_labels,
    fontsize = 5.8
  )
}
```

```{r appendix-ci-tree-cig, fig.width=7.2, fig.height=5.0}
#| label: fig-appendix-ci-tree-cig
#| fig-cap: "Concentration-index tree fitted with the CIg criterion."
#| fig-pos: "H"

if (isTRUE(params$render_appendix_tree_plots) && "CIg" %in% names(tree_models_by_type)) {
  ci_report_tree_plot(
    fit = tree_models_by_type[["CIg"]],
    data = congo_model_dt,
    outcome_name = "deadu5_num",
    outcome_label = "U5 death",
    ci_type = "CIg",
    var_labels = congo_var_labels,
    fontsize = 5.8
  )
}
```

```{r appendix-ci-tree-cic, fig.width=7.2, fig.height=5.0}
#| label: fig-appendix-ci-tree-cic
#| fig-cap: "Concentration-index tree fitted with the CIc criterion."
#| fig-pos: "H"

if (isTRUE(params$render_appendix_tree_plots) && "CIc" %in% names(tree_models_by_type)) {
  ci_report_tree_plot(
    fit = tree_models_by_type[["CIc"]],
    data = congo_model_dt,
    outcome_name = "deadu5_num",
    outcome_label = "U5 death",
    ci_type = "CIc",
    var_labels = congo_var_labels,
    fontsize = 5.8
  )
}
```

```{r appendix-ci-tree-l, fig.width=7.2, fig.height=5.0}
#| label: fig-appendix-ci-tree-l
#| fig-cap: "Concentration-index tree fitted with the L criterion."
#| fig-pos: "H"

if (isTRUE(params$render_appendix_tree_plots) && "L" %in% names(tree_models_by_type)) {
  ci_report_tree_plot(
    fit = tree_models_by_type[["L"]],
    data = congo_model_dt,
    outcome_name = "deadu5_num",
    outcome_label = "U5 death",
    ci_type = "L",
    var_labels = congo_var_labels,
    fontsize = 5.8
  )
}
```

### Appendix R4: Concentration-Index Forest Details

```{r appendix-forest-best-controls}
#| label: tbl-appendix-forest-best-controls
#| tbl-cap: "Best forest hyperparameters selected by validation gain within each impurity criterion."

forest_best_controls_out <- copy(drc_results_objects$forest_best_controls %||% data.table())
if (!nrow(forest_best_controls_out)) {
  forest_best_controls_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with RUN_FOREST=true"
  )
}

report_compact_table(
  forest_best_controls_out,
  digits = 4,
  caption = "Best forest hyperparameters selected by validation gain within each impurity criterion"
)
```

```{r appendix-forest-selected-setting-summary}
#| label: tbl-appendix-forest-selected-setting-summary
#| tbl-cap: "Forest summary for the selected settings."

forest_fit_table_out <- copy(drc_results_objects$forest_fit_table %||% data.table())
if (!nrow(forest_fit_table_out)) {
  forest_fit_table_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with RUN_FOREST=true"
  )
}

report_compact_table(
  forest_fit_table_out,
  digits = 4,
  caption = "Forest summary for the selected settings"
)
```

```{r appendix-forest-selection}
#| label: tbl-appendix-forest-selection
#| tbl-cap: "Selection metrics for fitted concentration-index forests."

forest_selection_out <- copy(drc_results_objects$forest_selection)
if (is.null(forest_selection_out) || !nrow(forest_selection_out)) {
  forest_selection_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with RUN_FOREST=true"
  )
} else {
  forest_selection_out <- forest_selection_out[
    ,
    .(
      criterion = type,
      trees = ntree,
      mtry,
      root_objective = mean_root_objective,
      train_gain = mean_train_gain,
      train_relative_gain = mean_train_relative_gain,
      validation_gain = mean_validation_gain,
      validation_relative_gain = mean_validation_relative_gain,
      mean_terminal_nodes,
      mean_max_depth
    )
  ]
  setnames(
    forest_selection_out,
    c(
      "criterion",
      "trees",
      "mtry",
      "root_objective",
      "train_gain",
      "train_relative_gain",
      "validation_gain",
      "validation_relative_gain",
      "mean_terminal_nodes",
      "mean_max_depth"
    ),
    c(
      "Criterion",
      "Trees",
      "mtry",
      "Root objective",
      "Train gain",
      "Train gain/root",
      "Validation gain",
      "Validation gain/root",
      "Mean terminal nodes",
      "Mean max depth"
    )
  )
}

report_compact_table(
  forest_selection_out,
  digits = 4,
  caption = "Selection metrics for fitted concentration-index forests"
)
```

```{r appendix-forest-complexity-metrics}
#| label: tbl-appendix-forest-complexity-metrics
#| tbl-cap: "Collected forest complexity metrics from cross-validation."

forest_complexity_out <- copy(drc_results_objects$forest_complexity_metrics %||% data.table())
if (!nrow(forest_complexity_out)) {
  forest_complexity_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with RUN_FOREST_CV=true"
  )
} else {
  forest_complexity_out <- forest_complexity_out[
    ,
    .(
      type,
      ntree,
      maxdepth,
      min_relative_gain,
      mean_terminal_nodes,
      mean_train_relative_gain,
      mean_validation_relative_gain,
      std_err_validation_relative_gain
    )
  ]
}

report_compact_table(
  forest_complexity_out,
  digits = 4,
  caption = "Collected forest complexity metrics from cross-validation"
)
```

```{r appendix-forest-validation-path-table}
#| label: tbl-appendix-forest-validation-path
#| tbl-cap: "Average forest validation percentage and complexity by minimum relative split gain."

forest_path_out <- copy(drc_results_objects$forest_min_relative_gain_path %||% data.table())
if (!nrow(forest_path_out)) {
  forest_path_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with RUN_FOREST_CV=true"
  )
} else {
  forest_path_out <- forest_path_out[
    ,
    .(
      type,
      min_relative_gain,
      n_settings,
      mean_root_objective,
      mean_validation_gain,
      validation_percent_from_mean_root,
      mean_terminal_nodes,
      validation_percent_gain,
      validation_percent_se,
      selected
    )
  ]
}

report_compact_table(
  forest_path_out,
  digits = 4,
  caption = "Average forest validation percentage and complexity by minimum relative split gain"
)
```

```{r appendix-forest-validation-path-plots, fig.width=12, fig.height=4.8}
#| label: fig-appendix-forest-validation-path
#| fig-cap: "Forest validation gain and complexity by minimum relative split gain."
#| fig-pos: "H"

forest_path_plot_dt <- copy(drc_results_objects$forest_min_relative_gain_path %||% data.table())
gridExtra::grid.arrange(
  plot_validation_percent_gain_path(
    forest_path_plot_dt,
    title = "Forest Validation Gain (% Root)"
  ),
  plot_terminal_nodes_path(
    forest_path_plot_dt,
    title = "Forest Complexity"
  ),
  ncol = 2
)
```

```{r appendix-forest-variable-importance-plot, fig.width=10, fig.height=7}
#| label: fig-appendix-forest-variable-importance
#| fig-cap: "Top forest variables by concentration-index criterion."
#| fig-pos: "H"

plot_importance_by_criterion(
  drc_results_objects$forest_variable_importance,
  "Top Forest Variables by Concentration-Index Criterion",
  fill = "#4B7F52"
)
```

#### Non-Selected Forest Surrogate Trees

The selected forest surrogate appears in the results section. The remaining
criterion-specific forest surrogates are retained here for comparison.

```{r appendix-forest-surrogate-ci, fig.width=12, fig.height=7}
#| label: fig-appendix-forest-surrogate-ci
#| fig-cap: "Surrogate tree for the CI concentration-index forest."
#| fig-pos: "H"

if (!identical(selected_tree_type, "CI")) {
  plot_forest_surrogate("CI")
}
```

```{r appendix-forest-surrogate-cig, fig.width=12, fig.height=7}
#| label: fig-appendix-forest-surrogate-cig
#| fig-cap: "Surrogate tree for the CIg concentration-index forest."
#| fig-pos: "H"

if (!identical(selected_tree_type, "CIg")) {
  plot_forest_surrogate("CIg")
}
```

```{r appendix-forest-surrogate-cic, fig.width=12, fig.height=7}
#| label: fig-appendix-forest-surrogate-cic
#| fig-cap: "Surrogate tree for the CIc concentration-index forest."
#| fig-pos: "H"

if (!identical(selected_tree_type, "CIc")) {
  plot_forest_surrogate("CIc")
}
```

```{r appendix-forest-surrogate-l, fig.width=12, fig.height=7}
#| label: fig-appendix-forest-surrogate-l
#| fig-cap: "Surrogate tree for the L concentration-index forest."
#| fig-pos: "H"

if (!identical(selected_tree_type, "L")) {
  plot_forest_surrogate("L")
}
```

```{r appendix-forest-variable-importance}
#| label: tbl-appendix-forest-variable-importance
#| tbl-cap: "Forest variable importance by concentration-index criterion."

forest_importance_out <- copy(drc_results_objects$forest_variable_importance)
if (!nrow(forest_importance_out)) {
  forest_importance_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with run_forest: true"
  )
} else {
  forest_importance_out[, variable_label := format_variable_label(variable)]
  forest_importance_out <- forest_importance_out[
    ,
    .(criterion, rank, variable = variable_label, importance)
  ]
}

report_compact_table(
  forest_importance_out,
  digits = 4,
  caption = "Forest variable importance by concentration-index criterion"
)
```

```{r appendix-surrogate-variable-importance}
#| label: tbl-appendix-surrogate-variable-importance
#| tbl-cap: "Surrogate-tree variable importance for the selected concentration-index forest."

surrogate_importance_out <- extract_surrogate_importance(
  drc_results_objects$selected_forest_surrogate
)
if (!nrow(surrogate_importance_out)) {
  surrogate_importance_out <- data.table(
    note = "Run Generate_DRC_Results_Objects.qmd with RUN_FOREST=true"
  )
} else {
  surrogate_importance_out[, variable_label := format_variable_label(variable)]
  surrogate_importance_out[
    ,
    rank := seq_len(.N)
  ]
  surrogate_importance_out <- surrogate_importance_out[
    ,
    .(rank, variable = variable_label, importance)
  ]
}

report_compact_table(
  surrogate_importance_out,
  digits = 4,
  caption = "Surrogate-tree variable importance for the selected concentration-index forest"
)
```

### Appendix R5: Implementation Status

Implemented in the report: selected concentration-index tree and selected
forest/surrogate-tree variable-importance plots in the results section; tree
and concentration-index forest best-control tables, selected-setting validation
summaries, complexity metrics, validation-path plots, and variable-importance
outputs by criterion in the appendix; ranger cross-validation and SHAP summary
outputs in the appendix.

To implement for the final analysis run: rerun the generator with `RUN_CV=true`
and `RUN_FOREST_CV=true` when the full cross-validation objects should be
recomputed from scratch rather than read from previously generated results.
