---
title: "DRC Selected Tree Bootstrap Stability"
author: "Moses Mburu"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
params:
  bootstrap_replicates: 1000
  bootstrap_seed: 20260615
  bootstrap_unit: "PSU"
  refresh_bootstrap: false
  save_results: true
---

```{r setup}
knitr::opts_chunk$set(
  fig.width = 9,
  fig.height = 5.5,
  dpi = 500,
  dev = "png"
)

library(data.table)
library(ggplot2)
library(here)
library(ineqTrees)
library(knitr)
library(partykit)

source(here("R", "report_helpers.R"))
source(here("R", "ci_rpart_comparison_methods.R"))
```

# Purpose

This notebook does not retune the concentration-index trees. The generator QMD
now handles design-aware validation using PSU fold IDs. This file only checks
the stability of the already selected tree settings by refitting those settings
under bootstrap samples.

The selected tuning parameters are loaded from the saved results object. Each
bootstrap sample refits the same tree criterion and the same selected controls,
then compares the refitted tree with the corresponding selected tree from the
generator.

The default bootstrap unit is the primary sampling unit (PSU). This keeps the
stability check aligned with the clustered DHS design and avoids treating all
children as independent row-level draws. The bootstrap is still a sensitivity
analysis for tree structure. It is not a replicate-weight analysis and it does
not produce formal survey-design confidence intervals for individual splits.

# Load Generated Results

```{r load-generated-results}
results_object_file <- here("data", "drc_report_results_objects.rda")
stopifnot(file.exists(results_object_file))

load(results_object_file)
stopifnot(exists("drc_results_objects"))
list2env(drc_results_objects, envir = environment())

required_objects <- c(
  "congo_model_dt",
  "raw_congo_model_dt",
  "congo_ci_formula",
  "congo_var_labels",
  "tree_best_by_type"
)
missing_objects <- setdiff(required_objects, names(drc_results_objects))
if (length(missing_objects)) {
  stop(
    "Missing required generated objects: ",
    paste(missing_objects, collapse = ", "),
    call. = FALSE
  )
}

stopifnot(nrow(congo_model_dt) == nrow(raw_congo_model_dt))
if (!"PSU" %in% names(raw_congo_model_dt)) {
  stop(
    "The generated data do not contain `PSU`. Rerender the generator QMD ",
    "so that `raw_congo_model_dt` includes the DHS cluster variable.",
    call. = FALSE
  )
}

criterion_types <- intersect(c("CI", "CIg", "CIc", "L"), tree_best_by_type$type)
if (!length(criterion_types)) {
  stop("No selected tree criteria were found in `tree_best_by_type`.", call. = FALSE)
}
bootstrap_unit <- toupper(as.character(params$bootstrap_unit)[1L])
if (!bootstrap_unit %in% c("PSU", "ROW")) {
  stop("`bootstrap_unit` must be either 'PSU' or 'ROW'.", call. = FALSE)
}

bootstrap_replicates <- max(1L, as.integer(params$bootstrap_replicates))
bootstrap_seed <- as.integer(params$bootstrap_seed)

bootstrap_source_dt <- data.table(
  PSU = raw_congo_model_dt$PSU,
  congo_model_dt
)

loaded_summary <- data.table(
  item = c(
    "Generated object",
    "Rows",
    "PSUs",
    "Selected criteria",
    "Bootstrap unit",
    "Bootstrap replicates"
  ),
  value = c(
    results_object_file,
    format(nrow(congo_model_dt), big.mark = ","),
    format(uniqueN(raw_congo_model_dt$PSU), big.mark = ","),
    paste(criterion_types, collapse = ", "),
    bootstrap_unit,
    format(bootstrap_replicates, big.mark = ",")
  )
)

report_compact_table(
  loaded_summary,
  caption = "Generated results used for selected-tree bootstrap stability checks"
)
```

# Selected Settings

```{r selected-tree-settings}
selected_tree_settings <- copy(tree_best_by_type[type %in% criterion_types])

report_compact_table(
  selected_tree_settings[
    ,
    .(
      type,
      grid_id,
      minsplit,
      minbucket,
      minprob,
      maxdepth,
      min_gain,
      min_relative_gain,
      mean_score,
      mean_terminal_nodes
    )
  ],
  digits = 4,
  caption = "Selected tuning settings held fixed across bootstrap refits"
)
```

```{r fit-reference-trees}
fit_selected_tree <- function(setting, data) {
  ci_tree(
    formula = congo_ci_formula,
    data = data,
    rank_name = "wealth",
    outcome_name = "deadu5_num",
    weights = data[["sample_weight"]],
    type = setting$type[1L],
    control = ci_control_from_row(setting)
  )
}

have_saved_tree_models <- exists("tree_models_by_type") &&
  all(criterion_types %in% names(tree_models_by_type))

reference_tree_models_by_type <- if (have_saved_tree_models) {
  tree_models_by_type[criterion_types]
} else {
  setNames(
    lapply(seq_len(nrow(selected_tree_settings)), function(i) {
      fit_selected_tree(selected_tree_settings[i], congo_model_dt)
    }),
    selected_tree_settings$type
  )
}

reference_tree_summary <- rbindlist(
  lapply(names(reference_tree_models_by_type), function(type) {
    out <- as.data.table(
      ci_tree_terminal_summary(reference_tree_models_by_type[[type]])
    )
    out[, criterion := type]
    setcolorder(out, "criterion")
    out
  }),
  fill = TRUE
)

report_compact_table(
  reference_tree_summary,
  digits = 4,
  caption = "Reference selected-tree terminal-node summaries before bootstrap refitting"
)
```

# Bootstrap Stability

```{r bootstrap-helpers}
scalar_numeric <- function(x) {
  if (is.null(x) || !length(x)) {
    return(NA_real_)
  }
  x <- suppressWarnings(as.numeric(x[1L]))
  if (is.na(x) || !is.finite(x)) NA_real_ else x
}

scalar_character <- function(x) {
  if (is.null(x) || !length(x) || is.na(x[1L]) || !nzchar(as.character(x[1L]))) {
    return(NA_character_)
  }
  as.character(x[1L])
}

label_tree_variable <- function(x) {
  if (is.na(x) || !nzchar(x)) {
    return(NA_character_)
  }
  if (x %in% names(congo_var_labels)) {
    return(unname(congo_var_labels[x]))
  }
  x
}

collapse_split_variables <- function(x) {
  x <- unique(x[!is.na(x) & nzchar(x)])
  if (!length(x)) {
    return(NA_character_)
  }
  paste(x, collapse = "; ")
}

collect_split_variables <- function(node) {
  if (partykit::is.terminal(node)) {
    return(character())
  }

  info <- partykit::info_node(node)
  this_var <- scalar_character(info$varname)
  kids <- partykit::kids_node(node)
  child_vars <- if (is.null(kids) || !length(kids)) {
    character()
  } else {
    unlist(lapply(kids, collect_split_variables), use.names = FALSE)
  }

  c(this_var, child_vars)
}

terminal_node_summary <- function(fit) {
  out <- as.data.table(ci_tree_terminal_summary(fit))
  out[, abs_impurity := abs(ci)]
  out
}

tree_leaf_similarity <- function(reference_fit,
                                 bootstrap_fit,
                                 reference_data,
                                 recovery_threshold = 0.50) {
  reference_leaf <- as.integer(partykit::predict.party(
    reference_fit,
    newdata = as.data.frame(reference_data),
    type = "node"
  ))
  bootstrap_leaf <- as.integer(partykit::predict.party(
    bootstrap_fit,
    newdata = as.data.frame(reference_data),
    type = "node"
  ))

  reference_leaves <- sort(unique(reference_leaf))
  bootstrap_leaves <- sort(unique(bootstrap_leaf))

  best_jaccard <- vapply(reference_leaves, function(reference_id) {
    reference_rows <- reference_leaf == reference_id
    max(vapply(bootstrap_leaves, function(bootstrap_id) {
      bootstrap_rows <- bootstrap_leaf == bootstrap_id
      sum(reference_rows & bootstrap_rows) / sum(reference_rows | bootstrap_rows)
    }, numeric(1L)))
  }, numeric(1L))

  list(
    tree_similarity_score = stats::median(best_jaccard, na.rm = TRUE),
    reference_leaves_recovered_percent =
      100 * mean(best_jaccard >= recovery_threshold, na.rm = TRUE)
  )
}

draw_bootstrap_data <- function(unit) {
  if (identical(unit, "PSU")) {
    sampled_psus <- sample(
      unique(bootstrap_source_dt$PSU),
      size = uniqueN(bootstrap_source_dt$PSU),
      replace = TRUE
    )
    boot_key <- data.table(
      bootstrap_draw_id = seq_along(sampled_psus),
      PSU = sampled_psus
    )
    out <- bootstrap_source_dt[boot_key, on = "PSU", allow.cartesian = TRUE]
    out[, bootstrap_draw_id := NULL]
    out[, PSU := NULL]
    return(out)
  }

  congo_model_dt[sample.int(nrow(congo_model_dt), size = nrow(congo_model_dt), replace = TRUE)]
}

summarise_bootstrap_tree <- function(fit, type, replicate, reference_fit) {
  root_info <- partykit::info_node(partykit::node_party(fit))
  terminal_summary <- terminal_node_summary(fit)
  split_variables <- collect_split_variables(partykit::node_party(fit))
  similarity <- tree_leaf_similarity(
    reference_fit = reference_fit,
    bootstrap_fit = fit,
    reference_data = congo_model_dt
  )

  weighted_terminal_impurity <- if (nrow(terminal_summary)) {
    sum(
      terminal_summary$abs_impurity * terminal_summary$weight,
      na.rm = TRUE
    ) / sum(terminal_summary$weight, na.rm = TRUE)
  } else {
    NA_real_
  }

  data.table(
    type = type,
    bootstrap_replicate = replicate,
    bootstrap_unit = bootstrap_unit,
    success = TRUE,
    error_message = NA_character_,
    terminal_nodes = nrow(terminal_summary),
    no_split = nrow(terminal_summary) <= 1L,
    root_split_variable = scalar_character(root_info$varname),
    root_split_variable_label = label_tree_variable(scalar_character(root_info$varname)),
    all_split_variables = collapse_split_variables(split_variables),
    all_split_variable_labels = collapse_split_variables(vapply(
      split_variables,
      label_tree_variable,
      character(1L)
    )),
    root_impurity = abs(scalar_numeric(root_info$ci)),
    weighted_terminal_impurity = weighted_terminal_impurity,
    absolute_gain = scalar_numeric(root_info$gain),
    relative_gain = scalar_numeric(root_info$relative_gain),
    terminal_mortality_min = min(terminal_summary$outcome_mean, na.rm = TRUE),
    terminal_mortality_median = stats::median(terminal_summary$outcome_mean, na.rm = TRUE),
    terminal_mortality_max = max(terminal_summary$outcome_mean, na.rm = TRUE),
    terminal_abs_impurity_min = min(terminal_summary$abs_impurity, na.rm = TRUE),
    terminal_abs_impurity_median = stats::median(terminal_summary$abs_impurity, na.rm = TRUE),
    terminal_abs_impurity_max = max(terminal_summary$abs_impurity, na.rm = TRUE),
    tree_similarity_score = similarity$tree_similarity_score,
    reference_leaves_recovered_percent = similarity$reference_leaves_recovered_percent
  )
}

failed_bootstrap_tree <- function(type, replicate, error_message) {
  data.table(
    type = type,
    bootstrap_replicate = replicate,
    bootstrap_unit = bootstrap_unit,
    success = FALSE,
    error_message = error_message,
    terminal_nodes = NA_integer_,
    no_split = NA,
    root_split_variable = NA_character_,
    root_split_variable_label = NA_character_,
    all_split_variables = NA_character_,
    all_split_variable_labels = NA_character_,
    root_impurity = NA_real_,
    weighted_terminal_impurity = NA_real_,
    absolute_gain = NA_real_,
    relative_gain = NA_real_,
    terminal_mortality_min = NA_real_,
    terminal_mortality_median = NA_real_,
    terminal_mortality_max = NA_real_,
    terminal_abs_impurity_min = NA_real_,
    terminal_abs_impurity_median = NA_real_,
    terminal_abs_impurity_max = NA_real_,
    tree_similarity_score = NA_real_,
    reference_leaves_recovered_percent = NA_real_
  )
}

format_interval <- function(x, digits = 3L, percent = FALSE) {
  x <- x[is.finite(x)]
  if (!length(x)) {
    return(NA_character_)
  }
  qs <- stats::quantile(x, probs = c(0.025, 0.5, 0.975), na.rm = TRUE)
  if (percent) {
    return(sprintf(
      "%s%% (%s%%, %s%%)",
      format(round(qs[[2L]], 1), nsmall = 1),
      format(round(qs[[1L]], 1), nsmall = 1),
      format(round(qs[[3L]], 1), nsmall = 1)
    ))
  }
  fmt <- paste0("%.", digits, "f (%.", digits, "f, %.", digits, "f)")
  sprintf(fmt, qs[[2L]], qs[[1L]], qs[[3L]])
}

format_count_interval <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) {
    return(NA_character_)
  }
  qs <- stats::quantile(x, probs = c(0.025, 0.5, 0.975), na.rm = TRUE)
  sprintf("%s (%s, %s)", round(qs[[2L]]), round(qs[[1L]]), round(qs[[3L]]))
}

most_common_root <- function(dt) {
  dt <- dt[success == TRUE & !is.na(root_split_variable_label)]
  if (!nrow(dt)) {
    return(NA_character_)
  }
  root_dt <- dt[
    ,
    .(n = .N),
    by = .(root_split_variable_label)
  ][order(-n, root_split_variable_label)]
  sprintf(
    "%s, %.1f%%",
    root_dt$root_split_variable_label[1L],
    100 * root_dt$n[1L] / nrow(dt)
  )
}

most_common_split_variables <- function(dt, max_vars = 5L) {
  dt <- dt[success == TRUE]
  if (!nrow(dt)) {
    return(NA_character_)
  }

  split_dt <- rbindlist(lapply(seq_len(nrow(dt)), function(i) {
    value <- dt$all_split_variable_labels[i]
    if (is.na(value) || !nzchar(value)) {
      return(NULL)
    }
    data.table(
      bootstrap_replicate = dt$bootstrap_replicate[i],
      variable = unique(strsplit(value, "; ", fixed = TRUE)[[1L]])
    )
  }), fill = TRUE)

  if (!nrow(split_dt)) {
    return(NA_character_)
  }

  split_dt <- split_dt[
    ,
    .(n = uniqueN(bootstrap_replicate)),
    by = variable
  ][order(-n, variable)]

  top_split_variables <- split_dt[
    seq_len(min(.N, max_vars)),
    paste0(variable, ", ", sprintf("%.1f%%", 100 * n / nrow(dt)))
  ]

  paste(top_split_variables, collapse = "; ")
}

summarise_bootstrap_table <- function(raw_dt) {
  raw_dt[
    ,
    {
      successful <- .SD[success == TRUE]
      .(
        successful_replicates = nrow(successful),
        failed_replicates = sum(success == FALSE),
        no_split_trees = if (nrow(successful)) {
          sprintf("%.1f%%", 100 * mean(successful$no_split, na.rm = TRUE))
        } else {
          NA_character_
        },
        terminal_nodes = format_count_interval(successful$terminal_nodes),
        most_common_root_split = most_common_root(.SD),
        most_common_split_variables = most_common_split_variables(.SD),
        absolute_gain = format_interval(successful$absolute_gain, digits = 3L),
        relative_gain = format_interval(successful$relative_gain, digits = 3L),
        tree_similarity_score = format_interval(
          successful$tree_similarity_score,
          digits = 3L
        ),
        recovered_reference_leaves = format_interval(
          successful$reference_leaves_recovered_percent,
          digits = 1L,
          percent = TRUE
        )
      )
    },
    by = type
  ][match(criterion_types, type)]
}
```

```{r run-bootstrap}
results_dir <- here("data")
bootstrap_label <- paste0(
  "selected_tree_",
  tolower(bootstrap_unit),
  "_bootstrap_B",
  bootstrap_replicates,
  "_seed",
  bootstrap_seed
)
bootstrap_tree_stability_raw_file <- file.path(
  results_dir,
  paste0(bootstrap_label, "_raw.rds")
)
bootstrap_tree_stability_table_file <- file.path(
  results_dir,
  paste0(bootstrap_label, "_table.rds")
)
bootstrap_tree_stability_table_csv <- file.path(
  results_dir,
  paste0(bootstrap_label, "_table.csv")
)

if (
  !isTRUE(params$refresh_bootstrap) &&
    file.exists(bootstrap_tree_stability_raw_file) &&
    file.exists(bootstrap_tree_stability_table_file)
) {
  bootstrap_tree_stability_raw <- readRDS(bootstrap_tree_stability_raw_file)
  bootstrap_tree_stability_table <- readRDS(bootstrap_tree_stability_table_file)
} else {
  set.seed(bootstrap_seed)
  bootstrap_rows <- vector("list", length(criterion_types) * bootstrap_replicates)
  row_id <- 0L

  for (criterion in criterion_types) {
    setting <- selected_tree_settings[type == criterion]
    reference_fit <- reference_tree_models_by_type[[criterion]]

    for (b in seq_len(bootstrap_replicates)) {
      row_id <- row_id + 1L
      bootstrap_dt <- draw_bootstrap_data(bootstrap_unit)

      bootstrap_rows[[row_id]] <- tryCatch({
        fit <- fit_selected_tree(setting, bootstrap_dt)
        summarise_bootstrap_tree(
          fit = fit,
          type = criterion,
          replicate = b,
          reference_fit = reference_fit
        )
      }, error = function(e) {
        failed_bootstrap_tree(
          type = criterion,
          replicate = b,
          error_message = conditionMessage(e)
        )
      })
    }
  }

  bootstrap_tree_stability_raw <- rbindlist(bootstrap_rows, fill = TRUE)
  bootstrap_tree_stability_table <- summarise_bootstrap_table(
    bootstrap_tree_stability_raw
  )

  saveRDS(bootstrap_tree_stability_raw, bootstrap_tree_stability_raw_file)
  saveRDS(bootstrap_tree_stability_table, bootstrap_tree_stability_table_file)
  fwrite(bootstrap_tree_stability_table, bootstrap_tree_stability_table_csv)
}

if (!file.exists(bootstrap_tree_stability_table_csv)) {
  fwrite(bootstrap_tree_stability_table, bootstrap_tree_stability_table_csv)
}

report_compact_table(
  bootstrap_tree_stability_table,
  caption = "Cluster-level bootstrap stability of selected concentration-index trees"
)
```

For tree similarity, each bootstrap tree is applied to the original analysis
data and compared with the corresponding selected tree from the generator.
Terminal-leaf membership is compared using best Jaccard overlap, and a reference
leaf is counted as recovered when its best overlap is at least 0.50.

The bootstrap analysis keeps the tree tuning settings fixed. It therefore checks
stability of the fitted subgroup structure after model selection, rather than
repeating the full tuning process. These results should be described in the
thesis as stability or sensitivity evidence. They should not be reported as
p-values, formal split-level inference, or full survey-design uncertainty.

# Save Objects

```{r save-bootstrap-objects}
bootstrap_tree_results <- list(
  results_object_file = results_object_file,
  bootstrap_seed = bootstrap_seed,
  bootstrap_replicates = bootstrap_replicates,
  bootstrap_unit = bootstrap_unit,
  selected_tree_settings = selected_tree_settings,
  reference_tree_summary = reference_tree_summary,
  bootstrap_tree_stability_raw_file = bootstrap_tree_stability_raw_file,
  bootstrap_tree_stability_table_file = bootstrap_tree_stability_table_file,
  bootstrap_tree_stability_table_csv = bootstrap_tree_stability_table_csv,
  bootstrap_tree_stability_table = bootstrap_tree_stability_table
)

bootstrap_tree_results_file <- here(
  "data",
  paste0(bootstrap_label, "_objects.rda")
)

if (isTRUE(params$save_results)) {
  save(bootstrap_tree_results, file = bootstrap_tree_results_file)
}

data.table(
  saved = isTRUE(params$save_results),
  file = bootstrap_tree_results_file,
  bootstrap_unit = bootstrap_unit,
  bootstrap_replicates = bootstrap_replicates
)
```
