---
title: "DRC Results Generator"
author: "Moses Mburu"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
    df-print: paged
    dpi: 500
  pdf:
    documentclass: article
    papersize: a4
    fontsize: 10pt
    geometry:
      - margin=25mm
    pdf-engine: xelatex
    toc: true
    number-sections: true
    fig-cap-location: bottom
    tbl-cap-location: top
execute:
  echo: true
  warning: false
  message: false
  fig-dpi: 500
params:
  analysis_rows: 1
  tuning_folds: 10
  undersample: 0.4
always_allow_html: true
---

```{r setup}
knitr::opts_chunk$set(
  fig.width = 9,
  fig.height = 5.5,
  dpi = 500,
  dev = "png"
)

library(data.table)
library(ggplot2)
library(grid)
library(knitr)
library(ineqTrees)
library(partykit)
library(here)
source(here("R", "report_helpers.R"))
source(here("R", "ci_known_subgroup_demo.R"))
source(here("R", "ci_rpart_comparison_methods.R"))

kableExtra::use_latex_packages()
```

```{r env-int-helper}
env_int <- function(name, default) {
  value <- suppressWarnings(as.integer(Sys.getenv(name, unset = NA_character_)))
  as.integer(ifelse(is.na(value) || value <= 0L, default, value))
}

env_num_or_null <- function(name, default = NULL) {
  value <- Sys.getenv(name, unset = NA_character_)
  if (is.na(value) || !nzchar(value)) {
    return(default)
  }

  parsed <- suppressWarnings(as.numeric(value))
  if (is.na(parsed)) {
    return(default)
  }
  parsed
}
```

```{r runtime-parameter-config}
default_params <- list(
  analysis_rows = 1L,
  tuning_folds = 10L,
  undersample = 0.41
)

runtime_params <- utils::modifyList(default_params, as.list(params))

runtime_params$analysis_rows <- env_int("ANALYSIS_ROWS", runtime_params$analysis_rows)
runtime_params$tuning_folds <- max(2L, env_int("TUNING_FOLDS", runtime_params$tuning_folds))
runtime_params$undersample <- env_num_or_null("UNDERSAMPLE", runtime_params$undersample)

shap_n <- 300L
shap_nsim <- 8L
```

```{r criterion-config}
criterion_types <- c("CI", "CIg", "CIc", "L")
rineq_ci_types <- c("CI", "CIg", "CIc", "CIw")
results_object_file <- here("data", "drc_report_results_objects.rda")
source_data_file <- here("prev_analysis", "DRCongo .dta")
paper_reference_file <- here("prev_analysis", "667.full.pdf")
outcome_labels <- c(
  alive = "Alive at age 5",
  died = "Died before age 5"
)

main_tree_control <-ci_tree_control(
  minsplit = 500L,
  minbucket = 250L,
  minprob = 0.02,
  maxdepth = 5L,
  min_gain = 0.00001,
  min_relative_gain = 0.20
)
```

```{r map-model-terms-helper}
map_model_terms_to_predictors <- function(term_names, predictors) {
  vapply(term_names, function(term) {
    hit <- predictors[term == predictors | startsWith(term, predictors)]
    c(hit[which.max(nchar(hit))], term)[1L]
  }, character(1L))
}
```

```{r canonicalize-model-terms-helper}
canonicalize_model_terms <- function(term_names, reference_terms) {
  lookup <- data.table(
    variable = reference_terms,
    syntactic = make.names(reference_terms, unique = FALSE)
  )
  lookup <- lookup[
    !duplicated(syntactic) & !duplicated(syntactic, fromLast = TRUE)
  ]

  out <- term_names
  exact <- match(out, reference_terms)
  out[!is.na(exact)] <- reference_terms[exact[!is.na(exact)]]

  needs_lookup <- which(is.na(exact))
  syntactic <- match(out[needs_lookup], lookup$syntactic)
  out[needs_lookup[!is.na(syntactic)]] <- lookup$variable[syntactic[!is.na(syntactic)]]

  out
}
```

```{r rank-from-named-vector-helper}
rank_from_named_vector <- function(x, method_name) {
  x <- x %||% numeric()
  data.table(variable = names(x), score = as.numeric(x))[
    is.finite(score) & score != 0
  ][
    order(-abs(score))
  ][
    ,
    .(variable, method = method_name, rank = seq_len(.N), score)
  ]
}
```

# Purpose

This document is the canonical generator for the objects used by the main DRC
results report. It is written as a readable workflow so a lecturer can follow
what is loaded, what is fitted, what is compared, and what is finally saved.

If you need a plain R script for batch execution, extract it from this document
with `knitr::purl("Generate_DRC_Results_Objects.qmd", output = "R/generate_drc_results_objects.R")`.
Display-only chunks are excluded from the purl output so the extracted script
stays focused on computation and saving.

# Run Configuration

The runtime of this generator is controlled by two quantities: the number of
analysis rows and the number of tuning folds. The remaining modelling choices
are fixed below so the workflow stays readable. Set `analysis_rows` to `1` to
use the full data set.

```{r load-2007-drc-data}
required_source_packages <- c("haven")
missing_source_packages <- required_source_packages[
  !vapply(required_source_packages, requireNamespace, logical(1L), quietly = TRUE)
]
stopifnot(length(missing_source_packages) == 0L)

drc_2007_dt <- as.data.table(haven::read_dta(source_data_file))
```

```{r prepare-2007-drc-analysis-data}
factor_from_codes <- function(x, labels) {
  factor(labels[as.character(as.integer(x))], levels = unname(labels))
}

binary_factor <- function(x, labels) {
  factor(labels[as.character(as.integer(x))], levels = unname(labels))
}

drc_birth_labels <- c(
  "1" = "First birth",
  "2" = "2-4 short interval",
  "3" = "2-4 long interval",
  "4" = "5+ short interval",
  "5" = "5+ long interval"
)
drc_wealth_group_labels <- c("1" = "High household wealth", "2" = "Low household wealth")
drc_age_mother_labels <- c("1" = "20 or more", "2" = "Less than 20")
drc_education_labels <- c("1" = "Any education", "2" = "No education")
drc_occupation_labels <- c(
  "1" = "Other",
  "2" = "Household, unskilled, not working",
  "3" = "Agriculture"
)
drc_region_labels <- c(
  "1" = "Bandundu",
  "2" = "Bas-Congo",
  "3" = "Equateur",
  "4" = "Kasai Occidental",
  "5" = "Kasai Oriental",
  "6" = "Katanga",
  "7" = "Kinshasa",
  "8" = "Maniema",
  "9" = "Nord-Kivu",
  "10" = "Orientale",
  "11" = "Sud-Kivu"
)

raw_congo_model_dt <- drc_2007_dt[
  ,
  .(
    wealth = as.numeric(wealth),
    deadu5_num = as.integer(deadu5),
    sample_weight = as.numeric(weight),
    quint = factor_from_codes(quint, drc_wealth_group_labels),
    unskilled = binary_factor(
      unskilled,
      c("0" = "Skilled birth attendance", "1" = "No skilled birth attendance")
    ),
    male = binary_factor(male, c("0" = "Female", "1" = "Male")),
    birth = factor_from_codes(birth, drc_birth_labels),
    agemoth = factor_from_codes(agemoth, drc_age_mother_labels),
    rural = binary_factor(rural, c("0" = "Urban", "1" = "Rural")),
    ed = factor_from_codes(ed, drc_education_labels),
    ped = factor_from_codes(ped, drc_education_labels),
    mocc = factor_from_codes(mocc, drc_occupation_labels),
    pocc = factor_from_codes(pocc, drc_occupation_labels),
    reg = factor_from_codes(reg, drc_region_labels)
  )
]
raw_congo_model_dt <- na.omit(raw_congo_model_dt)

raw_congo_var_labels <- c(
  quint = "Household wealth",
  unskilled = "Skilled birth attendance",
  male = "Sex of the child",
  birth = "Birth order and interval",
  agemoth = "Mother's age at birth",
  rural = "Type of residence",
  ed = "Mother's education",
  ped = "Father's education",
  mocc = "Mother's occupation",
  pocc = "Father's occupation",
  reg = "Region"
)
raw_congo_predictors <- names(raw_congo_var_labels)
```

```{r sample-2007-drc-analysis-data}
raw_congo_model_dt <- sample_analysis_rows(
    raw_congo_model_dt,
    use_full = runtime_params$analysis_rows == 1L,
    n = runtime_params$analysis_rows,
    seed = 20260516
)
```

```{r encode-analysis-predictors}
predictor_level_counts <- vapply(
    raw_congo_model_dt[, raw_congo_predictors, with = FALSE],
    function(x) length(levels(as.factor(x))),
    integer(1L)
)
nominal_predictors <- raw_congo_predictors[
    vapply(
        raw_congo_model_dt[, raw_congo_predictors, with = FALSE],
        function(x) is.factor(x) || is.character(x),
        logical(1L)
    )
]
multi_level_predictors <- intersect(
    names(predictor_level_counts[predictor_level_counts > 2L]),
    nominal_predictors
)
kept_predictors <- setdiff(raw_congo_predictors, multi_level_predictors)

make_indicator_name <- function(variable, level) {
    make.names(paste(variable, level, sep = "_"), unique = FALSE)
}

indicator_level_map <- rbindlist(lapply(multi_level_predictors, function(variable) {
    levels_i <- levels(as.factor(raw_congo_model_dt[[variable]]))
    data.table(
        raw_variable = variable,
        level = levels_i,
        variable = make_indicator_name(variable, levels_i),
        variable_label = paste0(unname(raw_congo_var_labels[variable]), ": ", levels_i)
    )
}), fill = TRUE)

stopifnot(!anyDuplicated(indicator_level_map$variable))

model_predictor_dt <- copy(raw_congo_model_dt[, kept_predictors, with = FALSE])
for (i in seq_len(nrow(indicator_level_map))) {
    source_variable <- indicator_level_map$raw_variable[i]
    source_level <- indicator_level_map$level[i]
    target_variable <- indicator_level_map$variable[i]
    model_predictor_dt[
        ,
        (target_variable) := factor(
            fifelse(as.character(raw_congo_model_dt[[source_variable]]) == source_level, "Yes", "No"),
            levels = c("No", "Yes")
        )
    ]
}

congo_model_dt <- data.table(
    raw_congo_model_dt[, .(wealth, deadu5_num, sample_weight)],
    model_predictor_dt
)
congo_predictors <- names(model_predictor_dt)
congo_var_labels <- setNames(congo_predictors, congo_predictors)
congo_var_labels[intersect(kept_predictors, names(raw_congo_var_labels))] <-
    raw_congo_var_labels[intersect(kept_predictors, names(raw_congo_var_labels))]
congo_var_labels[indicator_level_map$variable] <- indicator_level_map$variable_label

congo_ci_formula <- stats::as.formula(
    paste("cbind(wealth, deadu5_num) ~", paste(congo_predictors, collapse = " + "))
)
```

```{r format-variable-label-helper}
canonicalize_regression_terms <- function(term_names) {
    vapply(term_names, function(term) {
        parent <- map_model_terms_to_predictors(term, raw_congo_predictors)
        source_level <- sub(paste0("^", parent), "", term)
        hit <- indicator_level_map[raw_variable == parent & level == source_level]
        choices <- c(hit$variable[1L], parent, term)
        choice_id <- which(c(
            parent %in% multi_level_predictors && parent != term && nrow(hit) > 0L,
            parent %in% raw_congo_predictors && parent != term,
            TRUE
        ))[1L]
        choices[choice_id]
    }, character(1L))
}

format_variable_label <- function(x) {
    vapply(x, function(value) {
        parent <- map_model_terms_to_predictors(value, raw_congo_predictors)
        suffix <- trimws(sub(paste0("^", parent), "", value))
        suffix <- gsub("^[:._]+", "", suffix)
        parent_label <- unname(raw_congo_var_labels[parent])
        choices <- c(
            NA_character_,
            unname(congo_var_labels[value]),
            unname(raw_congo_var_labels[value]),
            paste0(parent_label, ": ", suffix),
            parent_label,
            value
        )
        choice_id <- which(c(
            is.na(value) || !nzchar(value),
            value %in% names(congo_var_labels),
            value %in% names(raw_congo_var_labels),
            parent %in% names(raw_congo_var_labels) && parent != value && nzchar(suffix),
            parent %in% names(raw_congo_var_labels) && parent != value,
            TRUE
        ))[1L]
        choices[choice_id]
    }, character(1L))
}
```

```{r run-configuration-preview, purl=FALSE}
report_compact_table(
    data.table(
        item = c(
            "Source data",
            "Rows requested",
            "Rows used",
            "Tuning folds",
            "Published benchmark",
            "Candidate predictors",
            "Output file"
        ),
        value = c(
            source_data_file,
            fifelse(
                runtime_params$analysis_rows == 1L,
                "Full data set",
                format(runtime_params$analysis_rows, big.mark = ",", scientific = FALSE)
            ),
            format(nrow(congo_model_dt), big.mark = ",", scientific = FALSE),
            runtime_params$tuning_folds,
            "Van Malderen et al. 2013 Table 3, DRCongo 2007",
            length(congo_predictors),
            results_object_file
        )
    ),
    caption = "Run configuration for the DRC results generator",
    column_widths = c("4.5cm", "9cm")
)
```

# Build Descriptive Objects

This section constructs the basic descriptive outputs that anchor the rest of
the report: outcome prevalence, predictor coverage, and root-node inequality.

```{r build-descriptive-objects}
outcome_plot_dt <- congo_model_dt[
    ,
    .(weighted_n = sum(sample_weight)),
    by = .(deadu5_num)
][
    ,
    outcome_label := fifelse(deadu5_num == 1, outcome_labels[["died"]], outcome_labels[["alive"]])
][
    ,
    weighted_percent := 100 * weighted_n / sum(weighted_n)
]

candidate_predictor_table <- rbindlist(lapply(raw_congo_predictors, function(var) {
    x <- raw_congo_model_dt[[var]]
    level_dt <- raw_congo_model_dt[, .(weighted_n = sum(sample_weight)), by = var][order(-weighted_n)]
    level_dt[, weighted_percent := 100 * weighted_n / sum(weighted_n)]
    data.table(
        variable = var,
        label = unname(raw_congo_var_labels[var]),
        class = paste(class(x), collapse = "/"),
        n_levels = uniqueN(x),
        most_common_level = as.character(level_dt[[var]][1L]),
        most_common_percent = level_dt$weighted_percent[1L]
    )
}), fill = TRUE)

drc_root_ci_table <- rbindlist(lapply(criterion_types, function(type) {
    ci_fun <- ci_factory(type)
    data.table(
        criterion = type,
        estimate = ci_fun(
            cbind(congo_model_dt$wealth, congo_model_dt$deadu5_num),
            congo_model_dt$sample_weight
        )
    )
}))

drc_data_summary_table <- data.table(
    item = c(
        "Analysis rows",
        "Weighted sample size",
        "Weighted under-five mortality",
        "Ranking variable",
        "Candidate predictors"
    ),
    value = c(
        format(nrow(congo_model_dt), big.mark = ",", scientific = FALSE),
        format(round(sum(congo_model_dt$sample_weight)), big.mark = ",", scientific = FALSE),
        sprintf("%.2f%%", 100 * weighted_mean_safe(congo_model_dt$deadu5_num, congo_model_dt$sample_weight)),
        "DHS wealth index",
        paste(unname(raw_congo_var_labels), collapse = "; ")
    )
)
```

```{r descriptive-summary-preview, purl=FALSE}
report_compact_table(
  drc_data_summary_table,
  caption = "Analysis sample summary used by the DRC results report",
  column_widths = c("4cm", "10cm")
)
```

```{r descriptive-root-ci-preview, purl=FALSE}
report_compact_table(
  drc_root_ci_table,
  digits = 4,
  caption = "Root-node concentration-index estimates across criteria"
)
```

# Build Classical Decomposition Objects

This section now follows a result-first workflow. The survey model is fitted in
its own chunk, the summary is shown immediately after, and each `rineq`
decomposition method is built in its own chunk before being displayed.

```{r fit-classical-model}
library(survey)
library(rineq)
linear_decomp_formula <- stats::as.formula(
  paste("deadu5_num ~", paste(raw_congo_predictors, collapse = " + "))
)
linear_decomp_design <- survey::svydesign(
  ids = ~1,
  weights = ~sample_weight,
  data = raw_congo_model_dt
)
linear_decomp_fit <- survey::svyglm(
  linear_decomp_formula,
  design = linear_decomp_design,
  family = stats::quasibinomial()
)
```

```{r classical-model-summary, purl=FALSE}
print(summary(linear_decomp_fit))
```

```{r extract-classical-model-components}
linear_x <- stats::model.matrix(linear_decomp_formula, data = raw_congo_model_dt)
linear_beta <- stats::coef(linear_decomp_fit)
linear_terms <- setdiff(colnames(linear_x), "(Intercept)")

linear_term_info <- data.table(
  term = linear_terms,
  variable = canonicalize_regression_terms(linear_terms),
  linear_coefficient = unname(linear_beta[linear_terms]),
  mean_of_variable = vapply(
    linear_terms,
    function(term) weighted_mean_safe(linear_x[, term], raw_congo_model_dt$sample_weight),
    numeric(1L)
  )
)

regression_model_summary <- as.data.table(
  as.data.frame(broom::tidy(linear_decomp_fit, conf.int = TRUE))
)
regression_model_summary[, variable := fifelse(
  term == "(Intercept)",
  term,
  canonicalize_regression_terms(term)
)]
regression_model_summary[, term_label := fifelse(
  term == "(Intercept)",
  "Intercept",
  format_variable_label(variable)
)]
data.table::setcolorder(
  regression_model_summary,
  c("term", "term_label", "variable", setdiff(names(regression_model_summary), c("term", "term_label", "variable")))
)
```

```{r decomposition-to-table-helper}
decomposition_to_table <- function(object, digits = 4L) {
  data.table(
    term = names(object$rel_contribution),
    `Contribution (%)` = round(object$rel_contribution, digits),
    `Contribution (Abs)` = round(object$ci_contribution, digits),
    Elasticity = round(c(0, object$elasticities), digits),
    `Concentration Index` = c(NA_real_, round(object$partial_cis, digits)),
    `lower 5%` = c(NA_real_, round(object$confints[1, ], digits)),
    `upper 5%` = c(NA_real_, round(object$confints[2, ], digits)),
    Corrected = c("", ifelse(object$corrected_coefficients, "yes", "no"))
  )
}
```

```{r summarise-rineq-decomposition-helper}
summarise_rineq_decomposition <- function(object, method_name) {
  out <- merge(
    linear_term_info,
    decomposition_to_table(object, digits = 6L)[
      term != "residual",
      .(
        term,
        elasticity = Elasticity,
        concentration_index_of_variable = `Concentration Index`,
        regression_contribution = `Contribution (Abs)`
      )
    ],
    by = "term",
    all.y = TRUE
  )
  out[, variable := fifelse(!is.na(variable) & nzchar(variable), variable, term)]
  out[, `:=`(
    criterion = method_name,
    variable_label = format_variable_label(variable)
  )]

  total_abs_contribution <- out[, sum(abs(regression_contribution), na.rm = TRUE)]
  out[, regression_pct_contribution := fifelse(
    rep(is.finite(total_abs_contribution) && total_abs_contribution > .Machine$double.eps, .N),
    100 * abs(regression_contribution) / total_abs_contribution,
    NA_real_
  )]

  data.table::setcolorder(
    out,
    c(
      "variable",
      "variable_label",
      "criterion",
      "linear_coefficient",
      "mean_of_variable",
      "elasticity",
      "concentration_index_of_variable",
      "regression_contribution",
      "regression_pct_contribution"
    )
  )
  out[order(-abs(regression_pct_contribution))]
}
```

```{r build-rineq-ci-decomposition}
rineq_ci_decomp <- rineq::decomposition(
  outcome = raw_congo_model_dt$deadu5_num,
  betas = unname(linear_beta[linear_terms]),
  mm = linear_x[, linear_terms, drop = FALSE],
  ranker = raw_congo_model_dt$wealth,
  wt = raw_congo_model_dt$sample_weight,
  correction = TRUE,
  citype = "CI"
)
rineq_ci_summary <- decomposition_to_table(rineq_ci_decomp, digits = 4L)
classical_ci_table <- summarise_rineq_decomposition(rineq_ci_decomp, "CI")
```

```{r rineq-ci-output, purl=FALSE, fig.height=7}
report_compact_table(
  rineq_ci_summary,
  digits = 4,
  caption = "rineq decomposition summary: CI"
)
plot(
  rineq_ci_decomp,
  horiz = TRUE,
  decreasing = FALSE,
  main = "rineq decomposition contributions: CI"
)
```

```{r build-rineq-cig-decomposition}
rineq_cig_decomp <- rineq::decomposition(
  outcome = raw_congo_model_dt$deadu5_num,
  betas = unname(linear_beta[linear_terms]),
  mm = linear_x[, linear_terms, drop = FALSE],
  ranker = raw_congo_model_dt$wealth,
  wt = raw_congo_model_dt$sample_weight,
  correction = TRUE,
  citype = "CIg"
)
rineq_cig_summary <- decomposition_to_table(rineq_cig_decomp, digits = 4L)
classical_cig_table <- summarise_rineq_decomposition(rineq_cig_decomp, "CIg")
```

```{r rineq-cig-output, purl=FALSE, fig.height=7}
report_compact_table(
  rineq_cig_summary,
  digits = 4,
  caption = "rineq decomposition summary: CIg"
)
plot(
  rineq_cig_decomp,
  horiz = TRUE,
  decreasing = FALSE,
  main = "rineq decomposition contributions: CIg"
)
```

```{r build-rineq-cic-decomposition}
rineq_cic_decomp <- rineq::decomposition(
  outcome = raw_congo_model_dt$deadu5_num,
  betas = unname(linear_beta[linear_terms]),
  mm = linear_x[, linear_terms, drop = FALSE],
  ranker = raw_congo_model_dt$wealth,
  wt = raw_congo_model_dt$sample_weight,
  correction = TRUE,
  citype = "CIc"
)
rineq_cic_summary <- decomposition_to_table(rineq_cic_decomp, digits = 4L)
classical_cic_table <- summarise_rineq_decomposition(rineq_cic_decomp, "CIc")
```

```{r rineq-cic-output, purl=FALSE, fig.height=7}
report_compact_table(
  rineq_cic_summary,
  digits = 4,
  caption = "rineq decomposition summary: CIc"
)
plot(
  rineq_cic_decomp,
  horiz = TRUE,
  decreasing = FALSE,
  main = "rineq decomposition contributions: CIc"
)
```

```{r build-rineq-ciw-decomposition}
rineq_ciw_decomp <- rineq::decomposition(
  outcome = raw_congo_model_dt$deadu5_num,
  betas = unname(linear_beta[linear_terms]),
  mm = linear_x[, linear_terms, drop = FALSE],
  ranker = raw_congo_model_dt$wealth,
  wt = raw_congo_model_dt$sample_weight,
  correction = TRUE,
  citype = "CIw"
)
rineq_ciw_summary <- decomposition_to_table(rineq_ciw_decomp, digits = 4L)
classical_ciw_table <- summarise_rineq_decomposition(rineq_ciw_decomp, "CIw")
```

```{r rineq-ciw-output, purl=FALSE, fig.height=7}
report_compact_table(
  rineq_ciw_summary,
  digits = 4,
  caption = "rineq decomposition summary: CIw"
)
plot(
  rineq_ciw_decomp,
  horiz = TRUE,
  decreasing = FALSE,
  main = "rineq decomposition contributions: CIw"
)
```

```{r assemble-rineq-objects}
rineq_decomp_by_type <- list(
  CI = rineq_ci_decomp,
  CIg = rineq_cig_decomp,
  CIc = rineq_cic_decomp,
  CIw = rineq_ciw_decomp
)

rineq_summary_by_type <- list(
  CI = rineq_ci_summary,
  CIg = rineq_cig_summary,
  CIc = rineq_cic_summary,
  CIw = rineq_ciw_summary
)

rineq_summary <- rbindlist(lapply(names(rineq_summary_by_type), function(type) {
  out <- copy(rineq_summary_by_type[[type]])
  out[, criterion := type]
  data.table::setcolorder(out, c("criterion", setdiff(names(out), "criterion")))
  out
}), fill = TRUE)

classical_decomposition_table <- rbindlist(
  list(classical_ci_table, classical_cig_table, classical_cic_table, classical_ciw_table),
  fill = TRUE
)
```

```{r published-2007-reference}
published_drc_table3_reference <- rbindlist(list(
  data.table(
    criterion = "G",
    determinant = c(
      "Household wealth",
      "Skilled birth attendance",
      "Sex of the child",
      "Birth order and interval",
      "Mother's age at birth",
      "Type of residence",
      "Mother's education",
      "Father's education",
      "Mother's occupation",
      "Father's occupation",
      "Region"
    ),
    published_index = -0.128,
    published_percent = c(2.4, 7.3, 5.3, 35.6, 0.4, 3.4, 17.4, -0.7, 5.8, 9.1, 14.0)
  ),
  data.table(
    criterion = "C",
    determinant = c(
      "Household wealth",
      "Skilled birth attendance",
      "Sex of the child",
      "Birth order and interval",
      "Mother's age at birth",
      "Type of residence",
      "Mother's education",
      "Father's education",
      "Mother's occupation",
      "Father's occupation",
      "Region"
    ),
    published_index = 0.057,
    published_percent = c(13.1, 11.3, -0.6, -1.2, 0.4, 12.5, 33.1, -2.1, 17.6, 20.8, -4.9)
  )
))

determinant_map <- data.table(
  variable = c("quint", "unskilled", "male", "birth", "agemoth", "rural", "ed", "ped", "mocc", "pocc", "reg"),
  raw_variable = c("quint", "unskilled", "male", "birth", "agemoth", "rural", "ed", "ped", "mocc", "pocc", "reg"),
  determinant = c(
    "Household wealth",
    "Skilled birth attendance",
    "Sex of the child",
    "Birth order and interval",
    "Mother's age at birth",
    "Type of residence",
    "Mother's education",
    "Father's education",
    "Mother's occupation",
    "Father's occupation",
    "Region"
  )
)
determinant_map <- rbindlist(
  list(
    determinant_map,
    merge(
      indicator_level_map[, .(variable, raw_variable)],
      determinant_map[, .(raw_variable, determinant)],
      by = "raw_variable",
      all.x = TRUE,
      sort = FALSE
    )[, .(variable, raw_variable, determinant)]
  ),
  fill = TRUE
)

classical_decomposition_grouped <- classical_decomposition_table[
  determinant_map,
  on = "variable",
  nomatch = 0L
][
  ,
  .(
    regression_contribution = sum(regression_contribution, na.rm = TRUE),
    regression_pct_contribution = sum(regression_pct_contribution, na.rm = TRUE)
  ),
  by = .(criterion, determinant)
]

published_drc_table3_comparison <- merge(
  published_drc_table3_reference[criterion == "C"],
  classical_decomposition_grouped[criterion == "CI"],
  by = "determinant",
  all.x = TRUE,
  sort = FALSE
)
published_drc_table3_comparison[
  ,
  difference_from_published := regression_pct_contribution - published_percent
]
```

```{r published-2007-reference-preview, purl=FALSE}
report_compact_table(
  published_drc_table3_comparison[
    ,
    .(
      determinant,
      paper_c_percent = published_percent,
      generated_ci_percent = regression_pct_contribution,
      difference_from_published
    )
  ],
  digits = 2,
  caption = "Published DRCongo 2007 concentration-index decomposition compared with generated regression decomposition"
)
```

# Fit Concentration-Index Trees

This is the core generator step. The grid is cross-validated, then each
criterion is refitted at its selected setting.

```{r tree-control-grid}
tree_grid <- ci_tree_control_grid(
    minsplit = c(500L), # 750L worked
    minbucket = c(100L), # 100 worked
    minprob = 0.01,
    maxdepth = 10L,
    min_gain = 0.0001,
    min_relative_gain = c(0.15,  0.20, 0.25,  0.30)
)

# filter where minbucket is at least 50% of minsplit to avoid invalid settings
tree_grid <- tree_grid[minbucket <= 0.5 * minsplit]
```

```{r tune-ci-trees}
ci_tuning_metrics <- c(
    "validation_gain",
    "relative_validation_gain",
    "percent_validation_root_recovered",
    "brier",
    "log_loss",
    "roc_auc"
)

tree_tuning <- tune_ci_tree(
    formula = congo_ci_formula,
    data = congo_model_dt,
    rank_name = "wealth",
    outcome_name = "deadu5_num",
    weights = congo_model_dt$sample_weight,
    type = criterion_types,
    control_grid = tree_grid,
    v = runtime_params$tuning_folds,
    strata = "deadu5_num",
    seed = 20260508,
    metric = ci_tuning_metrics,
    refit = FALSE,
    control = control_ci_tune(save_pred = TRUE)
)
```

```{r select-ci-tree-settings}
tree_best_by_type <- ci_select_best(tree_tuning,
    metric = "relative_validation_gain"
)
tree_selection_table <- ci_fit_summary_table(
    tree_tuning,
    selected = tree_best_by_type,
    metrics = unique(tree_tuning$summary$metric),
    include_percent = FALSE
)
tree_selection_table[
    tree_best_by_type,
    terminal_nodes := i.mean_terminal_nodes,
    on = c("type", "grid_id")
]
tree_selection_table[, selection_basis := paste(
    runtime_params$tuning_folds,
    "fold cross-validation"
)]
report_compact_table(
    tree_selection_table,
    digits = 4,
    caption = "Summary of the concentration-index tree selection results across criteria"
)
```

```{r fit-selected-ci-trees}
tree_models_by_type <- setNames(
    lapply(seq_len(nrow(tree_best_by_type)), function(i) {
        setting <- tree_best_by_type[i]
        ci_tree(
            formula = congo_ci_formula,
            data = congo_model_dt,
            rank_name = "wealth",
            outcome_name = "deadu5_num",
            weights = congo_model_dt$sample_weight,
            type = setting$type[1L],
            control = ci_control_from_row(setting)
        )
    }),
    tree_best_by_type$type
)
```

```{r summarise-selected-ci-tree}
selected_tree_type <- tree_selection_table[order(-mean_validation_relative_gain)][1L, type]
selected_tree_fit <- tree_models_by_type[[selected_tree_type]]
selected_tree_summary <- as.data.table(ci_tree_terminal_summary(selected_tree_fit))

all_tree_terminal_summaries <- rbindlist(lapply(names(tree_models_by_type), function(type) {
    out <- as.data.table(ci_tree_terminal_summary(tree_models_by_type[[type]]))
    out[, criterion := type]
    data.table::setcolorder(out, "criterion")
    out
}), fill = TRUE)
```

```{r summarise-ci-tree-selection-objects}
tree_variable_importance <- collect_variable_importance(tree_models_by_type)

tree_best_controls <- tree_best_by_type[
    ,
    .(type, minsplit, minbucket, minprob, maxdepth, min_gain, min_relative_gain)
]

tree_complexity_metrics <- collect_complexity_metrics(tree_tuning)
tree_min_relative_gain_path <- min_relative_gain_path(tree_complexity_metrics,
    selected = tree_best_by_type
)

tree_fit_wide <- ci_fit_summary_table(
    tree_tuning,
    selected = tree_best_by_type,
    metrics = unique(tree_tuning$summary$metric),
    include_percent = FALSE
)
tree_fit_table <- ci_collect_metrics(
    tree_tuning,
    selected = tree_best_by_type,
    metric = unique(tree_tuning$summary$metric),
    format = "tidy",
    include_train = TRUE
)
```

```{r tree-selection-preview, purl=FALSE}
report_compact_table(
  tree_selection_table,
  digits = 4,
  caption = "Selection summary for the fitted concentration-index trees"
)
```

```{r selected-tree-preview, purl=FALSE, fig.width=12, fig.height=8}
ci_report_tree_plot(
  fit = selected_tree_fit,
  data = congo_model_dt,
  outcome_name = "deadu5_num",
  outcome_label = "U5 death",
  ci_type = "L",
  var_labels = congo_var_labels
)
```

```{r ci-tree-preview, purl=FALSE, fig.width=12, fig.height=8}
ci_report_tree_plot(
  fit = tree_models_by_type[["CI"]],
  data = congo_model_dt,
  outcome_name = "deadu5_num",
  outcome_label = "U5 death",
  ci_type = "CI",
  var_labels = congo_var_labels
)
```

```{r cig-tree-preview, purl=FALSE, fig.width=12, fig.height=8}
ci_report_tree_plot(
  fit = tree_models_by_type[["CIg"]],
  data = congo_model_dt,
  outcome_name = "deadu5_num",
  outcome_label = "U5 death",
  ci_type = "CIg",
  var_labels = congo_var_labels
)
```

```{r cic-tree-preview, purl=FALSE, fig.width=12, fig.height=8}
ci_report_tree_plot(
  fit = tree_models_by_type[["CIc"]],
  data = congo_model_dt,
  outcome_name = "deadu5_num",
  outcome_label = "U5 death",
  ci_type = "CIc",
  var_labels = congo_var_labels
)
```

# Fit Lecturer Rpart Comparison Trees

The benchmark rpart trees are retained because they are easy to discuss beside
the inequality-driven tree.

```{r build-rpart-comparison}
lecturer_rpart_models <- fit_ci_rpart_comparison(
  data = congo_model_dt,
  predictors = congo_predictors,
  outcome = "deadu5_num",
  rank = "wealth",
  weights = "sample_weight"
)
lecturer_rpart_summary <- summarise_ci_rpart_comparison(lecturer_rpart_models)
```

```{r rpart-comparison-preview, purl=FALSE}
rpart_summary_out <- copy(lecturer_rpart_summary)[is_terminal == TRUE]

report_compact_table(
  rpart_summary_out,
  digits = 4,
  caption = "Terminal-node summaries for the lecturer rpart comparison trees"
)
```

```{r rpart-comparison-plots, purl=FALSE, fig.width=12, fig.height=8, results='asis'}
plot_labels <- c(
    simple_abs = "Lecturer rpart method 1: absolute CI impurity",
    robust_abs = "Lecturer rpart method 2: robust absolute CI impurity"
)
plot_fontsize <- c(simple_abs = 4.2, robust_abs = 4.6)

for (method_name in names(lecturer_rpart_models)) {
    cat("\n\n### ", plot_labels[[method_name]] %||% method_name, "\n\n", sep = "")
    ci_report_tree_plot(
        fit = lecturer_rpart_models[[method_name]],
        data = congo_model_dt,
        outcome_name = "deadu5_num",
        outcome_label = "U5 death",
        ci_type = "CI",
        var_labels = congo_var_labels,
        fontsize = plot_fontsize[[method_name]]
    )
}
```

# Forest and SHAP Objects

These objects are the most expensive part of the generator, so their runtime is
kept understandable by using the same row count and fold count set above.
`rineq` does not expose a SHAP decomposition API, so the SHAP section uses a
cross-validated `ranger` model and `shap_conc_decomp` for the four
reported inequality criteria.

```{r forest-control-grid}
p <- length(congo_predictors)
forest_grid <- ci_tree_control_grid(
    minsplit = 500L,
    minbucket = 100L,
    minprob = 0.01,
    maxdepth = 15L,
    min_gain = 0.00001,
    min_relative_gain = c( 0.25, 0.30, 0.35, 0.40),
    mtry = c(10),
    ntree = 50L
)
```

```{r tune-ci-forests}
forest_tuning <- run_forest_tuning_batches(
  formula = congo_ci_formula,
  data = congo_model_dt,
  forest_grid = forest_grid,
  criterion_types = criterion_types,
  tuning_metrics = ci_tuning_metrics,
  tuning_selection_metric = "relative_validation_gain",
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  weights = congo_model_dt$sample_weight,
  folds = runtime_params$tuning_folds,
  workers = 4L,
  progress_steps = min(20L, nrow(forest_grid)),
  log_file = here("logs", "forest_tuning.log"),
  seed = 20260508,
  perturb = list(replace = FALSE, fraction = 0.7)
)
```

```{r select-ci-forest-settings}
forest_best_by_type <- ci_select_best(
  forest_tuning,
  metric = "relative_validation_gain"
)
```

```{r fit-selected-ci-forests}
set.seed(20260509)
forest_models_by_type <- setNames(
  lapply(seq_len(nrow(forest_best_by_type)), function(i) {
    setting <- forest_best_by_type[i]
    ci_forest(
      formula = congo_ci_formula,
      data = congo_model_dt,
      rank_name = "wealth",
      outcome_name = "deadu5_num",
      weights = congo_model_dt$sample_weight,
      type = setting$type[1L],
      control = ci_control_from_row(setting, include_mtry = FALSE),
      ntree = as.integer(setting$ntree[1L]),
      mtry = as.integer(setting$mtry[1L]),
      perturb = list(replace = FALSE, fraction = 0.632)
    )
  }),
  forest_best_by_type$type
)
selected_forest_fit <- forest_models_by_type[[selected_tree_type]]
forest_variable_importance <- collect_variable_importance(forest_models_by_type)
```

```{r summarise-ci-forest-complexity}
forest_complexity_metrics <- collect_complexity_metrics(forest_tuning)
forest_min_relative_gain_path <- min_relative_gain_path(
  forest_complexity_metrics,
  selected = forest_best_by_type
)
```

```{r summarise-ci-forest-selection}
forest_selection_table <- ci_fit_summary_table(
    forest_tuning,
    selected = forest_best_by_type,
    metrics = unique(forest_tuning$summary$metric),
    include_percent = FALSE
)
forest_selection_table <- merge(
    forest_selection_table,
    forest_best_by_type[, .(type, ntree, mtry)],
    by = "type",
    all.x = TRUE,
    sort = FALSE
)
forest_selection_table[
    forest_best_by_type,
    `:=`(
        mean_terminal_nodes = i.mean_terminal_nodes,
        mean_max_depth = i.maxdepth
    ),
    on = "type"
]
forest_selection_table[, selection_basis := paste(
    runtime_params$tuning_folds,
    "fold cross-validation"
)]

forest_fit_wide <- copy(forest_selection_table)
forest_fit_table <- ci_collect_metrics(
    forest_tuning,
    selected = forest_best_by_type,
    metric = unique(forest_tuning$summary$metric),
    format = "tidy",
    include_train = TRUE
)
forest_best_controls <- forest_best_by_type[
    ,
    .(
        type, ntree, mtry, minsplit, minbucket,
        minprob, maxdepth,
        min_gain, min_relative_gain
    )
]
```

```{r fit-ci-forest-surrogates}
forest_surrogates_by_type <- setNames(
  lapply(seq_len(nrow(forest_best_by_type)), function(i) {
    setting <- copy(forest_best_by_type[i])
    setting$min_relative_gain <- 0.30
    setting$minsplit <- 1500L
    setting$minbucket <- 100L
    setting$maxdepth <- 5L
    ci_forest_surrogate(
      forest_fit = forest_models_by_type[[setting$type[1L]]],
      data = congo_model_dt,
      formula = congo_ci_formula,
      rank_name = "wealth",
      weights = congo_model_dt$sample_weight,
      type = setting$type[1L],
      control = ci_control_from_row(setting, include_mtry = FALSE),
      prediction_name = "forest_risk"
    )
  }),
  forest_best_by_type$type
)
selected_forest_surrogate <- forest_surrogates_by_type[[selected_tree_type]]
selected_forest_surrogate_summary <- as.data.table(
  ci_tree_terminal_summary(selected_forest_surrogate$fit)
)
```

```{r ci-forest-surrogate-preview, purl=FALSE, fig.width=12, fig.height=8}
ci_report_tree_plot(
  fit = forest_surrogates_by_type[["CI"]]$fit,
  data = forest_surrogates_by_type[["CI"]]$data,
  outcome_name = forest_surrogates_by_type[["CI"]]$prediction_name,
  outcome_label = "Forest predicted U5 death",
  ci_type = "CI",
  var_labels = congo_var_labels
)
```

```{r cig-forest-surrogate-preview, purl=FALSE, fig.width=12, fig.height=8}
ci_report_tree_plot(
  fit = forest_surrogates_by_type[["CIg"]]$fit,
  data = forest_surrogates_by_type[["CIg"]]$data,
  outcome_name = forest_surrogates_by_type[["CIg"]]$prediction_name,
  outcome_label = "Forest predicted U5 death",
  ci_type = "CIg",
  var_labels = congo_var_labels
)
```

```{r cic-forest-surrogate-preview, purl=FALSE, fig.width=12, fig.height=8}
ci_report_tree_plot(
  fit = forest_surrogates_by_type[["CIc"]]$fit,
  data = forest_surrogates_by_type[["CIc"]]$data,
  outcome_name = forest_surrogates_by_type[["CIc"]]$prediction_name,
  outcome_label = "Forest predicted U5 death",
  ci_type = "CIc",
  var_labels = congo_var_labels
)
```


```{r build-ranger-analysis-data}
library(tidymodels)
ranger_analysis <- build_ranger_analysis_data(
  data = congo_model_dt,
  predictor_terms = congo_predictors,
  undersample = runtime_params$undersample,
  outcome_labels = outcome_labels
)
list2env(ranger_analysis, envir = environment())
```

```{r build-ranger-recipe}
ranger_recipe <- recipes::recipe(deadu5_factor ~ ., data = ranger_fit_df)
```

```{r build-ranger-model-spec}
ranger_model_spec <- parsnip::rand_forest(
  mtry = tune::tune(),
  min_n = tune::tune(),
  trees = 500L
) |>
  parsnip::set_mode("classification") |>
  parsnip::set_engine("ranger", probability = TRUE, importance = "impurity")
```

```{r build-ranger-resamples}
ranger_resamples <- rsample::vfold_cv(
  ranger_fit_df,
  v = runtime_params$tuning_folds,
  strata = deadu5_factor
)
```

```{r build-ranger-workflow}
ranger_workflow <- workflows::workflow() |>
  workflows::add_model(ranger_model_spec) |>
  workflows::add_recipe(ranger_recipe) |>
  workflows::add_case_weights(sample_weight_case)
```

```{r build-ranger-grid}
ranger_parameter_set <- hardhat::extract_parameter_set_dials(ranger_workflow) |>
  update(
    mtry = dials::mtry(range = c(1L, length(ranger_predictor_terms))),
    min_n = dials::min_n(range = c(200L, 2000L))
  )
ranger_grid <- dials::grid_regular(
  ranger_parameter_set,
  levels = c(mtry = 4, min_n = 3)
)
```

```{r tune-ranger-model}
ranger_metric_set <- yardstick::metric_set(
  yardstick::roc_auc,
  yardstick::accuracy,
  yardstick::sens
)
ranger_tune_results <- tune::tune_grid(
  ranger_workflow,
  resamples = ranger_resamples,
  grid = ranger_grid,
  metrics = ranger_metric_set,
  control = tune::control_grid(save_pred = TRUE)
)
ranger_tuning <- as.data.table(tune::collect_metrics(ranger_tune_results))
```

```{r ranger-cv-preview, purl=FALSE}
report_compact_table(
  ranger_tuning,
  digits = 4,
  caption = "Predictive ranger cross-validation summary"
)
```

```{r select-best-ranger-model}
selected_ranger_setting <- tune::select_best(ranger_tune_results, metric = "roc_auc")
```

```{r selected-ranger-preview, purl=FALSE}
report_compact_table(
  as.data.table(as.data.frame(selected_ranger_setting)),
  digits = 4,
  caption = "Selected predictive ranger setting"
)
```

```{r fit-selected-ranger-model}
final_ranger_workflow <- tune::finalize_workflow(ranger_workflow, selected_ranger_setting)
predictive_forest_fit <- workflows::fit(final_ranger_workflow, data = ranger_fit_df)
```

```{r predict-ranger-risk-helper}
predict_ranger_risk <- function(object, newdata) {
  predict(object, new_data = as.data.frame(newdata), type = "prob")$.pred_Died
}
```

```{r prepare-shap-data}
shap_rows <- sort(sample.int(nrow(ranger_fit_df), min(shap_n, nrow(ranger_fit_df))))
shap_x_background <- ranger_fit_df[, ranger_predictor_terms, drop = FALSE]
shap_x_eval <- ranger_fit_df[shap_rows, ranger_predictor_terms, drop = FALSE]
shap_prediction <- as.numeric(predict_ranger_risk(predictive_forest_fit, shap_x_eval))
```

```{r compute-fastshap-values}
selected_forest_shap <- fastshap::explain(
  object = predictive_forest_fit,
  X = shap_x_background,
  pred_wrapper = predict_ranger_risk,
  newdata = shap_x_eval,
  nsim = shap_nsim,
  adjust = shap_nsim > 1L
)
colnames(selected_forest_shap) <- canonicalize_model_terms(
  colnames(selected_forest_shap),
  congo_predictors
)

shap_feature_values <- copy(as.data.table(shap_x_eval))
shap_feature_values[, observation_id := seq_len(.N)]
selected_forest_shap_dt <- as.data.table(selected_forest_shap)
selected_forest_shap_dt[, observation_id := seq_len(.N)]

shap_summary_long <- melt(
  selected_forest_shap_dt,
  id.vars = "observation_id",
  variable.name = "feature",
  value.name = "shap_value",
  variable.factor = FALSE
)
shap_value_long <- melt(
  shap_feature_values,
  id.vars = "observation_id",
  variable.name = "feature",
  value.name = "feature_value",
  variable.factor = FALSE
)
shap_summary_long <- merge(
  shap_summary_long,
  shap_value_long,
  by = c("observation_id", "feature"),
  all.x = TRUE,
  sort = FALSE
)
shap_summary_long[
  ,
  feature_value_numeric := suppressWarnings(as.numeric(as.character(feature_value)))
]
shap_summary_long[
  is.na(feature_value_numeric),
  feature_value_numeric := as.numeric(factor(feature_value)) - 1
]
shap_summary_long[
  ,
  feature_value_scaled := {
    range_value <- range(feature_value_numeric, na.rm = TRUE)
    fifelse(
      rep(all(is.finite(range_value)) && diff(range_value) > .Machine$double.eps, .N),
      (feature_value_numeric - range_value[1L]) / diff(range_value),
      0.5
    )
  },
  by = feature
]
shap_summary_long[, feature_label := format_variable_label(feature)]
shap_summary_long[
  ,
  mean_abs_shap := mean(abs(shap_value), na.rm = TRUE),
  by = feature
]
data.table::setcolorder(
  shap_summary_long,
  c(
    "feature",
    "feature_label",
    "observation_id",
    "shap_value",
    "feature_value",
    "feature_value_numeric",
    "feature_value_scaled",
    "mean_abs_shap"
  )
)
```

```{r build-shap-ci-decomposition}
shap_ci_decomp <- shap_conc_decomp(
  shap = selected_forest_shap,
  rank = ranger_dt$wealth[shap_rows],
  type = "CI",
  prediction = shap_prediction,
  weights = ranger_dt$sample_weight[shap_rows]
)
```

```{r shap-ci-output, purl=FALSE, fig.height=7}
shap_ci_diag <- as.data.table(shap_ci_decomp$diagnostics)
shap_ci_contrib <- as.data.table(shap_ci_decomp$contributions)[order(pct_contribution)]
report_compact_table(
  shap_ci_diag,
  digits = 4,
  caption = "SHAP decomposition diagnostics: CI"
)
report_compact_table(
  shap_ci_contrib,
  digits = 4,
  caption = "SHAP decomposition contributions: CI"
)
ggplot(shap_ci_contrib, aes(x = pct_contribution, y = reorder(feature, pct_contribution))) +
  geom_col(fill = "#3268A8") +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.4) +
  labs(x = "Contribution (%)", y = NULL, title = "SHAP concentration decomposition: CI") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r build-shap-cig-decomposition}
shap_cig_decomp <- shap_conc_decomp(
  shap = selected_forest_shap,
  rank = ranger_dt$wealth[shap_rows],
  type = "CIg",
  prediction = shap_prediction,
  weights = ranger_dt$sample_weight[shap_rows]
)
```

```{r shap-cig-output, purl=FALSE, fig.height=7}
shap_cig_diag <- as.data.table(shap_cig_decomp$diagnostics)
shap_cig_contrib <- as.data.table(shap_cig_decomp$contributions)[order(pct_contribution)]
report_compact_table(
  shap_cig_diag,
  digits = 4,
  caption = "SHAP decomposition diagnostics: CIg"
)
report_compact_table(
  shap_cig_contrib,
  digits = 4,
  caption = "SHAP decomposition contributions: CIg"
)
ggplot(shap_cig_contrib, aes(x = pct_contribution, y = reorder(feature, pct_contribution))) +
  geom_col(fill = "#3268A8") +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.4) +
  labs(x = "Contribution (%)", y = NULL, title = "SHAP concentration decomposition: CIg") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r build-shap-cic-decomposition}
shap_cic_decomp <- shap_conc_decomp(
  shap = selected_forest_shap,
  rank = ranger_dt$wealth[shap_rows],
  type = "CIc",
  prediction = shap_prediction,
  weights = ranger_dt$sample_weight[shap_rows]
)
```

```{r shap-cic-output, purl=FALSE, fig.height=7}
shap_cic_diag <- as.data.table(shap_cic_decomp$diagnostics)
shap_cic_contrib <- as.data.table(shap_cic_decomp$contributions)[order(pct_contribution)]
report_compact_table(
  shap_cic_diag,
  digits = 4,
  caption = "SHAP decomposition diagnostics: CIc"
)
report_compact_table(
  shap_cic_contrib,
  digits = 4,
  caption = "SHAP decomposition contributions: CIc"
)
ggplot(shap_cic_contrib, aes(x = pct_contribution, y = reorder(feature, pct_contribution))) +
  geom_col(fill = "#3268A8") +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.4) +
  labs(x = "Contribution (%)", y = NULL, title = "SHAP concentration decomposition: CIc") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r build-shap-l-decomposition}
shap_l_decomp <- shap_conc_decomp(
  shap = selected_forest_shap,
  rank = ranger_dt$wealth[shap_rows],
  type = "L",
  prediction = shap_prediction,
  weights = ranger_dt$sample_weight[shap_rows]
)
```

```{r shap-l-output, purl=FALSE, fig.height=7}
shap_l_diag <- as.data.table(shap_l_decomp$diagnostics)
shap_l_contrib <- as.data.table(shap_l_decomp$contributions)[order(pct_contribution)]
report_compact_table(
  shap_l_diag,
  digits = 4,
  caption = "SHAP decomposition diagnostics: L"
)
report_compact_table(
  shap_l_contrib,
  digits = 4,
  caption = "SHAP decomposition contributions: L"
)
ggplot(shap_l_contrib, aes(x = pct_contribution, y = reorder(feature, pct_contribution))) +
  geom_col(fill = "#3268A8") +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.4) +
  labs(x = "Contribution (%)", y = NULL, title = "SHAP concentration decomposition: L") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r assemble-shap-objects}
selected_shap_contributions <- rbindlist(list(
  as.data.table(shap_ci_decomp$contributions)[, criterion := "CI"],
  as.data.table(shap_cig_decomp$contributions)[, criterion := "CIg"],
  as.data.table(shap_cic_decomp$contributions)[, criterion := "CIc"],
  as.data.table(shap_l_decomp$contributions)[, criterion := "L"]
), fill = TRUE)
selected_shap_contributions[, feature := canonicalize_model_terms(feature, congo_predictors)]
data.table::setcolorder(selected_shap_contributions, c("criterion", setdiff(names(selected_shap_contributions), "criterion")))

selected_shap_diagnostics <- rbindlist(list(
  as.data.table(shap_ci_decomp$diagnostics)[, criterion := "CI"],
  as.data.table(shap_cig_decomp$diagnostics)[, criterion := "CIg"],
  as.data.table(shap_cic_decomp$diagnostics)[, criterion := "CIc"],
  as.data.table(shap_l_decomp$diagnostics)[, criterion := "L"]
), fill = TRUE)
data.table::setcolorder(selected_shap_diagnostics, c("criterion", setdiff(names(selected_shap_diagnostics), "criterion")))
```

```{r forest-preview, purl=FALSE}
forest_importance_out <- copy(forest_variable_importance)
forest_importance_out[, variable_label := fifelse(
  variable %in% names(congo_var_labels),
  congo_var_labels[variable],
  variable
)]

report_compact_table(
  forest_importance_out,
  digits = 4,
  caption = "Forest variable importance preview"
)
```

# Compare Methods

Once the regression, tree, forest, and SHAP objects exist, the next step is to
rank variables and assemble the comparison tables used in the main report.

```{r build-method-comparison}
linear_rank <- classical_decomposition_table[
  criterion == "CI"
][
  order(-abs(regression_pct_contribution)),
  .(variable, method = "Linear decomposition", rank = seq_len(.N))
]
tree_rank <- rank_from_named_vector(selected_tree_fit$variable.importance, "Selected CI-tree")
forest_rank <- rank_from_named_vector(selected_forest_fit$variable.importance, "Selected CI-forest")
shap_rank <- selected_shap_contributions[
  criterion == "CI"
][
  order(-abs_contribution),
  .(variable = feature, method = "Selected forest SHAP", rank = seq_len(.N))
]
importance_long <- rbindlist(list(linear_rank, tree_rank, forest_rank, shap_rank), fill = TRUE)

selected_variables <- classical_decomposition_table[
  criterion == "CI"
][
  order(-abs(regression_pct_contribution))
]$variable[seq_len(min(15L, classical_decomposition_table[criterion == "CI", .N]))]
selected_variables <- selected_variables[!is.na(selected_variables) & nzchar(selected_variables)]

method_comparison_table <- data.table(variable = selected_variables)
method_comparison_table[, variable_label := format_variable_label(variable)]
regression_wide <- dcast(
  classical_decomposition_table[
    criterion %in% c("CI", "CIg", "CIc"),
    .(variable, criterion, regression_pct_contribution)
  ],
  variable ~ criterion,
  value.var = "regression_pct_contribution"
)
method_comparison_table <- merge(
  method_comparison_table,
  regression_wide[
    ,
    .(
      variable,
      regression_ci_pct_contribution = CI,
      regression_cig_pct_contribution = CIg,
      regression_cic_pct_contribution = CIc
    )
  ],
  by = "variable",
  all.x = TRUE
)
shap_wide <- dcast(
  selected_shap_contributions[
    criterion %in% criterion_types,
    .(variable = feature, criterion, pct_contribution)
  ],
  variable ~ criterion,
  value.var = "pct_contribution"
)
shap_wide <- shap_wide[
  ,
  .(
    variable,
    forest_shap_ci_pct_contribution = CI,
    forest_shap_cig_pct_contribution = CIg,
    forest_shap_cic_pct_contribution = CIc,
    forest_shap_l_pct_contribution = L
  )
]
method_comparison_table <- merge(
  method_comparison_table,
  shap_wide,
  by = "variable",
  all.x = TRUE
)
method_comparison_table <- method_comparison_table[
  order(-abs(regression_ci_pct_contribution), -abs(forest_shap_ci_pct_contribution))
]

importance_wide <- dcast(
  importance_long,
  rank ~ method,
  value.var = "variable",
  fun.aggregate = function(x) paste(unique(x), collapse = ", ")
)
for (col in setdiff(names(importance_wide), "rank")) {
  importance_wide[, (col) := format_variable_label(get(col))]
}
importance_wide <- importance_wide[rank <= 10L]
```

```{r method-comparison-preview, purl=FALSE}
report_compact_table(
  method_comparison_table,
  digits = 3,
  caption = "Selected percentage-contribution comparison table"
)
```

```{r importance-preview, purl=FALSE}
report_compact_table(
  importance_wide,
  caption = "Ranked variables across decomposition and tree-based methods"
)
```


# Save Results Objects

The last chunk assembles the full object list and writes it to disk for the
main report to consume.

```{r save-results-objects}
drc_results_objects <- list(
  generation_params = runtime_params,
  source_data_file = source_data_file,
  paper_reference_file = paper_reference_file,
  outcome_labels = outcome_labels,
  congo_model_dt = congo_model_dt,
  raw_congo_model_dt = raw_congo_model_dt,
  congo_var_labels = congo_var_labels,
  raw_congo_var_labels = raw_congo_var_labels,
  congo_predictors = congo_predictors,
  raw_congo_predictors = raw_congo_predictors,
  indicator_level_map = indicator_level_map,
  congo_ci_formula = congo_ci_formula,
  data_summary = drc_data_summary_table,
  outcome_distribution = outcome_plot_dt,
  candidate_predictors = candidate_predictor_table,
  root_ci = drc_root_ci_table,
  regression_model_summary = regression_model_summary,
  classical_decomposition = classical_decomposition_table,
  classical_decomposition_grouped = classical_decomposition_grouped,
  published_drc_table3_reference = published_drc_table3_reference,
  published_drc_table3_comparison = published_drc_table3_comparison,
  rineq_summary = rineq_summary,
  tree_tuning = tree_tuning,
  tree_best_by_type = tree_best_by_type,
  tree_best_controls = tree_best_controls,
  tree_selection = tree_selection_table,
  tree_fit_table = tree_fit_table,
  tree_complexity_metrics = tree_complexity_metrics,
  tree_min_relative_gain_path = tree_min_relative_gain_path,
  tree_models_by_type = tree_models_by_type,
  selected_tree_type = selected_tree_type,
  selected_tree_fit = selected_tree_fit,
  selected_tree_summary = selected_tree_summary,
  all_tree_terminal_summaries = all_tree_terminal_summaries,
  tree_variable_importance = tree_variable_importance,
  lecturer_rpart_models = lecturer_rpart_models,
  lecturer_rpart_summary = lecturer_rpart_summary,
  forest_tuning = forest_tuning,
  forest_best_by_type = forest_best_by_type,
  forest_best_controls = forest_best_controls,
  forest_fit_table = forest_fit_table,
  forest_complexity_metrics = forest_complexity_metrics,
  forest_min_relative_gain_path = forest_min_relative_gain_path,
  selected_forest_fit = selected_forest_fit,
  forest_surrogates_by_type = forest_surrogates_by_type,
  selected_forest_surrogate = selected_forest_surrogate,
  selected_forest_surrogate_summary = selected_forest_surrogate_summary,
  forest_selection = forest_selection_table,
  forest_variable_importance = forest_variable_importance,
  predictive_forest_tuning = ranger_tuning,
  predictive_forest_sampling = ranger_sampling_summary,
  predictive_forest_feature_lookup = ranger_feature_lookup,
  predictive_forest_fit = predictive_forest_fit,
  shap_contributions = selected_shap_contributions,
  shap_diagnostics = selected_shap_diagnostics,
  shap_summary_long = shap_summary_long,
  method_comparison = method_comparison_table,
  importance_wide = importance_wide
)

save(drc_results_objects, file = results_object_file)

data.table(
  saved_to = results_object_file,
  object_count = length(drc_results_objects),
  object_names = paste(names(drc_results_objects), collapse = ", ")
)
```
