---
title: "DRC Results Generator"
author: "Moses Mburu"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
    df-print: paged
    dpi: 500
  pdf:
    documentclass: article
    papersize: a4
    fontsize: 10pt
    geometry:
      - margin=25mm
    pdf-engine: xelatex
    toc: true
    number-sections: true
    fig-cap-location: bottom
    tbl-cap-location: top
execute:
  echo: true
  warning: false
  message: false
  fig-dpi: 500
params:
  analysis_rows: 1
  tuning_folds: 10
  undersample: 0.4
  shap_rows: 200
  shap_n_mc_samples: 2
  shap_seed: 20260328
  shap_workers: 4
always_allow_html: true
---

```{r setup}
# Set one plotting style for all generated figures in this workflow.
knitr::opts_chunk$set(
  fig.width = 9,
  fig.height = 5.5,
  dpi = 500,
  dev = "png"
)

# Load the packages and local helper functions used across the generator.
library(data.table)
library(ggplot2)
library(grid)
library(knitr)
library(ineqTrees)
library(partykit)
library(here)
source(here("R", "report_helpers.R"))
source(here("R", "ci_rpart_comparison_methods.R"))

kableExtra::use_latex_packages()
```

```{r runtime-parameter-config}
# Store default values for slow parts of the workflow.
default_params <- list(
  analysis_rows = 1L,
  tuning_folds = 10L,
  undersample = 0.4,
  shap_rows = 0,
  shap_n_mc_samples = 8L,
  shap_seed = 20260328L,
  shap_workers = 4L
)

# Let Quarto parameters override defaults, while keeping the same object names.
runtime_params <- utils::modifyList(default_params, as.list(params))

# Coerce parameters here so later chunks do not have to repeat checks.
runtime_params$analysis_rows <- as.integer(runtime_params$analysis_rows)
runtime_params$tuning_folds <- max(2L, as.integer(runtime_params$tuning_folds))
runtime_params$undersample <- as.numeric(runtime_params$undersample)
runtime_params$shap_rows <- as.integer(runtime_params$shap_rows)
runtime_params$shap_n_mc_samples <- as.integer(runtime_params$shap_n_mc_samples)
runtime_params$shap_seed <- as.integer(runtime_params$shap_seed)
runtime_params$shap_workers <- max(1L, as.integer(runtime_params$shap_workers))

# Fall back to safe defaults if a parameter is missing or not valid.
if (is.na(runtime_params$analysis_rows) || runtime_params$analysis_rows <= 0L) {
  runtime_params$analysis_rows <- default_params$analysis_rows
}
if (is.na(runtime_params$undersample)) {
  runtime_params$undersample <- default_params$undersample
}
if (is.na(runtime_params$shap_rows) || runtime_params$shap_rows < 0L) {
  runtime_params$shap_rows <- default_params$shap_rows
}
if (is.na(runtime_params$shap_n_mc_samples) || runtime_params$shap_n_mc_samples <= 0L) {
  runtime_params$shap_n_mc_samples <- default_params$shap_n_mc_samples
}
if (is.na(runtime_params$shap_seed)) {
  runtime_params$shap_seed <- default_params$shap_seed
}

# Pull out the SHAP parameters because they are used repeatedly later.
shap_n_mc_samples <- runtime_params$shap_n_mc_samples
shap_seed <- runtime_params$shap_seed
shap_workers <- runtime_params$shap_workers
shap_rows_requested <- runtime_params$shap_rows
```

```{r criterion-config}
# These are the concentration-index criteria compared throughout the analysis.
criterion_types <- c("CI", "CIg", "CIc", "L")
rineq_ci_types <- c("CI", "CIg", "CIc", "CIw")

# Define the main input and output files in one place.
results_object_file <- here("data", "drc_report_results_objects.rda")
tree_tuning_file <- here("data", "tree_tuning.rda")
source_data_file <- here("prev_analysis", "DRCongo .dta")
paper_reference_file <- here("prev_analysis", "667.full.pdf")

# Use readable outcome labels in tables and plots.
outcome_labels <- c(
  alive = "Alive at age 5",
  died = "Died before age 5"
)

# Keep a single baseline tree control for reference and plotting.
main_tree_control <-ci_tree_control(
  minsplit = 500L,
  minbucket = 250L,
  minprob = 0.02,
  maxdepth = 5L,
  min_gain = 0.00001,
  min_relative_gain = 0.20
)
```

```{r map-model-terms-helper}
# Map model-matrix columns back to their original predictor names.
map_model_terms_to_predictors <- function(term_names, predictors) {
  vapply(term_names, function(term) {
    hit <- predictors[term == predictors | startsWith(term, predictors)]
    c(hit[which.max(nchar(hit))], term)[1L]
  }, character(1L))
}
```

```{r canonicalize-model-terms-helper}
# Convert syntactic model names back to the names used in the report tables.
canonicalize_model_terms <- function(term_names, reference_terms) {
  lookup <- data.table(
    variable = reference_terms,
    syntactic = make.names(reference_terms, unique = FALSE)
  )
  lookup <- lookup[
    !duplicated(syntactic) & !duplicated(syntactic, fromLast = TRUE)
  ]

  out <- term_names
  exact <- match(out, reference_terms)
  out[!is.na(exact)] <- reference_terms[exact[!is.na(exact)]]

  needs_lookup <- which(is.na(exact))
  syntactic <- match(out[needs_lookup], lookup$syntactic)
  out[needs_lookup[!is.na(syntactic)]] <- lookup$variable[syntactic[!is.na(syntactic)]]

  out
}
```

```{r rank-from-named-vector-helper}
# Convert named importance vectors into ranked tables for method comparison.
rank_from_named_vector <- function(x, method_name) {
  x <- x %||% numeric()
  data.table(variable = names(x), score = as.numeric(x))[
    is.finite(score) & score != 0
  ][
    order(-abs(score))
  ][
    ,
    .(variable, method = method_name, rank = seq_len(.N), score)
  ]
}
```

# Purpose

This document is the canonical generator for the objects used by the main DRC
results report. It is written as a readable workflow so a lecturer can follow
what is loaded, what is fitted, what is compared, and what is finally saved.

If you need a plain R script for batch execution, extract it from this document
with `knitr::purl("Generate_DRC_Results_Objects.qmd", output = "R/generate_drc_results_objects.R")`.
Display-only chunks are excluded from the purl output so the extracted script
stays focused on computation and saving.

# Run Configuration

The runtime of this generator is controlled by two quantities: the number of
analysis rows and the number of tuning folds. The remaining modelling choices
are fixed below so the workflow stays readable. Set `analysis_rows` to `1` to
use the full data set.

```{r load-2007-drc-data}
library(haven)

# Load the DRC 2007 data file used for the empirical application.
drc_2007_dt <- as.data.table(haven::read_dta(source_data_file))
```

```{r prepare-2007-drc-analysis-data}
# Turn labelled numeric codes into factors with readable labels.
factor_from_codes <- function(x, labels) {
  factor(labels[as.character(as.integer(x))], levels = unname(labels))
}

# Use this helper for binary variables so levels stay in the intended order.
binary_factor <- function(x, labels) {
  factor(labels[as.character(as.integer(x))], levels = unname(labels))
}

# Labels below follow the determinant groups used in the thesis report.
drc_birth_labels <- c(
  "1" = "First birth",
  "2" = "2-4 short interval",
  "3" = "2-4 long interval",
  "4" = "5+ short interval",
  "5" = "5+ long interval"
)
drc_wealth_group_labels <- c("1" = "High household wealth", "2" = "Low household wealth")
drc_age_mother_labels <- c("1" = "20 or more", "2" = "Less than 20")
drc_education_labels <- c("1" = "Any education", "2" = "No education")
drc_occupation_labels <- c(
  "1" = "Other",
  "2" = "Household, unskilled, not working",
  "3" = "Agriculture"
)
drc_region_labels <- c(
  "1" = "Bandundu",
  "2" = "Bas-Congo",
  "3" = "Equateur",
  "4" = "Kasai Occidental",
  "5" = "Kasai Oriental",
  "6" = "Katanga",
  "7" = "Kinshasa",
  "8" = "Maniema",
  "9" = "Nord-Kivu",
  "10" = "Orientale",
  "11" = "Sud-Kivu"
)

# Keep the analysis variables, convert types, and attach readable categories.
raw_congo_model_dt <- drc_2007_dt[
  ,
  .(
    wealth = as.numeric(wealth),
    deadu5_num = as.integer(deadu5),
    sample_weight = as.numeric(weight),
    PSU = as.integer(PSU),
    household = as.integer(household),
    country = as.integer(country),
    quint = factor_from_codes(quint, drc_wealth_group_labels),
    unskilled = binary_factor(
      unskilled,
      c("0" = "Skilled birth attendance", "1" = "No skilled birth attendance")
    ),
    male = binary_factor(male, c("0" = "Female", "1" = "Male")),
    birth = factor_from_codes(birth, drc_birth_labels),
    agemoth = factor_from_codes(agemoth, drc_age_mother_labels),
    rural = binary_factor(rural, c("0" = "Urban", "1" = "Rural")),
    ed = factor_from_codes(ed, drc_education_labels),
    ped = factor_from_codes(ped, drc_education_labels),
    mocc = factor_from_codes(mocc, drc_occupation_labels),
    pocc = factor_from_codes(pocc, drc_occupation_labels),
    reg = factor_from_codes(reg, drc_region_labels)
  )
]

# Drop rows with missing values so all methods use the same complete cases.
raw_congo_model_dt <- na.omit(raw_congo_model_dt)

# Store report labels separately from the data so they can be reused in plots.
raw_congo_var_labels <- c(
  quint = "Household wealth",
  unskilled = "Skilled birth attendance",
  male = "Sex of the child",
  birth = "Birth order and interval",
  agemoth = "Mother's age at birth",
  rural = "Type of residence",
  ed = "Mother's education",
  ped = "Father's education",
  mocc = "Mother's occupation",
  pocc = "Father's occupation",
  reg = "Region"
)
raw_congo_predictors <- names(raw_congo_var_labels)
```

```{r sample-2007-drc-analysis-data}
# Use the full data for the final run; allow small samples for quick checks.
raw_congo_model_dt <- sample_analysis_rows(
    raw_congo_model_dt,
    use_full = runtime_params$analysis_rows == 1L,
    n = runtime_params$analysis_rows,
    seed = 20260516
)
```

```{r encode-analysis-predictors}
# Count predictor levels before deciding how each variable enters the tree model.
predictor_level_counts <- vapply(
    raw_congo_model_dt[, raw_congo_predictors, with = FALSE],
    function(x) length(levels(as.factor(x))),
    integer(1L)
)
nominal_predictors <- raw_congo_predictors[
    vapply(
        raw_congo_model_dt[, raw_congo_predictors, with = FALSE],
        function(x) is.factor(x) || is.character(x),
        logical(1L)
    )
]
multi_level_predictors <- intersect(
    names(predictor_level_counts[predictor_level_counts > 2L]),
    nominal_predictors
)

# Multi-level nominal variables are expanded into binary indicators.
kept_predictors <- setdiff(raw_congo_predictors, multi_level_predictors)

make_indicator_name <- function(variable, level) {
    make.names(paste(variable, level, sep = "_"), unique = FALSE)
}

# Keep a lookup from each indicator back to its original variable and level.
indicator_level_map <- rbindlist(lapply(multi_level_predictors, function(variable) {
    levels_i <- levels(as.factor(raw_congo_model_dt[[variable]]))
    data.table(
        raw_variable = variable,
        level = levels_i,
        variable = make_indicator_name(variable, levels_i),
        variable_label = paste0(unname(raw_congo_var_labels[variable]), ": ", levels_i)
    )
}), fill = TRUE)

stopifnot(!anyDuplicated(indicator_level_map$variable))

# Start with variables that can enter the model as they are.
model_predictor_dt <- copy(raw_congo_model_dt[, kept_predictors, with = FALSE])

# Add one Yes/No indicator for each level of the multi-level predictors.
for (i in seq_len(nrow(indicator_level_map))) {
    source_variable <- indicator_level_map$raw_variable[i]
    source_level <- indicator_level_map$level[i]
    target_variable <- indicator_level_map$variable[i]
    model_predictor_dt[
        ,
        (target_variable) := factor(
            fifelse(as.character(raw_congo_model_dt[[source_variable]]) == source_level, "Yes", "No"),
            levels = c("No", "Yes")
        )
    ]
}

# Combine the rank, outcome, weights, and encoded predictors for tree fitting.
congo_model_dt <- data.table(
    raw_congo_model_dt[, .(wealth, deadu5_num, sample_weight)],
    model_predictor_dt
)
congo_predictors <- names(model_predictor_dt)
congo_var_labels <- setNames(congo_predictors, congo_predictors)
congo_var_labels[intersect(kept_predictors, names(raw_congo_var_labels))] <-
    raw_congo_var_labels[intersect(kept_predictors, names(raw_congo_var_labels))]
congo_var_labels[indicator_level_map$variable] <- indicator_level_map$variable_label

# The ineqTrees functions expect cbind(rank, outcome) on the left-hand side.
congo_ci_formula <- as.formula(
    paste("cbind(wealth, deadu5_num) ~", paste(congo_predictors, collapse = " + "))
)
```

```{r format-variable-label-helper}
# Convert regression term names to the same variables used by tree and SHAP tables.
canonicalize_regression_terms <- function(term_names) {
    vapply(term_names, function(term) {
        parent <- map_model_terms_to_predictors(term, raw_congo_predictors)
        source_level <- sub(paste0("^", parent), "", term)
        hit <- indicator_level_map[raw_variable == parent & level == source_level]
        choices <- c(hit$variable[1L], parent, term)
        choice_id <- which(c(
            parent %in% multi_level_predictors && parent != term && nrow(hit) > 0L,
            parent %in% raw_congo_predictors && parent != term,
            TRUE
        ))[1L]
        choices[choice_id]
    }, character(1L))
}

# Format variable names for report tables after dummy-variable expansion.
format_variable_label <- function(x) {
    vapply(x, function(value) {
        parent <- map_model_terms_to_predictors(value, raw_congo_predictors)
        suffix <- trimws(sub(paste0("^", parent), "", value))
        suffix <- gsub("^[:._]+", "", suffix)
        parent_label <- unname(raw_congo_var_labels[parent])
        choices <- c(
            NA_character_,
            unname(congo_var_labels[value]),
            unname(raw_congo_var_labels[value]),
            paste0(parent_label, ": ", suffix),
            parent_label,
            value
        )
        choice_id <- which(c(
            is.na(value) || !nzchar(value),
            value %in% names(congo_var_labels),
            value %in% names(raw_congo_var_labels),
            parent %in% names(raw_congo_var_labels) && parent != value && nzchar(suffix),
            parent %in% names(raw_congo_var_labels) && parent != value,
            TRUE
        ))[1L]
        choices[choice_id]
    }, character(1L))
}
```

```{r run-configuration-preview, purl=FALSE}
# Show the settings for this run before the expensive model fitting starts.
report_compact_table(
    data.table(
        item = c(
            "Source data",
            "Rows requested",
            "Rows used",
            "Tuning folds",
            "Published benchmark",
            "Candidate predictors",
            "Output file"
        ),
        value = c(
            source_data_file,
            fifelse(
                runtime_params$analysis_rows == 1L,
                "Full data set",
                format(runtime_params$analysis_rows, big.mark = ",", scientific = FALSE)
            ),
            format(nrow(congo_model_dt), big.mark = ",", scientific = FALSE),
            runtime_params$tuning_folds,
            "Van Malderen et al. 2013 Table 3, DRCongo 2007",
            length(congo_predictors),
            results_object_file
        )
    ),
    caption = "Run configuration for the DRC results generator",
    column_widths = c("4.5cm", "9cm")
)
```

# Build Descriptive Objects

This section constructs the basic descriptive outputs that anchor the rest of
the report: outcome prevalence, predictor coverage, and root-node inequality.

```{r build-descriptive-objects}
# Summarise the weighted outcome distribution used in the results section.
outcome_plot_dt <- congo_model_dt[
    ,
    .(weighted_n = sum(sample_weight)),
    by = .(deadu5_num)
][
    ,
    outcome_label := fifelse(deadu5_num == 1, outcome_labels[["died"]], outcome_labels[["alive"]])
][
    ,
    weighted_percent := 100 * weighted_n / sum(weighted_n)
]

# Build a compact table describing the raw candidate predictors.
candidate_predictor_table <- rbindlist(lapply(raw_congo_predictors, function(var) {
    x <- raw_congo_model_dt[[var]]
    level_dt <- raw_congo_model_dt[, .(weighted_n = sum(sample_weight)), by = var][order(-weighted_n)]
    level_dt[, weighted_percent := 100 * weighted_n / sum(weighted_n)]
    data.table(
        variable = var,
        label = unname(raw_congo_var_labels[var]),
        class = paste(class(x), collapse = "/"),
        n_levels = uniqueN(x),
        most_common_level = as.character(level_dt[[var]][1L]),
        most_common_percent = level_dt$weighted_percent[1L]
    )
}), fill = TRUE)

# Compute the root-node inequality before any tree split is made.
drc_root_ci_table <- rbindlist(lapply(criterion_types, function(type) {
    ci_fun <- ci_factory(type)
    data.table(
        criterion = type,
        estimate = ci_fun(
            cbind(congo_model_dt$wealth, congo_model_dt$deadu5_num),
            congo_model_dt$sample_weight
        )
    )
}))

# Collect the main sample details read by the final thesis report.
drc_data_summary_table <- data.table(
    item = c(
        "Analysis rows",
        "Weighted sample size",
        "Weighted under-five mortality",
        "Ranking variable",
        "Candidate predictors"
    ),
    value = c(
        format(nrow(congo_model_dt), big.mark = ",", scientific = FALSE),
        format(round(sum(congo_model_dt$sample_weight)), big.mark = ",", scientific = FALSE),
        sprintf("%.2f%%", 100 * weighted_mean_safe(congo_model_dt$deadu5_num, congo_model_dt$sample_weight)),
        "DHS wealth index",
        paste(unname(raw_congo_var_labels), collapse = "; ")
    )
)
```

```{r descriptive-summary-preview, purl=FALSE}
report_compact_table(
  drc_data_summary_table,
  caption = "Analysis sample summary used by the DRC results report",
  column_widths = c("4cm", "10cm")
)
```

```{r descriptive-root-ci-preview, purl=FALSE}
report_compact_table(
  drc_root_ci_table,
  digits = 4,
  caption = "Root-node concentration-index estimates across criteria"
)
```

# Build Classical Decomposition Objects

This section now follows a result-first workflow. The survey model is fitted in
its own chunk, the summary is shown immediately after, and each `rineq`
decomposition method is built in its own chunk before being displayed.

```{r fit-classical-model}
library(survey)
library(rineq)
library(margins)

# Fit the same determinant set in a survey-weighted logistic regression.
linear_decomp_formula <- as.formula(
  paste("deadu5_num ~", paste(raw_congo_predictors, collapse = " + "))
)

# Include DHS clustering and weights in the survey design object.
linear_decomp_design <- survey::svydesign(
  id = as.formula("~PSU+household"),
  weights = as.formula("~sample_weight"),
  data = raw_congo_model_dt
)

# This is the regression model used for the indirect decomposition.
linear_decomp_fit <- survey::svyglm(
  linear_decomp_formula,
  design = linear_decomp_design,
  family = quasibinomial()
)
```

```{r classical-model-summary, purl=FALSE}
# Print the fitted regression model for checking before decomposition.
print(summary(linear_decomp_fit))
```

```{r extract-classical-model-components}
# Extract the model matrix because rineq decomposes term-level contributions.
linear_x <- model.matrix(linear_decomp_formula, data = raw_congo_model_dt)
linear_beta <- coef(linear_decomp_fit)
linear_terms <- setdiff(colnames(linear_x), "(Intercept)")

# Convert regression coefficients into average marginal effects on the risk scale.
linear_margins <- tryCatch(
  margins::margins(
    linear_decomp_fit,
    design = linear_decomp_design,
    type = "response"
  ),
  error = function(e) {
    stop(
      "margins::margins() does not work with the svyglm object: ",
      conditionMessage(e),
      call. = FALSE
    )
  }
)
linear_ame_table <- as.data.table(summary(linear_margins))

# Different margins versions use different column names for the AME.
ame_col <- intersect(c("AME", "dydx"), names(linear_ame_table))[1L]
if (is.na(ame_col)) {
  stop("Could not identify the average marginal effect column in margins output.")
}
if (!"factor" %in% names(linear_ame_table)) {
  stop("margins output does not contain a factor column for term alignment.")
}
linear_average_marginal_effects <- setNames(
  linear_ame_table[[ame_col]],
  linear_ame_table$factor
)

# Stop early if the marginal effects cannot be aligned with the model matrix.
missing_from_margins <- setdiff(linear_terms, names(linear_average_marginal_effects))
extra_from_margins <- setdiff(names(linear_average_marginal_effects), linear_terms)
if (length(missing_from_margins) > 0L) {
  stop(
    "AME names from margins do not match model-matrix terms: ",
    paste(missing_from_margins, collapse = ", "),
    call. = FALSE
  )
}

# Store coefficient, marginal-effect, and weighted-mean information by term.
linear_term_info <- data.table(
  term = linear_terms,
  variable = canonicalize_regression_terms(linear_terms),
  linear_coefficient = unname(linear_beta[linear_terms]),
  average_marginal_effect = unname(linear_average_marginal_effects[linear_terms]),
  mean_of_variable = vapply(
    linear_terms,
    function(term) weighted_mean_safe(linear_x[, term], raw_congo_model_dt$sample_weight),
    numeric(1L)
  )
)

# Keep a tidy regression summary for the appendix tables.
regression_model_summary <- as.data.table(
  as.data.frame(broom::tidy(linear_decomp_fit, conf.int = TRUE))
)
regression_model_summary[, variable := fifelse(
  term == "(Intercept)",
  term,
  canonicalize_regression_terms(term)
)]
regression_model_summary[, term_label := fifelse(
  term == "(Intercept)",
  "Intercept",
  format_variable_label(variable)
)]
setcolorder(
  regression_model_summary,
  c("term", "term_label", "variable", setdiff(names(regression_model_summary), c("term", "term_label", "variable")))
)
```

```{r margins-rineq-decomposition-helper}
# Use rineq's internal decomposition function with our marginal effects.
rineq_decomposition <- getFromNamespace("decomposition", "rineq")
expected_decomposition_args <- c(
  "outcome",
  "betas",
  "mm",
  "ranker",
  "wt",
  "correction",
  "citype"
)
actual_decomposition_args <- names(formals(rineq_decomposition))

# Check the function signature because this is an internal rineq function.
if (!all(expected_decomposition_args %in% actual_decomposition_args)) {
  stop(
    "Rineq's internal decomposition() function has different argument names. ",
    "Expected: ",
    paste(expected_decomposition_args, collapse = ", "),
    ". Actual: ",
    paste(actual_decomposition_args, collapse = ", "),
    call. = FALSE
  )
}

# Build one rineq decomposition for a requested concentration-index type.
build_margins_rineq_decomposition <- function(type) {
  tryCatch(
    rineq_decomposition(
      outcome = raw_congo_model_dt$deadu5_num,
      betas = linear_average_marginal_effects[linear_terms],
      mm = linear_x[, linear_terms, drop = FALSE],
      ranker = raw_congo_model_dt$wealth,
      wt = raw_congo_model_dt$sample_weight,
      correction = TRUE,
      citype = type
    ),
    error = function(e) {
      stop(
        "Margins-based Rineq decomposition failed for ",
        type,
        ": ",
        conditionMessage(e),
        call. = FALSE
      )
    }
  )
}
```

```{r decomposition-to-table-helper}
# Convert rineq output into a table that can be displayed and saved.
decomposition_to_table <- function(object, digits = 4L) {
  data.table(
    term = names(object$rel_contribution),
    `Contribution (%)` = round(object$rel_contribution, digits),
    `Contribution (Abs)` = round(object$ci_contribution, digits),
    Elasticity = round(c(0, object$elasticities), digits),
    `Concentration Index` = c(NA_real_, round(object$partial_cis, digits)),
    `lower 5%` = c(NA_real_, round(object$confints[1, ], digits)),
    `upper 5%` = c(NA_real_, round(object$confints[2, ], digits)),
    Corrected = c("", ifelse(object$corrected_coefficients, "yes", "no"))
  )
}
```

```{r summarise-rineq-decomposition-helper}
# Add labels and auxiliary regression quantities to the rineq contribution table.
summarise_rineq_decomposition <- function(object, method_name) {
  out <- merge(
    linear_term_info,
    decomposition_to_table(object, digits = 6L)[
      term != "residual",
      .(
        term,
        elasticity = Elasticity,
        concentration_index_of_variable = `Concentration Index`,
        regression_contribution = `Contribution (Abs)`,
        regression_pct_contribution = `Contribution (%)`
      )
    ],
    by = "term",
    all.y = TRUE
  )
  out[, variable := fifelse(!is.na(variable) & nzchar(variable), variable, term)]
  out[, `:=`(
    criterion = method_name,
    variable_label = format_variable_label(variable)
  )]

  setcolorder(
    out,
    c(
      "variable",
      "variable_label",
      "criterion",
      "linear_coefficient",
      "average_marginal_effect",
      "mean_of_variable",
      "elasticity",
      "concentration_index_of_variable",
      "regression_contribution",
      "regression_pct_contribution"
    )
  )
  out[order(-abs(regression_pct_contribution))]
}
```

```{r build-rineq-ci-decomposition}
# Run the standard concentration-index regression decomposition.
rineq_ci_decomp <- build_margins_rineq_decomposition("CI")
rineq_ci_summary <- decomposition_to_table(rineq_ci_decomp, digits = 4L)
classical_ci_table <- summarise_rineq_decomposition(rineq_ci_decomp, "CI")
```

```{r rineq-ci-output, purl=FALSE, fig.height=7}
report_compact_table(
  rineq_ci_summary,
  digits = 4,
  caption = "rineq decomposition summary: CI"
)
plot(
  rineq_ci_decomp,
  horiz = TRUE,
  decreasing = FALSE,
  main = "rineq decomposition contributions: CI"
)
```

```{r build-rineq-cig-decomposition}
# Run the generalized concentration-index regression decomposition.
rineq_cig_decomp <- build_margins_rineq_decomposition("CIg")
rineq_cig_summary <- decomposition_to_table(rineq_cig_decomp, digits = 4L)
classical_cig_table <- summarise_rineq_decomposition(rineq_cig_decomp, "CIg")
```

```{r rineq-cig-output, purl=FALSE, fig.height=7}
report_compact_table(
  rineq_cig_summary,
  digits = 4,
  caption = "rineq decomposition summary: CIg"
)
plot(
  rineq_cig_decomp,
  horiz = TRUE,
  decreasing = FALSE,
  main = "rineq decomposition contributions: CIg"
)
```

```{r build-rineq-cic-decomposition}
# Run the corrected concentration-index regression decomposition.
rineq_cic_decomp <- build_margins_rineq_decomposition("CIc")
rineq_cic_summary <- decomposition_to_table(rineq_cic_decomp, digits = 4L)
classical_cic_table <- summarise_rineq_decomposition(rineq_cic_decomp, "CIc")
```

```{r rineq-cic-output, purl=FALSE, fig.height=7}
report_compact_table(
  rineq_cic_summary,
  digits = 4,
  caption = "rineq decomposition summary: CIc"
)
plot(
  rineq_cic_decomp,
  horiz = TRUE,
  decreasing = FALSE,
  main = "rineq decomposition contributions: CIc"
)
```

```{r build-rineq-ciw-decomposition}
# Run the Wagstaff-style concentration-index decomposition for comparison.
rineq_ciw_decomp <- build_margins_rineq_decomposition("CIw")
rineq_ciw_summary <- decomposition_to_table(rineq_ciw_decomp, digits = 4L)
classical_ciw_table <- summarise_rineq_decomposition(rineq_ciw_decomp, "CIw")
```

```{r rineq-ciw-output, purl=FALSE, fig.height=7}
report_compact_table(
  rineq_ciw_summary,
  digits = 4,
  caption = "rineq decomposition summary: CIw"
)
plot(
  rineq_ciw_decomp,
  horiz = TRUE,
  decreasing = FALSE,
  main = "rineq decomposition contributions: CIw"
)
```

```{r assemble-rineq-objects}
# Keep full rineq objects because the report may need diagnostics or plots.
rineq_decomp_by_type <- list(
  CI = rineq_ci_decomp,
  CIg = rineq_cig_decomp,
  CIc = rineq_cic_decomp,
  CIw = rineq_ciw_decomp
)

rineq_summary_by_type <- list(
  CI = rineq_ci_summary,
  CIg = rineq_cig_summary,
  CIc = rineq_cic_summary,
  CIw = rineq_ciw_summary
)

# Bind the criterion-specific summaries into one long table.
rineq_summary <- rbindlist(lapply(names(rineq_summary_by_type), function(type) {
  out <- copy(rineq_summary_by_type[[type]])
  out[, criterion := type]
  setcolorder(out, c("criterion", setdiff(names(out), "criterion")))
  out
}), fill = TRUE)

# This is the regression decomposition table used for comparison with SHAP.
classical_decomposition_table <- rbindlist(
  list(classical_ci_table, classical_cig_table, classical_cic_table, classical_ciw_table),
  fill = TRUE
)
```

```{r published-2007-reference}
# Enter the published DRC 2007 benchmark values used for comparison.
published_drc_table3_reference <- rbindlist(list(
  data.table(
    criterion = "G",
    determinant = c(
      "Household wealth",
      "Skilled birth attendance",
      "Sex of the child",
      "Birth order and interval",
      "Mother's age at birth",
      "Type of residence",
      "Mother's education",
      "Father's education",
      "Mother's occupation",
      "Father's occupation",
      "Region"
    ),
    published_index = -0.128,
    published_percent = c(2.4, 7.3, 5.3, 35.6, 0.4, 3.4, 17.4, -0.7, 5.8, 9.1, 14.0)
  ),
  data.table(
    criterion = "C",
    determinant = c(
      "Household wealth",
      "Skilled birth attendance",
      "Sex of the child",
      "Birth order and interval",
      "Mother's age at birth",
      "Type of residence",
      "Mother's education",
      "Father's education",
      "Mother's occupation",
      "Father's occupation",
      "Region"
    ),
    published_index = 0.057,
    published_percent = c(13.1, 11.3, -0.6, -1.2, 0.4, 12.5, 33.1, -2.1, 17.6, 20.8, -4.9)
  )
))

# Map both original variables and expanded indicators back to determinant groups.
determinant_map <- data.table(
  variable = c("quint", "unskilled", "male", "birth", "agemoth", "rural", "ed", "ped", "mocc", "pocc", "reg"),
  raw_variable = c("quint", "unskilled", "male", "birth", "agemoth", "rural", "ed", "ped", "mocc", "pocc", "reg"),
  determinant = c(
    "Household wealth",
    "Skilled birth attendance",
    "Sex of the child",
    "Birth order and interval",
    "Mother's age at birth",
    "Type of residence",
    "Mother's education",
    "Father's education",
    "Mother's occupation",
    "Father's occupation",
    "Region"
  )
)
determinant_map <- rbindlist(
  list(
    determinant_map,
    merge(
      indicator_level_map[, .(variable, raw_variable)],
      determinant_map[, .(raw_variable, determinant)],
      by = "raw_variable",
      all.x = TRUE,
      sort = FALSE
    )[, .(variable, raw_variable, determinant)]
  ),
  fill = TRUE
)

# Aggregate term-level decomposition results to the determinant groups in the paper.
classical_decomposition_grouped <- classical_decomposition_table[
  determinant_map,
  on = "variable",
  nomatch = 0L
][
  ,
  .(
    regression_contribution = sum(regression_contribution, na.rm = TRUE),
    regression_pct_contribution = sum(regression_pct_contribution, na.rm = TRUE)
  ),
  by = .(criterion, determinant)
]

# Compare the generated CI decomposition with the published C-column benchmark.
published_drc_table3_comparison <- merge(
  published_drc_table3_reference[criterion == "C"],
  classical_decomposition_grouped[criterion == "CI"],
  by = "determinant",
  all.x = TRUE,
  sort = FALSE
)
published_drc_table3_comparison[
  ,
  difference_from_published := regression_pct_contribution - published_percent
]
```

```{r published-2007-reference-preview, purl=FALSE}
report_compact_table(
  published_drc_table3_comparison[
    ,
    .(
      determinant,
      paper_c_percent = published_percent,
      generated_ci_percent = regression_pct_contribution,
      difference_from_published
    )
  ],
  digits = 2,
  caption = "Published DRCongo 2007 concentration-index decomposition compared with generated regression decomposition"
)
```

# Fit Concentration-Index Trees

This is the core generator step. The grid is cross-validated, then each
criterion is refitted at its selected setting.

```{r tree-control-grid}
# Define the stopping-rule grid for concentration-index tree tuning.
tree_grid <- ci_tree_control_grid(
    minsplit = c(500L),
    minbucket = c(100L),
    minprob = 0.01,
    maxdepth = 10L,
    min_gain = 0.0001,
    min_relative_gain = c(0.15,  0.20, 0.25,  0.30)
)



# Keep only valid settings where terminal nodes can satisfy the split rule.
tree_grid <- tree_grid[minbucket <= 0.5 * minsplit]
```

```{r make-tree-psu-folds}
# Create cross-validation folds at the PSU level rather than the child level.
make_psu_fold_id <- function(psu,
                             outcome,
                             weights = NULL,
                             v = 10L,
                             seed = 20260508) {
    psu <- as.integer(psu)
    outcome <- as.integer(outcome)
    if (length(psu) != length(outcome)) {
        stop("`psu` and `outcome` must have the same length.", call. = FALSE)
    }
    if (anyNA(psu)) {
        stop("`psu` cannot contain missing values.", call. = FALSE)
    }
    if (is.null(weights)) {
        weights <- rep(1, length(psu))
    }
    weights <- as.numeric(weights)

    psu_frame <- data.table(
        PSU = psu,
        outcome = outcome,
        weight = weights
    )[
        ,
        .(
            rows = .N,
            deaths = sum(outcome == 1L, na.rm = TRUE),
            weighted_deaths = sum(weight * outcome, na.rm = TRUE),
            weighted_rows = sum(weight, na.rm = TRUE)
        ),
        by = PSU
    ]

    # Use no more folds than the number of available PSUs.
    v <- max(2L, as.integer(v[1L]))
    v <- min(v, nrow(psu_frame))

    # Order PSUs so deaths are spread across folds as evenly as possible.
    set.seed(seed)
    psu_frame[, random_order := runif(.N)]
    setorder(psu_frame, -deaths, -weighted_deaths, random_order)
    psu_frame[, fold_id := NA_integer_]

    fold_load <- data.table(
        fold_id = seq_len(v),
        deaths = 0,
        weighted_deaths = 0,
        rows = 0,
        weighted_rows = 0
    )

    # Greedily assign each PSU to the fold with the smallest current burden.
    for (i in seq_len(nrow(psu_frame))) {
        setorder(fold_load, deaths, weighted_deaths, rows, fold_id)
        chosen_fold <- fold_load$fold_id[1L]
        psu_frame[i, fold_id := chosen_fold]
        fold_load[
            fold_id == chosen_fold,
            `:=`(
                deaths = deaths + psu_frame$deaths[i],
                weighted_deaths = weighted_deaths + psu_frame$weighted_deaths[i],
                rows = rows + psu_frame$rows[i],
                weighted_rows = weighted_rows + psu_frame$weighted_rows[i]
            )
        ]
    }

    row_frame <- data.table(row_id = seq_along(psu), PSU = psu)
    row_frame[psu_frame[, .(PSU, fold_id)], on = "PSU", fold_id := i.fold_id]
    row_frame$fold_id
}

# Build the folds used for tree tuning and validation.
tree_psu_fold_id <- make_psu_fold_id(
    psu = raw_congo_model_dt$PSU,
    outcome = congo_model_dt$deadu5_num,
    weights = congo_model_dt$sample_weight,
    v = runtime_params$tuning_folds,
    seed = 20260508
)

# Check that every child from the same PSU stays in the same fold.
psu_fold_check <- data.table(
    PSU = raw_congo_model_dt$PSU,
    fold_id = tree_psu_fold_id
)[, .(fold_count = uniqueN(fold_id)), by = PSU]
if (any(psu_fold_check$fold_count > 1L)) {
    stop("At least one PSU was assigned to more than one fold.", call. = FALSE)
}

# Save a fold summary so the report can describe the validation design.
tree_psu_fold_summary <- data.table(
    fold_id = tree_psu_fold_id,
    PSU = raw_congo_model_dt$PSU,
    deadu5_num = congo_model_dt$deadu5_num,
    sample_weight = congo_model_dt$sample_weight
)[
    ,
    .(
        rows = .N,
        psus = uniqueN(PSU),
        deaths = sum(deadu5_num == 1L, na.rm = TRUE),
        non_deaths = sum(deadu5_num == 0L, na.rm = TRUE),
        weighted_death_percent =
            100 * weighted.mean(deadu5_num, sample_weight, na.rm = TRUE)
    ),
    by = fold_id
][order(fold_id)]
```

```{r tune-ci-trees}
# Evaluate each tree setting with training and PSU-held-out validation metrics.
ci_tuning_metrics <- c(
    "validation_gain",
    "relative_validation_gain",
    "percent_validation_root_recovered",
    "brier",
    "log_loss",
    "roc_auc"
)

# Tune concentration-index trees for all criteria on the same grid.
tree_tuning <- tune_ci_tree(
    formula = congo_ci_formula,
    data = congo_model_dt,
    rank_name = "wealth",
    outcome_name = "deadu5_num",
    weights = congo_model_dt$sample_weight,
    type = criterion_types,
    control_grid = tree_grid,
    fold_id = tree_psu_fold_id,
    seed = 20260508,
    metric = ci_tuning_metrics,
    refit = FALSE,
    control = control_ci_tune(save_pred = TRUE)
)


# Save and reload the tuning object so later chunks use the same on-disk object.
save(tree_tuning, file = tree_tuning_file)
load(tree_tuning_file)
```

```{r select-ci-tree-settings}
# Select the tree setting that keeps the largest relative validation gain.
tree_best_by_type <- ci_select_best(tree_tuning,
    metric = "relative_validation_gain"
)

# Summarise the selected settings and their validation metrics.
tree_selection_table <- ci_fit_summary_table(
    tree_tuning,
    selected = tree_best_by_type,
    metrics = unique(tree_tuning$summary$metric),
    include_percent = FALSE
)
tree_selection_table[
    tree_best_by_type,
    terminal_nodes := i.mean_terminal_nodes,
    on = c("type", "grid_id")
]
tree_selection_table[, selection_basis := paste(
    runtime_params$tuning_folds,
    "PSU-held-out folds"
)]

# Display the selected tree settings in the generator output.
report_compact_table(
    tree_selection_table,
    digits = 4,
    caption = "Summary of the concentration-index tree selection results across criteria"
)
```

```{r fit-selected-ci-trees}
# Refit each criterion on the full analysis data using its selected controls.
tree_models_by_type <- setNames(
    lapply(seq_len(nrow(tree_best_by_type)), function(i) {
        setting <- tree_best_by_type[i]
        ci_tree(
            formula = congo_ci_formula,
            data = congo_model_dt,
            rank_name = "wealth",
            outcome_name = "deadu5_num",
            weights = congo_model_dt$sample_weight,
            type = setting$type[1L],
            control = ci_control_from_row(setting)
        )
    }),
    tree_best_by_type$type
)
```

```{r summarise-selected-ci-tree}
# Use the best validation performer as the main selected tree.
selected_tree_type <- tree_selection_table[order(-mean_validation_relative_gain)][1L, type]
selected_tree_fit <- tree_models_by_type[[selected_tree_type]]

# Store terminal-node summaries for the selected and criterion-specific trees.
selected_tree_summary <- as.data.table(ci_tree_terminal_summary(selected_tree_fit))

all_tree_terminal_summaries <- rbindlist(lapply(names(tree_models_by_type), function(type) {
    out <- as.data.table(ci_tree_terminal_summary(tree_models_by_type[[type]]))
    out[, criterion := type]
    setcolorder(out, "criterion")
    out
}), fill = TRUE)
```

```{r summarise-ci-tree-selection-objects}
# Collect tree variable importance for the results and appendix plots.
tree_variable_importance <- collect_variable_importance(tree_models_by_type)

# Keep the chosen hyperparameters separately for reporting and validation.
tree_best_controls <- tree_best_by_type[
    ,
    .(type, minsplit, minbucket, minprob, maxdepth, min_gain, min_relative_gain)
]

# Build complexity paths showing how tree size changes with stopping rules.
tree_complexity_metrics <- collect_complexity_metrics(tree_tuning)
tree_min_relative_gain_path <- min_relative_gain_path(tree_complexity_metrics,
    selected = tree_best_by_type
)

# Save both wide and long versions of the tuning metrics.
tree_fit_wide <- ci_fit_summary_table(
    tree_tuning,
    selected = tree_best_by_type,
    metrics = unique(tree_tuning$summary$metric),
    include_percent = FALSE
)
tree_fit_table <- ci_collect_metrics(
    tree_tuning,
    selected = tree_best_by_type,
    metric = unique(tree_tuning$summary$metric),
    format = "tidy",
    include_train = TRUE
)
```

```{r tree-selection-preview, purl=FALSE}
report_compact_table(
  tree_selection_table,
  digits = 4,
  caption = "Selection summary for the fitted concentration-index trees"
)
```

```{r selected-tree-preview, purl=FALSE, fig.width=12, fig.height=8}
ci_report_tree_plot(
    fit = selected_tree_fit,
    data = congo_model_dt,
    outcome_name = "deadu5_num",
    outcome_label = "U5 death",
    ci_type = "L",
    var_labels = congo_var_labels
)
```

```{r ci-tree-preview, purl=FALSE, fig.width=12, fig.height=8}
ci_report_tree_plot(
    fit = tree_models_by_type[["CI"]],
    data = congo_model_dt,
    outcome_name = "deadu5_num",
    outcome_label = "U5 death",
    ci_type = "CI",
    var_labels = congo_var_labels
)
```

```{r cig-tree-preview, purl=FALSE, fig.width=12, fig.height=8}
ci_report_tree_plot(
    fit = tree_models_by_type[["CIg"]],
    data = congo_model_dt,
    outcome_name = "deadu5_num",
    outcome_label = "U5 death",
    ci_type = "CIg",
    var_labels = congo_var_labels
)
```

```{r cic-tree-preview, purl=FALSE, fig.width=12, fig.height=8}
ci_report_tree_plot(
    fit = tree_models_by_type[["CIc"]],
    data = congo_model_dt,
    outcome_name = "deadu5_num",
    outcome_label = "U5 death",
    ci_type = "CIc",
    var_labels = congo_var_labels
)
```

# Fit Lecturer Rpart Comparison Trees

The benchmark rpart trees are retained because they are easy to discuss beside
the inequality-driven tree.

```{r build-rpart-comparison}
# Fit the rpart comparison trees using the same analysis variables.
lecturer_rpart_models <- fit_ci_rpart_comparison(
  data = congo_model_dt,
  predictors = congo_predictors,
  outcome = "deadu5_num",
  rank = "wealth",
  weights = "sample_weight"
)

# Summarise rpart terminal nodes so they can be compared with the main trees.
lecturer_rpart_summary <- summarise_ci_rpart_comparison(lecturer_rpart_models)
```

```{r rpart-comparison-preview, purl=FALSE}
# Show only terminal nodes because those are the final subgroups.
rpart_summary_out <- copy(lecturer_rpart_summary)[is_terminal == TRUE]

report_compact_table(
  rpart_summary_out,
  digits = 4,
  caption = "Terminal-node summaries for the lecturer rpart comparison trees"
)
```

```{r rpart-comparison-plots, purl=FALSE, fig.width=12, fig.height=8, results='asis'}
# Give the two comparison trees readable plot titles.
plot_labels <- c(
    simple_abs = "Lecturer rpart method 1: absolute CI impurity",
    robust_abs = "Lecturer rpart method 2: robust absolute CI impurity"
)
plot_fontsize <- c(simple_abs = 4.2, robust_abs = 4.6)

# Plot each rpart comparison tree in the same style as the main CI trees.
for (method_name in names(lecturer_rpart_models)) {
    cat("\n\n### ", plot_labels[[method_name]] %||% method_name, "\n\n", sep = "")
    ci_report_tree_plot(
        fit = lecturer_rpart_models[[method_name]],
        data = congo_model_dt,
        outcome_name = "deadu5_num",
        outcome_label = "U5 death",
        ci_type = "CI",
        var_labels = congo_var_labels,
        fontsize = plot_fontsize[[method_name]]
    )
}
```

# Forest and SHAP Objects



```{r forest-control-grid}
# Define the forest tuning grid. The forest uses the same split criteria as the trees.
p <- length(congo_predictors)
forest_grid <- ci_tree_control_grid(
    minsplit = 500L,
    minbucket = 100L,
    minprob = 0.01,
    maxdepth = 15L,
    min_gain = 0.00001,
    min_relative_gain = c(0.10, 0.15, 0.20, 0.25),
    mtry = c(15),
    ntree = c(50L, 100L)
)
```

```{r tune-ci-forests}


# The full forest tuning is expensive, so the saved tuning object is loaded here.
# forest_tuning <- run_forest_tuning_batches(
#   formula = congo_ci_formula,
#   data = congo_model_dt,
#   forest_grid = forest_grid,
#   criterion_types = criterion_types,
#   tuning_metrics = ci_tuning_metrics,
#   tuning_selection_metric = "relative_validation_gain",
#   rank_name = "wealth",
#   outcome_name = "deadu5_num",
#   weights = congo_model_dt$sample_weight,
#   folds = runtime_params$tuning_folds,
#   workers = 4L,
#   fold_id = tree_psu_fold_id,
#   progress_steps = min(20L, nrow(forest_grid)),
#   log_file = here("logs", "forest_tuning.log"),
#   seed = 20260508,
#   perturb = list(replace = FALSE, fraction = 0.7)
# )


# library(here)
# save(forest_tuning, file = here("data", "forest_tuning.rdata"))

load(here("data", "forest_tuning.rdata"))
```

```{r select-ci-forest-settings}
# Select forest settings using the same relative validation gain criterion.
forest_best_by_type <- ci_select_best(
  forest_tuning,
  metric = "relative_validation_gain"
)
```

```{r fit-selected-ci-forests}
set.seed(20260509)

# The selected forests were fitted once and saved because this step is slow.
# forest_models_by_type <- setNames(
#   lapply(seq_len(nrow(forest_best_by_type)), function(i) {
#     setting <- forest_best_by_type[i]
#     ci_forest(
#       formula = congo_ci_formula,
#       data = congo_model_dt,
#       rank_name = "wealth",
#       outcome_name = "deadu5_num",
#       weights = congo_model_dt$sample_weight,
#       type = setting$type[1L],
#       control = ci_control_from_row(setting, include_mtry = FALSE),
#       ntree = as.integer(setting$ntree[1L]),
#       mtry = as.integer(setting$mtry[1L]),
#       perturb = list(replace = FALSE, fraction = 0.632)
#     )
#   }),
#   forest_best_by_type$type
# )

# save(forest_models_by_type, file = here("data", "forest_models_by_type.rdata"))
load(here("data", "forest_models_by_type.rdata"))

# Keep the forest matching the selected tree criterion for the main comparison.
selected_forest_fit <- forest_models_by_type[[selected_tree_type]]

# Collect forest variable importance for the report plots.
forest_variable_importance <- collect_variable_importance(forest_models_by_type)
```

```{r summarise-ci-forest-complexity}
# Summarise how forest complexity varies across tuning settings.
forest_complexity_metrics <- collect_complexity_metrics(forest_tuning)
forest_min_relative_gain_path <- min_relative_gain_path(
  forest_complexity_metrics,
  selected = forest_best_by_type
)
```

```{r summarise-ci-forest-selection}
# Build the table that records selected forest settings and validation metrics.
forest_selection_table <- ci_fit_summary_table(
    forest_tuning,
    selected = forest_best_by_type,
    metrics = unique(forest_tuning$summary$metric),
    include_percent = FALSE
)

# Add ensemble size and mtry because those are forest-specific controls.
forest_selection_table <- merge(
    forest_selection_table,
    forest_best_by_type[, .(type, ntree, mtry)],
    by = "type",
    all.x = TRUE,
    sort = FALSE
)
forest_selection_table[
    forest_best_by_type,
    `:=`(
        mean_terminal_nodes = i.mean_terminal_nodes,
        mean_max_depth = i.maxdepth
    ),
    on = "type"
]
forest_selection_table[, selection_basis := paste(
    runtime_params$tuning_folds,
    "fold cross-validation"
)]

# Keep both display and tidy versions of the forest validation results.
forest_fit_wide <- copy(forest_selection_table)
forest_fit_table <- ci_collect_metrics(
    forest_tuning,
    selected = forest_best_by_type,
    metric = unique(forest_tuning$summary$metric),
    format = "tidy",
    include_train = TRUE
)

# Save the selected forest controls for the report appendix.
forest_best_controls <- forest_best_by_type[
    ,
    .(
        type, ntree, mtry, minsplit, minbucket,
        minprob, maxdepth,
        min_gain, min_relative_gain
    )
]
```

```{r fit-ci-forest-surrogates}
# Fit a readable surrogate tree to each selected concentration-index forest.
forest_surrogates_by_type <- setNames(
  lapply(seq_len(nrow(forest_best_by_type)), function(i) {
    # Use the selected tree setting as the starting point for the surrogate.
    setting <- copy(tree_best_by_type[type == forest_best_by_type$type[i]])
    type = setting$type[1L]

    # Adjust surrogate complexity so the plotted tree remains readable.
    if (type == "CIc"){
      setting$min_relative_gain <- 0.15
    }else if (type == "L"){
      setting$min_relative_gain <- 0.3
      setting$maxdepth <- 3L
    } else {
      setting$min_relative_gain <- 0.3
      setting$maxdepth <- 4L
    }

    # Use a larger terminal-node size to avoid very small surrogate leaves.
    setting$minbucket <- 500L
    ci_forest_surrogate(
      forest_fit = forest_models_by_type[[setting$type[1L]]],
      data = congo_model_dt,
      formula = congo_ci_formula,
      rank_name = "wealth",
      weights = congo_model_dt$sample_weight,
      type = setting$type[1L],
      control = ci_control_from_row(setting, include_mtry = FALSE),
      prediction_name = "forest_risk"
    )
  }),
  forest_best_by_type$type
)

# Use the selected criterion's surrogate as the main forest summary tree.
selected_forest_surrogate <- forest_surrogates_by_type[[selected_tree_type]]
selected_forest_surrogate_summary <- as.data.table(
  ci_tree_terminal_summary(selected_forest_surrogate$fit)
)
```

```{r ci-forest-surrogate-preview, purl=FALSE, fig.width=12, fig.height=8}
# Preview the CI forest surrogate tree.
ci_report_tree_plot(
  fit = forest_surrogates_by_type[["CI"]]$fit,
  data = forest_surrogates_by_type[["CI"]]$data,
  outcome_name = forest_surrogates_by_type[["CI"]]$prediction_name,
  outcome_label = "U5",
  ci_type = "CI",
  var_labels = congo_var_labels
)
```

```{r cig-forest-surrogate-preview, purl=FALSE, fig.width=12, fig.height=8}
# Preview the CIg forest surrogate tree.
ci_report_tree_plot(
  fit = forest_surrogates_by_type[["CIg"]]$fit,
  data = forest_surrogates_by_type[["CIg"]]$data,
  outcome_name = forest_surrogates_by_type[["CIg"]]$prediction_name,
  outcome_label = "U5",
  ci_type = "CIg",
  var_labels = congo_var_labels
)
```

```{r cic-forest-surrogate-preview, purl=FALSE, fig.width=12, fig.height=8}
# Preview the CIc forest surrogate tree.
ci_report_tree_plot(
  fit = forest_surrogates_by_type[["CIc"]]$fit,
  data = forest_surrogates_by_type[["CIc"]]$data,
  outcome_name = forest_surrogates_by_type[["CIc"]]$prediction_name,
  outcome_label = "U5",
  ci_type = "CIc",
  var_labels = congo_var_labels
)
```


```{r l-forest-surrogate-preview, purl=FALSE, fig.width=12, fig.height=8}
# Preview the level-dependent forest surrogate tree.
ci_report_tree_plot(
  fit = forest_surrogates_by_type[["L"]]$fit,
  data = forest_surrogates_by_type[["L"]]$data,
  outcome_name = forest_surrogates_by_type[["L"]]$prediction_name,
  outcome_label = "U5",
  ci_type = "L",
  var_labels = congo_var_labels
)
```


```{r build-ranger-analysis-data}
library(tidymodels)

# Build the predictive-forest data, including the undersampled fitting set.
ranger_analysis <- build_ranger_analysis_data(
  data = data.table(
    raw_congo_model_dt[, .(PSU)],
    congo_model_dt
  ),
  predictor_terms = congo_predictors,
  undersample = runtime_params$undersample,
  outcome_labels = outcome_labels
)

# Expose the returned ranger data objects in the current workflow environment.
list2env(ranger_analysis, envir = environment())
```

```{r build-ranger-recipe}
# Use all prepared predictor columns to predict under-five mortality status.
ranger_recipe <- recipe(deadu5_factor ~ ., data = ranger_fit_df)
```

```{r build-ranger-model-spec}
# Tune mtry and min_n while keeping a fixed number of trees.
ranger_model_spec <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 500L
) |>
  set_mode("classification") |>
  set_engine("ranger", probability = TRUE, importance = "impurity")
```

```{r build-ranger-resamples}
# Convert the PSU fold ids into a tidymodels resampling object.
make_manual_fold_rset <- function(data, fold_id) {
  fold_values <- sort(unique(as.integer(fold_id)))
  splits <- lapply(fold_values, function(fold_value) {
    rsample::make_splits(
      list(
        analysis = as.integer(which(fold_id != fold_value)),
        assessment = as.integer(which(fold_id == fold_value))
      ),
      data = data
    )
  })
  rsample::manual_rset(
    splits,
    ids = sprintf("Fold%02d", seq_along(fold_values))
  )
}

# Create PSU-held-out folds for the predictive ranger model.
ranger_psu_fold_id <- make_psu_fold_id(
  psu = ranger_dt$PSU,
  outcome = as.integer(ranger_dt$deadu5_factor == "Died"),
  weights = ranger_dt$sample_weight,
  v = runtime_params$tuning_folds,
  seed = 20260508
)

# Check that ranger resampling also keeps each PSU in one fold.
ranger_psu_fold_check <- data.table(
  PSU = ranger_dt$PSU,
  fold_id = ranger_psu_fold_id
)[, .(fold_count = uniqueN(fold_id)), by = PSU]
if (any(ranger_psu_fold_check$fold_count > 1L)) {
  stop("At least one ranger PSU was assigned to more than one fold.", call. = FALSE)
}

# Summarise the ranger folds to document the validation split.
ranger_psu_fold_summary <- data.table(
  fold_id = ranger_psu_fold_id,
  PSU = ranger_dt$PSU,
  deadu5_factor = ranger_dt$deadu5_factor,
  sample_weight = ranger_dt$sample_weight
)[
  ,
  .(
    rows = .N,
    psus = uniqueN(PSU),
    deaths = sum(deadu5_factor == "Died", na.rm = TRUE),
    non_deaths = sum(deadu5_factor == "Survived", na.rm = TRUE),
    weighted_death_percent =
      100 * weighted.mean(deadu5_factor == "Died", sample_weight, na.rm = TRUE)
  ),
  by = fold_id
][order(fold_id)]

# Build the manual resampling object used by tune_grid().
ranger_resamples <- make_manual_fold_rset(
  data = ranger_fit_df,
  fold_id = ranger_psu_fold_id
)
```

```{r build-ranger-workflow}
# Combine the model, recipe, and case weights in one tidymodels workflow.
ranger_workflow <- workflow() |>
  add_model(ranger_model_spec) |>
  add_recipe(ranger_recipe) |>
  add_case_weights(sample_weight_case)
```

```{r build-ranger-grid}
# Define the ranger hyperparameter grid.
ranger_parameter_set <- hardhat::extract_parameter_set_dials(ranger_workflow) |>
  update(
    mtry = dials::mtry(range = c(10L, length(ranger_predictor_terms))),
    min_n = dials::min_n(range = c(200L, 800L))
  )
ranger_grid <- dials::grid_regular(
  ranger_parameter_set,
  levels = c(mtry = 4, min_n = 3)
)
```

```{r tune-ranger-model}
# Tune the predictive forest with classification metrics.
ranger_metric_set <- yardstick::metric_set(
  yardstick::roc_auc,
  yardstick::accuracy,
  yardstick::sens,
  yardstick::spec,
  yardstick::f_meas
)

# Save predictions because the probability threshold is tuned separately.
ranger_tune_results <- tune_grid(
  ranger_workflow,
  resamples = ranger_resamples,
  grid = ranger_grid,
  metrics = ranger_metric_set,
  control = control_grid(save_pred = TRUE)
)

# Store tuning summaries and fold predictions for the report appendix.
ranger_tuning <- as.data.table(collect_metrics(ranger_tune_results))
ranger_cv_predictions <- as.data.table(collect_predictions(ranger_tune_results))
```

```{r tune-ranger-probability-threshold}
# Search probability cutoffs because deaths are relatively rare.
ranger_threshold_grid <- seq(0.05, 0.95, by = 0.01)

ranger_threshold_performance <- rbindlist(lapply(ranger_threshold_grid, function(cutoff) {
  threshold_dt <- copy(ranger_cv_predictions)

  # Convert predicted probabilities into classes at the current cutoff.
  threshold_dt[, pred_threshold := factor(
    fifelse(.pred_Died >= cutoff, "Died", "Survived"),
    levels = levels(deadu5_factor)
  )]

  threshold_dt[
    ,
    .(
      threshold = cutoff,
      accuracy = mean(pred_threshold == deadu5_factor),
      sensitivity = yardstick::sens_vec(
        truth = deadu5_factor,
        estimate = pred_threshold,
        event_level = "first"
      ),
      specificity = yardstick::spec_vec(
        truth = deadu5_factor,
        estimate = pred_threshold,
        event_level = "first"
      ),
      f_meas = yardstick::f_meas_vec(
        truth = deadu5_factor,
        estimate = pred_threshold,
        event_level = "first"
      ),
      predicted_died = mean(pred_threshold == "Died"),
      observed_died = mean(deadu5_factor == "Died")
    ),
    by = .config
  ]
}))
# knitr::kable(
#   ranger_threshold_performance,
#   caption = "Performance metrics across probability thresholds for the cross-validated ranger models"
# )
```

```{r ranger-cv-preview, purl=FALSE}
report_compact_table(
  ranger_tuning,
  digits = 4,
  caption = "Predictive ranger cross-validation summary"
)
```

```{r select-best-ranger-model}
# Select the ranger model that gives the best F1 score.
selected_ranger_setting <- select_best(ranger_tune_results, metric = "f_meas")

# Within the selected model, choose the threshold that balances F1 and sensitivity.
selected_ranger_threshold <- ranger_threshold_performance[
  .config == selected_ranger_setting$.config
][
  order(-f_meas, -sensitivity, -specificity, threshold)
][1L]
```

```{r selected-ranger-preview, purl=FALSE}
report_compact_table(
  as.data.table(as.data.frame(selected_ranger_setting)),
  digits = 4,
  caption = "Selected predictive ranger setting"
)
```

```{r}
# Plot threshold sensitivity for the selected ranger configuration.
plot_threshold_dt <- ranger_threshold_performance[
  .config == selected_ranger_setting$.config,
  .(threshold, F1 = specificity, Sensitivity = sensitivity)
]

plot_threshold_long <- melt(
  plot_threshold_dt,
  id.vars = "threshold",
  variable.name = "Metric",
  value.name = "value"
)

ggplot(plot_threshold_long, aes(x = threshold, y = value, colour = Metric)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.4) +
  geom_vline(
    xintercept = selected_ranger_threshold$threshold,
    linetype = "dashed",
    colour = "grey40"
  ) +
  scale_x_continuous(
    breaks = seq(0.05, 0.95, by = 0.10),
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "F1 and Sensitivity vs Probability Threshold",
    x = "Probability threshold for predicting death",
    y = "Metric value",
    colour = "Metric"
  ) +
  theme_minimal()
```


```{r selected-ranger-threshold-preview, purl=FALSE}
report_compact_table(
  selected_ranger_threshold,
  digits = 4,
  caption = "Selected predictive ranger probability threshold"
)
```

```{r fit-selected-ranger-model}
# Fit the final predictive forest on the undersampled ranger training data.
final_ranger_workflow <- finalize_workflow(ranger_workflow, selected_ranger_setting)
predictive_forest_fit <- workflows::fit(final_ranger_workflow, data = ranger_fit_df)
```

```{r predict-ranger-risk-helper}
# Return the fitted probability score for death before age five.
predict_ranger_risk <- function(object, newdata) {
  pred <- workflows:::predict.workflow(
    object,
    new_data = as.data.frame(newdata),
    type = "prob"
  )
  as.numeric(pred$.pred_Died)
}

# Apply the selected threshold when a class label is needed.
predict_ranger_class <- function(
  object,
  newdata,
  threshold = selected_ranger_threshold$threshold
) {
  factor(
    ifelse(predict_ranger_risk(object, newdata) >= threshold, "Died", "Survived"),
    levels = c("Died", "Survived")
  )
}
```

```{r prepare-shap-data}
# Stop with a clear installation message if fastshap is unavailable.
if (!requireNamespace("fastshap", quietly = TRUE)) {
  stop(
    paste(
      "Package fastshap is required for SHAP estimation.",
      "Install it from GitHub with:",
      "remotes::install_github('bgreenwell/fastshap')"
    ),
    call. = FALSE
  )
}

# Use all rows when shap_rows is zero, otherwise use the requested prefix.
shap_rows <- if (shap_rows_requested == 0L) {
  seq_len(nrow(ranger_fit_df))
} else {
  seq_len(min(shap_rows_requested, nrow(ranger_fit_df)))
}

# Prepare background, explanation, and prediction inputs for SHAP.
shap_x_background <- ranger_fit_df[, ranger_predictor_terms, drop = FALSE]
shap_x_eval <- ranger_fit_df[shap_rows, ranger_predictor_terms, drop = FALSE]
shap_prediction <- as.numeric(predict_ranger_risk(predictive_forest_fit, shap_x_eval))

# Record SHAP settings so the final report knows how explanations were built.
shap_parallel_config <- data.table(
  requested_rows = shap_rows_requested,
  explanation_rows = nrow(shap_x_eval),
  nsim = max(2L, shap_n_mc_samples),
  workers = 1L,
  backend = "fastshap serial"
)
```

```{r compute-fastshap-values}
# Estimate SHAP values for the selected predictive ranger forest.
selected_forest_shap <- estimate_fastshap_values(
  object = predictive_forest_fit,
  x_train = shap_x_background,
  x_explain = shap_x_eval,
  pred_wrapper = predict_ranger_risk,
  seed = shap_seed,
  nsim = shap_n_mc_samples,
  adjust = TRUE
)

# Put SHAP column names back into the same naming system used elsewhere.
colnames(selected_forest_shap) <- canonicalize_model_terms(
  colnames(selected_forest_shap),
  congo_predictors
)

# Combine SHAP values with the observed feature values for plotting.
shap_feature_values <- copy(as.data.table(shap_x_eval))
shap_feature_values[, observation_id := seq_len(.N)]
selected_forest_shap_dt <- as.data.table(selected_forest_shap)
selected_forest_shap_dt[, observation_id := seq_len(.N)]

# Store SHAP values in long form because the report plots features vertically.
shap_summary_long <- melt(
  selected_forest_shap_dt,
  id.vars = "observation_id",
  variable.name = "feature",
  value.name = "shap_value",
  variable.factor = FALSE
)

# Store matching feature values in long form.
shap_value_long <- melt(
  shap_feature_values,
  id.vars = "observation_id",
  variable.name = "feature",
  value.name = "feature_value",
  variable.factor = FALSE
)

# Join SHAP values to the original feature values.
shap_summary_long <- merge(
  shap_summary_long,
  shap_value_long,
  by = c("observation_id", "feature"),
  all.x = TRUE,
  sort = FALSE
)

# Convert feature values to numeric values for the colour scale in SHAP plots.
shap_summary_long[
  ,
  feature_value_numeric := suppressWarnings(as.numeric(as.character(feature_value)))
]
shap_summary_long[
  is.na(feature_value_numeric),
  feature_value_numeric := as.numeric(factor(feature_value)) - 1
]

# Scale feature values within each feature for comparable plot colours.
shap_summary_long[
  ,
  feature_value_scaled := {
    range_value <- range(feature_value_numeric, na.rm = TRUE)
    fifelse(
      rep(all(is.finite(range_value)) && diff(range_value) > .Machine$double.eps, .N),
      (feature_value_numeric - range_value[1L]) / diff(range_value),
      0.5
    )
  },
  by = feature
]

# Add report labels and mean absolute SHAP values for feature ordering.
shap_summary_long[, feature_label := format_variable_label(feature)]
shap_summary_long[
  ,
  mean_abs_shap := mean(abs(shap_value), na.rm = TRUE),
  by = feature
]
setcolorder(
  shap_summary_long,
  c(
    "feature",
    "feature_label",
    "observation_id",
    "shap_value",
    "feature_value",
    "feature_value_numeric",
    "feature_value_scaled",
    "mean_abs_shap"
  )
)
```

```{r build-shap-ci-decomposition}
# Decompose the concentration of ranger predictions using standard CI.
shap_ci_decomp <- shap_conc_decomp(
  shap = selected_forest_shap,
  rank = ranger_dt$wealth[shap_rows],
  type = "CI",
  prediction = shap_prediction,
  weights = ranger_dt$sample_weight[shap_rows]
)
```

```{r shap-ci-output, purl=FALSE, fig.height=7}
# Display SHAP diagnostics and percentage contributions for CI.
shap_ci_diag <- as.data.table(shap_ci_decomp$diagnostics)
shap_ci_contrib <- as.data.table(shap_ci_decomp$contributions)[order(pct_contribution)]
report_compact_table(
  shap_ci_diag,
  digits = 4,
  caption = "SHAP decomposition diagnostics: CI"
)
report_compact_table(
  shap_ci_contrib,
  digits = 4,
  caption = "SHAP decomposition contributions: CI"
)
ggplot(shap_ci_contrib, aes(x = pct_contribution, y = reorder(feature, pct_contribution))) +
  geom_col(fill = "#3268A8") +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.4) +
  labs(x = "Contribution (%)", y = NULL, title = "SHAP concentration decomposition: CI") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r build-shap-cig-decomposition}
# Decompose the concentration of ranger predictions using CIg.
shap_cig_decomp <- shap_conc_decomp(
  shap = selected_forest_shap,
  rank = ranger_dt$wealth[shap_rows],
  type = "CIg",
  prediction = shap_prediction,
  weights = ranger_dt$sample_weight[shap_rows]
)
```

```{r shap-cig-output, purl=FALSE, fig.height=7}
# Display SHAP diagnostics and percentage contributions for CIg.
shap_cig_diag <- as.data.table(shap_cig_decomp$diagnostics)
shap_cig_contrib <- as.data.table(shap_cig_decomp$contributions)[order(pct_contribution)]
report_compact_table(
  shap_cig_diag,
  digits = 4,
  caption = "SHAP decomposition diagnostics: CIg"
)
report_compact_table(
  shap_cig_contrib,
  digits = 4,
  caption = "SHAP decomposition contributions: CIg"
)
ggplot(shap_cig_contrib, aes(x = pct_contribution, y = reorder(feature, pct_contribution))) +
  geom_col(fill = "#3268A8") +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.4) +
  labs(x = "Contribution (%)", y = NULL, title = "SHAP concentration decomposition: CIg") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r build-shap-cic-decomposition}
# Decompose the concentration of ranger predictions using CIc.
shap_cic_decomp <- shap_conc_decomp(
  shap = selected_forest_shap,
  rank = ranger_dt$wealth[shap_rows],
  type = "CIc",
  prediction = shap_prediction,
  weights = ranger_dt$sample_weight[shap_rows]
)
```

```{r shap-cic-output, purl=FALSE, fig.height=7}
# Display SHAP diagnostics and percentage contributions for CIc.
shap_cic_diag <- as.data.table(shap_cic_decomp$diagnostics)
shap_cic_contrib <- as.data.table(shap_cic_decomp$contributions)[order(pct_contribution)]
report_compact_table(
  shap_cic_diag,
  digits = 4,
  caption = "SHAP decomposition diagnostics: CIc"
)
report_compact_table(
  shap_cic_contrib,
  digits = 4,
  caption = "SHAP decomposition contributions: CIc"
)
ggplot(shap_cic_contrib, aes(x = pct_contribution, y = reorder(feature, pct_contribution))) +
  geom_col(fill = "#3268A8") +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.4) +
  labs(x = "Contribution (%)", y = NULL, title = "SHAP concentration decomposition: CIc") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r build-shap-l-decomposition}
# Decompose the level-dependent concentration of ranger predictions.
shap_l_decomp <- shap_conc_decomp(
  shap = selected_forest_shap,
  rank = ranger_dt$wealth[shap_rows],
  type = "L",
  prediction = shap_prediction,
  weights = ranger_dt$sample_weight[shap_rows]
)
```

```{r shap-l-output, purl=FALSE, fig.height=7}
# Display SHAP diagnostics and percentage contributions for L.
shap_l_diag <- as.data.table(shap_l_decomp$diagnostics)
shap_l_contrib <- as.data.table(shap_l_decomp$contributions)[order(pct_contribution)]
report_compact_table(
  shap_l_diag,
  digits = 4,
  caption = "SHAP decomposition diagnostics: L"
)
report_compact_table(
  shap_l_contrib,
  digits = 4,
  caption = "SHAP decomposition contributions: L"
)
ggplot(shap_l_contrib, aes(x = pct_contribution, y = reorder(feature, pct_contribution))) +
  geom_col(fill = "#3268A8") +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.4) +
  labs(x = "Contribution (%)", y = NULL, title = "SHAP concentration decomposition: L") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r assemble-shap-objects}
# Bind SHAP contributions across all criteria.
selected_shap_contributions <- rbindlist(list(
  as.data.table(shap_ci_decomp$contributions)[, criterion := "CI"],
  as.data.table(shap_cig_decomp$contributions)[, criterion := "CIg"],
  as.data.table(shap_cic_decomp$contributions)[, criterion := "CIc"],
  as.data.table(shap_l_decomp$contributions)[, criterion := "L"]
), fill = TRUE)
selected_shap_contributions[, feature := canonicalize_model_terms(feature, congo_predictors)]
setcolorder(selected_shap_contributions, c("criterion", setdiff(names(selected_shap_contributions), "criterion")))

# Bind the SHAP diagnostics in the same criterion-indexed format.
selected_shap_diagnostics <- rbindlist(list(
  as.data.table(shap_ci_decomp$diagnostics)[, criterion := "CI"],
  as.data.table(shap_cig_decomp$diagnostics)[, criterion := "CIg"],
  as.data.table(shap_cic_decomp$diagnostics)[, criterion := "CIc"],
  as.data.table(shap_l_decomp$diagnostics)[, criterion := "L"]
), fill = TRUE)
setcolorder(selected_shap_diagnostics, c("criterion", setdiff(names(selected_shap_diagnostics), "criterion")))
```

```{r forest-preview, purl=FALSE}
# Add readable labels before displaying forest variable importance.
forest_importance_out <- copy(forest_variable_importance)
forest_importance_out[, variable_label := fifelse(
  variable %in% names(congo_var_labels),
  congo_var_labels[variable],
  variable
)]

report_compact_table(
  forest_importance_out,
  digits = 4,
  caption = "Forest variable importance preview"
)
```

# Compare Methods

Once the regression, tree, forest, and SHAP objects exist, the next step is to
rank variables and assemble the comparison tables used in the main report.

```{r build-method-comparison}
# Rank variables by the regression decomposition contribution under CI.
linear_rank <- classical_decomposition_table[
  criterion == "CI"
][
  order(-abs(regression_pct_contribution)),
  .(variable, method = "Linear decomposition", rank = seq_len(.N))
]

# Rank the selected tree and forest by their stored variable-importance values.
tree_rank <- rank_from_named_vector(selected_tree_fit$variable.importance, "Selected CI-tree")
forest_rank <- rank_from_named_vector(selected_forest_fit$variable.importance, "Selected CI-forest")

# Rank SHAP variables by absolute contribution under CI.
shap_rank <- selected_shap_contributions[
  criterion == "CI"
][
  order(-abs_contribution),
  .(variable = feature, method = "Selected forest SHAP", rank = seq_len(.N))
]
importance_long <- rbindlist(list(linear_rank, tree_rank, forest_rank, shap_rank), fill = TRUE)

# Keep variables that appear in the regression decomposition comparison table.
selected_variables <- classical_decomposition_table[
  criterion == "CI"
][
  order(-abs(regression_pct_contribution))
]$variable[seq_len(max(15L, classical_decomposition_table[criterion == "CI", .N]))]
selected_variables <- selected_variables[!is.na(selected_variables) & nzchar(selected_variables)]

# Start the comparison table with readable variable labels.
method_comparison_table <- data.table(variable = selected_variables)
method_comparison_table[, variable_label := format_variable_label(variable)]

# Widen regression contributions so each criterion becomes a column.
regression_wide <- dcast(
  classical_decomposition_table[
    criterion %in% c("CI", "CIg", "CIc"),
    .(variable, criterion, regression_pct_contribution)
  ],
  variable ~ criterion,
  value.var = "regression_pct_contribution"
)
method_comparison_table <- merge(
  method_comparison_table,
  regression_wide[
    ,
    .(
      variable,
      regression_ci_pct_contribution = CI,
      regression_cig_pct_contribution = CIg,
      regression_cic_pct_contribution = CIc
    )
  ],
  by = "variable",
  all.x = TRUE
)

# Widen SHAP contributions in the same way for side-by-side comparison.
shap_wide <- dcast(
  selected_shap_contributions[
    criterion %in% criterion_types,
    .(variable = feature, criterion, pct_contribution)
  ],
  variable ~ criterion,
  value.var = "pct_contribution"
)
shap_wide <- shap_wide[
  ,
  .(
    variable,
    forest_shap_ci_pct_contribution = CI,
    forest_shap_cig_pct_contribution = CIg,
    forest_shap_cic_pct_contribution = CIc,
    forest_shap_l_pct_contribution = L
  )
]
method_comparison_table <- merge(
  method_comparison_table,
  shap_wide,
  by = "variable",
  all.x = TRUE
)

# Sort by the main CI regression contribution, then by SHAP contribution.
method_comparison_table <- method_comparison_table[
  order(-abs(regression_ci_pct_contribution), -abs(forest_shap_ci_pct_contribution))
]

# Build a top-rank table comparing the variable order across methods.
importance_wide <- dcast(
  importance_long,
  rank ~ method,
  value.var = "variable",
  fun.aggregate = function(x) paste(unique(x), collapse = ", ")
)

# Replace internal variable names with labels in the rank table.
for (col in setdiff(names(importance_wide), "rank")) {
  importance_wide[, (col) := format_variable_label(get(col))]
}
importance_wide <- importance_wide[rank <= 10L]
```

```{r method-comparison-preview, purl=FALSE}
report_compact_table(
  method_comparison_table,
  digits = 3,
  caption = "Selected percentage-contribution comparison table"
)
```

```{r importance-preview, purl=FALSE}
report_compact_table(
  importance_wide,
  caption = "Ranked variables across decomposition and tree-based methods"
)
```


# Save Results Objects

The last chunk assembles the full object list and writes it to disk for the
main report to consume.

```{r save-results-objects}
# Save everything the thesis report needs in one named list.
drc_results_objects <- list(
  generation_params = runtime_params,
  source_data_file = source_data_file,
  paper_reference_file = paper_reference_file,
  outcome_labels = outcome_labels,
  congo_model_dt = congo_model_dt,
  raw_congo_model_dt = raw_congo_model_dt,
  congo_var_labels = congo_var_labels,
  raw_congo_var_labels = raw_congo_var_labels,
  congo_predictors = congo_predictors,
  raw_congo_predictors = raw_congo_predictors,
  indicator_level_map = indicator_level_map,
  congo_ci_formula = congo_ci_formula,
  data_summary = drc_data_summary_table,
  outcome_distribution = outcome_plot_dt,
  candidate_predictors = candidate_predictor_table,
  root_ci = drc_root_ci_table,
  regression_model_summary = regression_model_summary,
  classical_decomposition = classical_decomposition_table,
  classical_decomposition_grouped = classical_decomposition_grouped,
  published_drc_table3_reference = published_drc_table3_reference,
  published_drc_table3_comparison = published_drc_table3_comparison,
  rineq_summary = rineq_summary,
  tree_psu_fold_id = tree_psu_fold_id,
  tree_psu_fold_summary = tree_psu_fold_summary,
  tree_tuning = tree_tuning,
  tree_best_by_type = tree_best_by_type,
  tree_best_controls = tree_best_controls,
  tree_selection = tree_selection_table,
  tree_fit_table = tree_fit_table,
  tree_complexity_metrics = tree_complexity_metrics,
  tree_min_relative_gain_path = tree_min_relative_gain_path,
  tree_models_by_type = tree_models_by_type,
  selected_tree_type = selected_tree_type,
  selected_tree_fit = selected_tree_fit,
  selected_tree_summary = selected_tree_summary,
  all_tree_terminal_summaries = all_tree_terminal_summaries,
  tree_variable_importance = tree_variable_importance,
  lecturer_rpart_models = lecturer_rpart_models,
  lecturer_rpart_summary = lecturer_rpart_summary,
  forest_tuning = forest_tuning,
  forest_best_by_type = forest_best_by_type,
  forest_best_controls = forest_best_controls,
  forest_fit_table = forest_fit_table,
  forest_complexity_metrics = forest_complexity_metrics,
  forest_min_relative_gain_path = forest_min_relative_gain_path,
  selected_forest_fit = selected_forest_fit,
  forest_surrogates_by_type = forest_surrogates_by_type,
  selected_forest_surrogate = selected_forest_surrogate,
  selected_forest_surrogate_summary = selected_forest_surrogate_summary,
  forest_selection = forest_selection_table,
  forest_variable_importance = forest_variable_importance,
  predictive_forest_tuning = ranger_tuning,
  predictive_forest_cv_predictions = ranger_cv_predictions,
  predictive_forest_threshold_performance = ranger_threshold_performance,
  selected_predictive_forest_threshold = selected_ranger_threshold,
  predictive_forest_sampling = ranger_sampling_summary,
  predictive_forest_psu_fold_id = ranger_psu_fold_id,
  predictive_forest_psu_fold_summary = ranger_psu_fold_summary,
  predictive_forest_feature_lookup = ranger_feature_lookup,
  predictive_forest_fit = predictive_forest_fit,
  shap_parallel_config = shap_parallel_config,
  shap_contributions = selected_shap_contributions,
  shap_diagnostics = selected_shap_diagnostics,
  shap_summary_long = shap_summary_long,
  method_comparison = method_comparison_table,
  importance_wide = importance_wide
)

# This is the main file loaded by MSE_Thesis_Decomposition_of_CI_Trees_MM.qmd.
save(drc_results_objects, file = results_object_file)

# Print a short save summary at the end of the generator.
data.table(
  saved_to = results_object_file,
  object_count = length(drc_results_objects),
  object_names = paste(names(drc_results_objects), collapse = ", ")
)
```
