---
title: "Validation-Gain Path Presentation Experiments"
author: "Moses Mburu"
format:
  html:
    toc: true
    number-sections: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 4.8,
  dpi = 300,
  warning = FALSE,
  message = FALSE
)

library(data.table)
library(ggplot2)
library(knitr)
library(here)

source(here("R", "report_helpers.R"))

results_object_file <- here("data", "drc_report_results_objects.rda")
report_results <- load_drc_report_results(results_object_file)
drc_results_objects <- report_results$drc_results_objects
```

# Purpose

This notebook inspects the exact data frames used by
`plot_validation_percent_gain_path()` and experiments with alternative summaries
that make the distinction between the averaged tuning path and the selected full
grid setting clearer.

The original figure uses:

- `drc_results_objects$tree_min_relative_gain_path`
- `drc_results_objects$forest_min_relative_gain_path`

Inside `plot_validation_percent_gain_path()`, these are copied to `plot_dt`, and
rows where `selected == TRUE` are redrawn as the larger white circle.

```{r helper-functions}
show_table <- function(x, caption = NULL, digits = 3L) {
  out <- copy(as.data.table(x))
  numeric_cols <- names(out)[vapply(out, is.numeric, logical(1))]
  if (length(numeric_cols)) {
    out[, (numeric_cols) := lapply(.SD, round, digits = digits),
        .SDcols = numeric_cols]
  }

  if (requireNamespace("DT", quietly = TRUE)) {
    return(DT::datatable(
      out,
      caption = caption,
      rownames = FALSE,
      filter = "top",
      options = list(pageLength = 20, scrollX = TRUE)
    ))
  }

  knitr::kable(out, caption = caption)
}

path_object <- function(model) {
  object_name <- paste0(model, "_min_relative_gain_path")
  out <- copy(as.data.table(drc_results_objects[[object_name]] %||% data.table()))
  if (nrow(out)) {
    out[, model := model]
    data.table::setcolorder(out, c("model", setdiff(names(out), "model")))
  }
  out[]
}

raw_grid_object <- function(model) {
  object_name <- paste0(model, "_complexity_metrics")
  out <- copy(as.data.table(drc_results_objects[[object_name]] %||% data.table()))
  if (!nrow(out)) {
    return(out)
  }

  out[
    ,
    `:=`(
      model = model,
      validation_percent_gain = 100 * mean_validation_relative_gain,
      validation_percent_se = if (
        "std_err_validation_relative_gain" %in% names(out)
      ) {
        100 * std_err_validation_relative_gain
      } else {
        NA_real_
      }
    )
  ]
  data.table::setcolorder(out, c("model", setdiff(names(out), "model")))
  out[]
}

selected_settings_object <- function(model) {
  object_name <- paste0(model, "_best_by_type")
  out <- copy(as.data.table(drc_results_objects[[object_name]] %||% data.table()))
  if (nrow(out)) {
    out[, model := model]
    data.table::setcolorder(out, c("model", setdiff(names(out), "model")))
  }
  out[]
}

tree_path_plot_dt <- path_object("tree")
forest_path_plot_dt <- path_object("forest")
path_plot_dt <- rbindlist(
  list(tree_path_plot_dt, forest_path_plot_dt),
  use.names = TRUE,
  fill = TRUE
)

tree_raw_grid_dt <- raw_grid_object("tree")
forest_raw_grid_dt <- raw_grid_object("forest")
raw_grid_dt <- rbindlist(
  list(tree_raw_grid_dt, forest_raw_grid_dt),
  use.names = TRUE,
  fill = TRUE
)

tree_selected_settings <- selected_settings_object("tree")
forest_selected_settings <- selected_settings_object("forest")
selected_settings <- rbindlist(
  list(tree_selected_settings, forest_selected_settings),
  use.names = TRUE,
  fill = TRUE
)

selected_keys <- selected_settings[, .(model, type, grid_id)]
selected_actual_dt <- raw_grid_dt[
  selected_keys,
  on = c("model", "type", "grid_id"),
  nomatch = 0
]

selected_display_cols <- intersect(
  c(
    "model", "type", "grid_id", "min_relative_gain", "validation_percent_gain",
    "validation_percent_se", "mean_validation_gain", "mean_root_objective",
    "mean_terminal_nodes", "ntree", "mtry", "minsplit", "minbucket",
    "minprob", "maxdepth", "min_gain"
  ),
  names(selected_actual_dt)
)
```

# Exact Plot Data

These are the data frames passed into `plot_validation_percent_gain_path()`.
They are already summarized by `type` and `min_relative_gain`.

```{r forest-plot-data}
show_table(
  forest_path_plot_dt,
  "Exact data frame used for the forest validation-gain path figure"
)
```

```{r tree-plot-data}
show_table(
  tree_path_plot_dt,
  "Exact data frame used for the tree validation-gain path figure"
)
```

The important columns are:

- `validation_percent_gain`: mean validation gain at this criterion and
  minimum relative split-gain value.
- `validation_percent_se`: summarized standard error used for the vertical
  bars.
- `min_validation_percent_gain` and `max_validation_percent_gain`: spread of
  raw grid settings collapsed into the same `min_relative_gain` bucket.
- `mean_terminal_nodes`: mean tree complexity at this bucket.
- `selected`: whether the selected full tuning grid row falls in this bucket.

# Selected Full-Grid Settings

The selected model is chosen from the full tuning grid, not from the collapsed
path above. This table shows the selected rows recovered from the raw complexity
metrics.

```{r selected-actual-rows}
show_table(
  selected_actual_dt[, ..selected_display_cols],
  "Selected full-grid rows with their actual validation scores"
)
```

# Option 1: Reproduce the Current Plot

This is the current helper output. The large circle marks the summarized
`min_relative_gain` bucket containing the selected full-grid setting.

```{r current-tree-plot}
plot_validation_percent_gain_path(
  tree_path_plot_dt,
  title = "Tree Validation Gain (% Root): Current Summary"
)
```

```{r current-forest-plot}
plot_validation_percent_gain_path(
  forest_path_plot_dt,
  title = "Forest Validation Gain (% Root): Current Summary"
)
```

# Option 2: Raw Grid Points Plus Mean Path

This version shows all raw tuning-grid settings as faint points, overlays the
mean path, and marks the selected full-grid row at its actual validation score.
This makes it clearer when the selected full-grid setting differs from the
highest averaged bucket.

```{r raw-plus-mean-function}
plot_raw_plus_mean_path <- function(model_name, title) {
  raw_dt <- raw_grid_dt[model == model_name]
  mean_dt <- path_plot_dt[model == model_name]
  selected_dt <- selected_actual_dt[model == model_name]

  ggplot() +
    geom_hline(
      yintercept = 0,
      linetype = 2,
      color = "gray55",
      linewidth = 0.4
    ) +
    geom_point(
      data = raw_dt,
      aes(x = min_relative_gain, y = validation_percent_gain),
      color = "gray45",
      alpha = 0.28,
      size = 1.6,
      na.rm = TRUE
    ) +
    geom_errorbar(
      data = mean_dt,
      aes(
        x = min_relative_gain,
        ymin = validation_percent_gain - validation_percent_se,
        ymax = validation_percent_gain + validation_percent_se
      ),
      width = 0,
      alpha = 0.35,
      color = "#00A6B2",
      na.rm = TRUE
    ) +
    geom_line(
      data = mean_dt,
      aes(x = min_relative_gain, y = validation_percent_gain, group = type),
      linewidth = 0.7,
      color = "#00A6B2",
      na.rm = TRUE
    ) +
    geom_point(
      data = mean_dt,
      aes(x = min_relative_gain, y = validation_percent_gain),
      size = 2,
      color = "#00A6B2",
      na.rm = TRUE
    ) +
    geom_point(
      data = selected_dt,
      aes(x = min_relative_gain, y = validation_percent_gain),
      shape = 21,
      size = 3.2,
      stroke = 0.9,
      fill = "white",
      color = "black",
      na.rm = TRUE
    ) +
    facet_wrap(ggplot2::vars(type), scales = "free_y") +
    labs(
      x = "Minimum relative split gain",
      y = "Validation gain (% root)",
      title = title,
      subtitle = "Gray points are raw grid settings; turquoise line is the grouped mean; large circle is the selected full-grid row."
    ) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank())
}
```

```{r raw-plus-mean-tree}
plot_raw_plus_mean_path(
  "tree",
  "Tree Validation Gain: Raw Grid Settings and Mean Path"
)
```

```{r raw-plus-mean-forest}
plot_raw_plus_mean_path(
  "forest",
  "Forest Validation Gain: Raw Grid Settings and Mean Path"
)
```

# Option 3: Mean Path With Raw-Grid Range

This version keeps the simple mean path but uses the min/max columns already in
the plot data frame to show how much variation was collapsed into each
`min_relative_gain` bucket.

```{r range-path-function}
plot_mean_range_path <- function(model_name, title) {
  mean_dt <- path_plot_dt[model == model_name]
  selected_bucket_dt <- mean_dt[selected == TRUE]
  selected_dt <- selected_actual_dt[model == model_name]

  ggplot(
    mean_dt,
    aes(x = min_relative_gain, y = validation_percent_gain, group = type)
  ) +
    geom_hline(
      yintercept = 0,
      linetype = 2,
      color = "gray55",
      linewidth = 0.4
    ) +
    geom_ribbon(
      aes(
        ymin = min_validation_percent_gain,
        ymax = max_validation_percent_gain
      ),
      fill = "#7A7A7A",
      alpha = 0.16,
      color = NA,
      na.rm = TRUE
    ) +
    geom_line(linewidth = 0.7, color = "#00A6B2", na.rm = TRUE) +
    geom_point(size = 2, color = "#00A6B2", na.rm = TRUE) +
    geom_point(
      data = selected_bucket_dt,
      aes(x = min_relative_gain, y = validation_percent_gain),
      inherit.aes = FALSE,
      shape = 21,
      size = 3.4,
      stroke = 0.9,
      fill = "white",
      color = "black",
      na.rm = TRUE
    ) +
    geom_point(
      data = selected_dt,
      aes(x = min_relative_gain, y = validation_percent_gain),
      inherit.aes = FALSE,
      shape = 24,
      size = 3,
      stroke = 0.8,
      fill = "#D55E00",
      color = "black",
      na.rm = TRUE
    ) +
    facet_wrap(ggplot2::vars(type), scales = "free_y") +
    labs(
      x = "Minimum relative split gain",
      y = "Validation gain (% root)",
      title = title,
      subtitle = "Band is min-max across raw grid settings; circle is selected bucket mean; triangle is selected row."
    ) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank())
}
```

```{r range-path-tree}
plot_mean_range_path(
  "tree",
  "Tree Validation Gain: Mean Path With Raw-Grid Range"
)
```

```{r range-path-forest}
plot_mean_range_path(
  "forest",
  "Forest Validation Gain: Mean Path With Raw-Grid Range"
)
```

# Option 4: Validation Gain and Complexity Tradeoff

This version removes `min_relative_gain` from the x-axis and directly shows the
tradeoff between validation performance and mean terminal-node count. The label
on the highlighted point reports the selected split-gain threshold.

```{r tradeoff-function}
plot_gain_complexity_tradeoff <- function(model_name, title) {
  mean_dt <- path_plot_dt[model == model_name]
  selected_bucket_dt <- mean_dt[selected == TRUE]

  ggplot(
    mean_dt,
    aes(
      x = mean_terminal_nodes,
      y = validation_percent_gain,
      group = type
    )
  ) +
    geom_hline(
      yintercept = 0,
      linetype = 2,
      color = "gray55",
      linewidth = 0.4
    ) +
    geom_path(
      aes(color = min_relative_gain),
      linewidth = 0.7,
      alpha = 0.75,
      na.rm = TRUE
    ) +
    geom_point(
      aes(color = min_relative_gain),
      size = 2.2,
      na.rm = TRUE
    ) +
    geom_point(
      data = selected_bucket_dt,
      aes(x = mean_terminal_nodes, y = validation_percent_gain),
      inherit.aes = FALSE,
      shape = 21,
      size = 3.4,
      stroke = 0.9,
      fill = "white",
      color = "black",
      na.rm = TRUE
    ) +
    geom_text(
      data = selected_bucket_dt,
      aes(
        x = mean_terminal_nodes,
        y = validation_percent_gain,
        label = sprintf("%.2f", min_relative_gain)
      ),
      inherit.aes = FALSE,
      nudge_y = 4,
      size = 3,
      na.rm = TRUE
    ) +
    facet_wrap(ggplot2::vars(type), scales = "free") +
    scale_color_viridis_c(option = "C", end = 0.92) +
    labs(
      x = "Mean terminal nodes",
      y = "Validation gain (% root)",
      color = "Min relative\nsplit gain",
      title = title
    ) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank())
}
```

```{r tradeoff-tree}
plot_gain_complexity_tradeoff(
  "tree",
  "Tree Validation Gain and Complexity Tradeoff"
)
```

```{r tradeoff-forest}
plot_gain_complexity_tradeoff(
  "forest",
  "Forest Validation Gain and Complexity Tradeoff"
)
```

# Option 5: Fixed Forest Settings

This section filters the raw forest tuning grid before summarising. It is useful
when we want to ask a narrower question such as:

> If the CI forest uses 50 trees, `mtry = 15`, and
> `min_relative_gain = 0.3`, what are the mean train/validation gains and
> complexity for that fixed setting?

This avoids mixing that setting with other forest settings that happen to share
the same `min_relative_gain`.

```{r fixed-forest-functions}
filter_equal <- function(dt, column, value) {
  if (!column %in% names(dt) || is.null(value) || length(value) == 0L) {
    return(dt)
  }

  if (is.numeric(dt[[column]]) && is.numeric(value)) {
    tolerance <- sqrt(.Machine$double.eps)
    return(dt[is.finite(get(column)) & abs(get(column) - value[1L]) < tolerance])
  }

  dt[get(column) == value[1L]]
}

filter_fixed_forest <- function(type_value, fixed = list()) {
  out <- copy(forest_raw_grid_dt[type == type_value])
  if (!nrow(out)) {
    return(out)
  }

  for (column in names(fixed)) {
    out <- filter_equal(out, column, fixed[[column]])
  }

  out[]
}

selected_full_forest_rows <- function(type_values = NULL) {
  keys <- copy(forest_selected_settings)
  if (!is.null(type_values)) {
    keys <- keys[type %in% type_values]
  }
  if (!nrow(keys)) {
    return(data.table())
  }

  selected_cols <- intersect(
    c(
      "type", "ntree", "mtry", "minsplit", "minbucket", "minprob",
      "maxdepth", "min_gain", "min_relative_gain"
    ),
    names(keys)
  )

  rows <- lapply(seq_len(nrow(keys)), function(i) {
    key <- keys[i]
    out <- copy(forest_raw_grid_dt[type == key$type])
    for (column in setdiff(selected_cols, "type")) {
      out <- filter_equal(out, column, key[[column]])
    }
    out[]
  })

  rbindlist(rows, use.names = TRUE, fill = TRUE)
}

summarise_fixed_forest <- function(dt, label, group_cols = NULL) {
  if (!nrow(dt)) {
    return(data.table())
  }

  if (is.null(group_cols)) {
    group_cols <- c("model", "type")
  }
  group_cols <- intersect(group_cols, names(dt))

  dt[
    ,
    .(
      label = label,
      n_grid_rows = .N,
      n_grid_ids = uniqueN(grid_id),
      train_gain_percent = mean(100 * mean_train_relative_gain, na.rm = TRUE),
      validation_gain_percent = mean(validation_percent_gain, na.rm = TRUE),
      validation_se_percent = {
        se <- validation_percent_se[is.finite(validation_percent_se)]
        if (length(se)) sqrt(sum(se^2)) / length(se) else NA_real_
      },
      terminal_nodes = mean(mean_terminal_nodes, na.rm = TRUE),
      mean_root_objective = mean(mean_root_objective, na.rm = TRUE)
    ),
    by = group_cols
  ][]
}

fixed_summary_long <- function(summary_dt) {
  if (!nrow(summary_dt)) {
    return(data.table())
  }

  measure_cols <- intersect(
    c(
      "train_gain_percent", "validation_gain_percent",
      "validation_se_percent", "terminal_nodes"
    ),
    names(summary_dt)
  )

  melt(
    summary_dt,
    id.vars = setdiff(names(summary_dt), measure_cols),
    measure.vars = measure_cols,
    variable.name = "metric",
    value.name = "value"
  )[
    ,
    metric_label := fifelse(
      metric == "train_gain_percent",
      "Train gain (% root)",
      fifelse(
        metric == "validation_gain_percent",
        "Validation gain (% root)",
        fifelse(
          metric == "validation_se_percent",
          "Validation SE (% root)",
          "Mean terminal nodes"
        )
      )
    )
  ][]
}

plot_fixed_forest_summary <- function(summary_dt, title) {
  long_dt <- fixed_summary_long(summary_dt)
  if (!nrow(long_dt)) {
    return(
      ggplot() +
        annotate("text", x = 0, y = 0, label = "No matching forest rows") +
        labs(title = title, x = NULL, y = NULL) +
        theme_void(base_size = 11)
    )
  }

  ggplot(
    long_dt,
    aes(x = type, y = value, fill = metric_label)
  ) +
    geom_hline(
      data = long_dt[grepl("gain", metric_label, ignore.case = TRUE)],
      inherit.aes = FALSE,
      yintercept = 0,
      linetype = 2,
      color = "gray55",
      linewidth = 0.35
    ) +
    geom_col(width = 0.62, color = "gray30", linewidth = 0.2) +
    geom_text(
      aes(label = sprintf("%.2f", value)),
      vjust = ifelse(long_dt$value >= 0, -0.35, 1.25),
      size = 3,
      na.rm = TRUE
    ) +
    facet_wrap(ggplot2::vars(metric_label), scales = "free_y") +
    scale_fill_manual(
      values = c(
        "Train gain (% root)" = "#3268A8",
        "Validation gain (% root)" = "#00A6B2",
        "Validation SE (% root)" = "#8A8A8A",
        "Mean terminal nodes" = "#D55E00"
      ),
      guide = "none"
    ) +
    labs(
      x = NULL,
      y = NULL,
      title = title
    ) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank())
}
```

## CI Forest: 50 Trees, `mtry = 15`, `min_relative_gain = 0.3`

This is the specific fixed setting described above. If several grid rows match
these three fixed values because other controls differ, the table and plot show
the mean over those matching rows.

```{r ci-forest-fixed-setting}
ci_forest_fixed_rows <- filter_fixed_forest(
  type_value = "CI",
  fixed = list(
    ntree = 50L,
    mtry = 15L,
    min_relative_gain = 0.3
  )
)

ci_forest_fixed_summary <- summarise_fixed_forest(
  ci_forest_fixed_rows,
  label = "CI forest: ntree 50, mtry 15, min relative gain 0.3",
  group_cols = c("model", "type", "ntree", "mtry", "min_relative_gain")
)

fixed_row_display_cols <- intersect(
  c(
    "model", "type", "grid_id", "ntree", "mtry", "minsplit", "minbucket",
    "minprob", "maxdepth", "min_gain", "min_relative_gain",
    "validation_percent_gain", "validation_percent_se",
    "mean_train_relative_gain", "mean_validation_relative_gain",
    "mean_terminal_nodes"
  ),
  names(ci_forest_fixed_rows)
)

show_table(
  ci_forest_fixed_rows[, ..fixed_row_display_cols],
  "Forest grid rows included in the fixed CI setting"
)
```

```{r ci-forest-fixed-summary-table}
show_table(
  ci_forest_fixed_summary,
  "Mean metrics for the fixed CI forest setting"
)
```

```{r ci-forest-fixed-summary-plot}
plot_fixed_forest_summary(
  ci_forest_fixed_summary,
  "Fixed CI Forest Setting: Mean Metrics"
)
```

## Selected Full Forest Settings by Criterion

This version fixes the complete selected forest setting for each criterion,
using all split-control columns available in `forest_best_by_type`. It answers:

> What does each selected forest look like at its exact selected full-grid
> configuration?

```{r selected-full-forest-settings}
selected_full_forest_dt <- selected_full_forest_rows()
selected_full_forest_summary <- summarise_fixed_forest(
  selected_full_forest_dt,
  label = "Selected full forest setting",
  group_cols = c(
    "model", "type", "ntree", "mtry", "minsplit", "minbucket",
    "minprob", "maxdepth", "min_gain", "min_relative_gain"
  )
)

show_table(
  selected_full_forest_summary,
  "Mean metrics for the selected full forest settings"
)
```

```{r selected-full-forest-settings-plot}
plot_fixed_forest_summary(
  selected_full_forest_summary,
  "Selected Full Forest Settings: Mean Metrics"
)
```

# Option 6: Fixed Forest Controls, Vary `min_relative_gain`

This is the least confusing way to show the selected minimum relative split-gain
value. For each criterion, we hold the selected forest controls fixed, except
for `min_relative_gain`, and then plot the validation path over
`min_relative_gain`.

For example, if the selected CI forest used `ntree = 50`, `mtry = 15`,
`minsplit = 200`, `minbucket = 100`, and `maxdepth = 10`, this plot compares
only CI forests with those controls fixed while `min_relative_gain` varies. The
highlighted point is then selected from the same conditional path being plotted.

```{r fixed-control-min-gain-path-functions}
selected_forest_fixed_controls <- function(type_value, extra_fixed = list()) {
  selected_row <- forest_selected_settings[type == type_value]
  if (!nrow(selected_row)) {
    return(list())
  }
  selected_row <- copy(selected_row[1L])

  control_cols <- intersect(
    c(
      "ntree", "mtry", "minsplit", "minbucket", "minprob",
      "maxdepth", "min_gain"
    ),
    names(selected_row)
  )

  fixed <- as.list(selected_row[, ..control_cols])
  fixed <- fixed[!vapply(fixed, function(x) length(x) == 0L || is.na(x[1L]), logical(1))]
  modifyList(fixed, extra_fixed)
}

fixed_control_min_gain_path <- function(type_value, fixed_controls, label = NULL) {
  out <- copy(forest_raw_grid_dt[type == type_value])
  if (!nrow(out)) {
    return(data.table())
  }

  for (column in names(fixed_controls)) {
    out <- filter_equal(out, column, fixed_controls[[column]])
  }
  if (!nrow(out)) {
    return(data.table())
  }

  selected_row <- forest_selected_settings[type == type_value]
  selected_min_gain <- if (nrow(selected_row)) {
    selected_row$min_relative_gain[1L]
  } else {
    NA_real_
  }

  out[
    ,
    .(
      label = label %||% paste(type_value, "selected controls"),
      n_grid_rows = .N,
      n_grid_ids = uniqueN(grid_id),
      train_gain_percent = mean(100 * mean_train_relative_gain, na.rm = TRUE),
      validation_gain_percent = mean(validation_percent_gain, na.rm = TRUE),
      validation_se_percent = {
        se <- validation_percent_se[is.finite(validation_percent_se)]
        if (length(se)) sqrt(sum(se^2)) / length(se) else NA_real_
      },
      min_validation_gain_percent = min(validation_percent_gain, na.rm = TRUE),
      max_validation_gain_percent = max(validation_percent_gain, na.rm = TRUE),
      terminal_nodes = mean(mean_terminal_nodes, na.rm = TRUE),
      selected = isTRUE(
        is.finite(selected_min_gain) &&
          abs(min_relative_gain[1L] - selected_min_gain) < sqrt(.Machine$double.eps)
      )
    ),
    by = .(type, min_relative_gain)
  ][order(type, min_relative_gain)]
}

selected_controls_min_gain_paths <- function(type_values = NULL) {
  if (is.null(type_values)) {
    type_values <- unique(forest_selected_settings$type)
  }
  if (!length(type_values)) {
    return(data.table())
  }

  paths <- lapply(type_values, function(type_value) {
    fixed_controls <- selected_forest_fixed_controls(type_value)
    fixed_control_min_gain_path(
      type_value = type_value,
      fixed_controls = fixed_controls,
      label = paste(type_value, "selected controls")
    )
  })

  rbindlist(paths, use.names = TRUE, fill = TRUE)
}

plot_fixed_control_min_gain_path <- function(path_dt, title) {
  if (!nrow(path_dt)) {
    return(
      ggplot() +
        annotate("text", x = 0, y = 0, label = "No matching forest path") +
        labs(title = title, x = NULL, y = NULL) +
        theme_void(base_size = 11)
    )
  }

  selected_dt <- path_dt[selected == TRUE]

  ggplot(
    path_dt,
    aes(x = min_relative_gain, y = validation_gain_percent, group = type)
  ) +
    geom_hline(
      yintercept = 0,
      linetype = 2,
      color = "gray55",
      linewidth = 0.4
    ) +
    geom_ribbon(
      aes(
        ymin = min_validation_gain_percent,
        ymax = max_validation_gain_percent
      ),
      fill = "#7A7A7A",
      alpha = 0.14,
      color = NA,
      na.rm = TRUE
    ) +
    geom_errorbar(
      aes(
        ymin = validation_gain_percent - validation_se_percent,
        ymax = validation_gain_percent + validation_se_percent
      ),
      width = 0,
      alpha = 0.35,
      color = "#00A6B2",
      na.rm = TRUE
    ) +
    geom_line(linewidth = 0.7, color = "#00A6B2", na.rm = TRUE) +
    geom_point(size = 2, color = "#00A6B2", na.rm = TRUE) +
    geom_point(
      data = selected_dt,
      aes(x = min_relative_gain, y = validation_gain_percent),
      inherit.aes = FALSE,
      shape = 21,
      size = 3.4,
      stroke = 0.9,
      fill = "white",
      color = "black",
      na.rm = TRUE
    ) +
    facet_wrap(ggplot2::vars(type), scales = "free_y") +
    labs(
      x = "Minimum relative split gain",
      y = "Validation gain (% root)",
      title = title,
      subtitle = "Forest controls are fixed at the selected setting; only min_relative_gain varies."
    ) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank())
}
```

## CI Forest Path Conditional on the Selected Setting

This view takes the selected CI forest setting, removes only
`min_relative_gain` from the fixed controls, and then plots the path over
`min_relative_gain`.

```{r ci-fixed-controls-min-gain-path}
ci_selected_fixed_controls <- selected_forest_fixed_controls("CI")

ci_fixed_controls_path <- fixed_control_min_gain_path(
  type_value = "CI",
  fixed_controls = ci_selected_fixed_controls,
  label = "CI forest: selected controls except min_relative_gain"
)

show_table(
  as.data.table(ci_selected_fixed_controls),
  "CI forest controls held fixed"
)

show_table(
  ci_fixed_controls_path,
  "CI forest path conditional on selected controls except min_relative_gain"
)
```

```{r ci-fixed-controls-min-gain-path-plot}
plot_fixed_control_min_gain_path(
  ci_fixed_controls_path,
  "CI Forest Validation Path Conditional on Selected Controls"
)
```

## Selected-Control Forest Paths by Criterion

This version uses each criterion's selected forest controls from
`forest_best_by_type`, but deliberately excludes `min_relative_gain` from the
fixed controls so it can form the x-axis path.

```{r selected-controls-min-gain-path}
selected_controls_path_dt <- selected_controls_min_gain_paths()

show_table(
  selected_controls_path_dt,
  "Forest validation paths with selected controls fixed and min_relative_gain varying"
)
```

```{r selected-controls-min-gain-path-plot}
plot_fixed_control_min_gain_path(
  selected_controls_path_dt,
  "Forest Validation Path With Selected Controls Fixed"
)
```

# Compact Selection Summary

This table is a thesis-friendly companion to any of the plots above.

```{r compact-selection-summary}
compact_selection <- selected_actual_dt[
  ,
  .(
    model,
    type,
    grid_id,
    min_relative_gain,
    validation_gain_percent = validation_percent_gain,
    validation_se_percent = validation_percent_se,
    mean_terminal_nodes,
    ntree = if ("ntree" %in% names(selected_actual_dt)) ntree else NA_integer_,
    mtry = if ("mtry" %in% names(selected_actual_dt)) mtry else NA_integer_,
    minsplit = if ("minsplit" %in% names(selected_actual_dt)) minsplit else NA_integer_,
    minbucket = if ("minbucket" %in% names(selected_actual_dt)) minbucket else NA_integer_
  )
]

show_table(
  compact_selection,
  "Compact selected-setting summary"
)
```

# Suggested Reading

For the thesis text, the clearest wording is:

> Points show mean validation gain across tuning settings with the same minimum
> relative split-gain value. The highlighted marker indicates the selected full
> tuning configuration, which was selected from the complete tuning grid rather
> than from this one-dimensional summary alone.
