---
title: "Test svyglm Margins Rineq Decomposition"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
params:
  analysis_rows: 1
---

```{r setup}
library(data.table)
library(haven)
library(survey)
library(rineq)
library(margins)
library(here)

source(here("R", "report_helpers.R"))
```

# Aim

This notebook tests whether survey-weighted average marginal effects can be
extracted from the existing `survey::svyglm()` model using `margins::margins()`
and then passed into Rineq's lower-level decomposition function.

It does not run the tree, forest, SHAP, simulation, or final saving sections.

# Prepare Data

```{r runtime-config}
runtime_params <- list(
  analysis_rows = as.integer(params$analysis_rows)
)

if (is.na(runtime_params$analysis_rows) || runtime_params$analysis_rows <= 0L) {
  runtime_params$analysis_rows <- 1L
}

source_data_file <- here("prev_analysis", "DRCongo .dta")
stopifnot(file.exists(source_data_file))
```

```{r load-2007-drc-data}
drc_2007_dt <- as.data.table(haven::read_dta(source_data_file))
```

```{r prepare-2007-drc-analysis-data}
factor_from_codes <- function(x, labels) {
  factor(labels[as.character(as.integer(x))], levels = unname(labels))
}

binary_factor <- function(x, labels) {
  factor(labels[as.character(as.integer(x))], levels = unname(labels))
}

drc_birth_labels <- c(
  "1" = "First birth",
  "2" = "2-4 short interval",
  "3" = "2-4 long interval",
  "4" = "5+ short interval",
  "5" = "5+ long interval"
)
drc_wealth_group_labels <- c(
  "1" = "High household wealth",
  "2" = "Low household wealth"
)
drc_age_mother_labels <- c("1" = "20 or more", "2" = "Less than 20")
drc_education_labels <- c("1" = "Any education", "2" = "No education")
drc_occupation_labels <- c(
  "1" = "Other",
  "2" = "Household, unskilled, not working",
  "3" = "Agriculture"
)
drc_region_labels <- c(
  "1" = "Bandundu",
  "2" = "Bas-Congo",
  "3" = "Equateur",
  "4" = "Kasai Occidental",
  "5" = "Kasai Oriental",
  "6" = "Katanga",
  "7" = "Kinshasa",
  "8" = "Maniema",
  "9" = "Nord-Kivu",
  "10" = "Orientale",
  "11" = "Sud-Kivu"
)

raw_congo_model_dt <- drc_2007_dt[
  ,
  .(
    wealth = as.numeric(wealth),
    deadu5_num = as.integer(deadu5),
    sample_weight = as.numeric(weight),
    PSU = as.integer(PSU),
    household = as.integer(household),
    country = as.integer(country),
    quint = factor_from_codes(quint, drc_wealth_group_labels),
    unskilled = binary_factor(
      unskilled,
      c("0" = "Skilled birth attendance", "1" = "No skilled birth attendance")
    ),
    male = binary_factor(male, c("0" = "Female", "1" = "Male")),
    birth = factor_from_codes(birth, drc_birth_labels),
    agemoth = factor_from_codes(agemoth, drc_age_mother_labels),
    rural = binary_factor(rural, c("0" = "Urban", "1" = "Rural")),
    ed = factor_from_codes(ed, drc_education_labels),
    ped = factor_from_codes(ped, drc_education_labels),
    mocc = factor_from_codes(mocc, drc_occupation_labels),
    pocc = factor_from_codes(pocc, drc_occupation_labels),
    reg = factor_from_codes(reg, drc_region_labels)
  )
]
raw_congo_model_dt <- na.omit(raw_congo_model_dt)

raw_congo_var_labels <- c(
  quint = "Household wealth",
  unskilled = "Skilled birth attendance",
  male = "Sex of the child",
  birth = "Birth order and interval",
  agemoth = "Mother's age at birth",
  rural = "Type of residence",
  ed = "Mother's education",
  ped = "Father's education",
  mocc = "Mother's occupation",
  pocc = "Father's occupation",
  reg = "Region"
)
raw_congo_predictors <- names(raw_congo_var_labels)
```

```{r sample-2007-drc-analysis-data}
raw_congo_model_dt <- sample_analysis_rows(
  raw_congo_model_dt,
  use_full = runtime_params$analysis_rows == 1L,
  n = runtime_params$analysis_rows,
  seed = 20260516
)

data.table(
  rows = nrow(raw_congo_model_dt),
  deaths = sum(raw_congo_model_dt$deadu5_num == 1L),
  non_deaths = sum(raw_congo_model_dt$deadu5_num == 0L),
  psus = uniqueN(raw_congo_model_dt$PSU)
)
```

# Fit Survey-Weighted Regression

```{r fit-classical-model}
linear_decomp_formula <- as.formula(
  paste("deadu5_num ~", paste(raw_congo_predictors, collapse = " + "))
)

linear_decomp_design <- survey::svydesign(
  id = as.formula("~PSU+household"),
  weights = as.formula("~sample_weight"),
  data = raw_congo_model_dt
)

linear_decomp_fit <- survey::svyglm(
  linear_decomp_formula,
  design = linear_decomp_design,
  family = quasibinomial()
)

cat("svyglm fitted: yes\n")
```

# Test Margins

```{r create-model-matrix}
linear_x <- model.matrix(linear_decomp_formula, data = raw_congo_model_dt)
linear_terms <- setdiff(colnames(linear_x), "(Intercept)")
```

```{r run-margins}
linear_margins <- tryCatch(
  margins::margins(
    linear_decomp_fit,
    design = linear_decomp_design,
    type = "response"
  ),
  error = function(e) {
    stop(
      "margins::margins() does not work with the svyglm object: ",
      conditionMessage(e),
      call. = FALSE
    )
  }
)

linear_ame_table <- as.data.table(summary(linear_margins))
print(linear_ame_table)
```

```{r inspect-margins-columns}
print(names(linear_ame_table))

ame_col <- intersect(c("AME", "dydx"), names(linear_ame_table))[1L]

if (is.na(ame_col)) {
  stop("Could not identify the average marginal effect column in margins output.")
}

if (!"factor" %in% names(linear_ame_table)) {
  stop("margins output does not contain a factor column for term alignment.")
}
```

# Check Term Alignment

```{r inspect-term-alignment}
print(linear_terms)
print(linear_ame_table$factor)

linear_average_marginal_effects <- setNames(
  linear_ame_table[[ame_col]],
  linear_ame_table$factor
)

missing_from_margins <- setdiff(
  linear_terms,
  names(linear_average_marginal_effects)
)
extra_from_margins <- setdiff(
  names(linear_average_marginal_effects),
  linear_terms
)

cat("\nTerms missing from margins output:\n")
print(missing_from_margins)

cat("\nExtra terms from margins output:\n")
print(extra_from_margins)

if (length(missing_from_margins) > 0L) {
  stop(
    "AME names from margins do not match model-matrix terms. ",
    "Need a term-mapping step before using these in Rineq.",
    call. = FALSE
  )
}
```

# Test Rineq Lower-Level Decomposition

```{r check-rineq-decomposition-signature}
rineq_decomposition <- getFromNamespace("decomposition", "rineq")
expected_decomposition_args <- c(
  "outcome",
  "betas",
  "mm",
  "ranker",
  "wt",
  "correction",
  "citype"
)
actual_decomposition_args <- names(formals(rineq_decomposition))

if (!all(expected_decomposition_args %in% actual_decomposition_args)) {
  stop(
    "Rineq's internal decomposition() function has different argument names. ",
    "Expected: ",
    paste(expected_decomposition_args, collapse = ", "),
    ". Actual: ",
    paste(actual_decomposition_args, collapse = ", "),
    call. = FALSE
  )
}
```

```{r run-ci-decomposition}
test_ci_decomp <- tryCatch(
  rineq_decomposition(
    outcome = raw_congo_model_dt$deadu5_num,
    betas = linear_average_marginal_effects[linear_terms],
    mm = linear_x[, linear_terms, drop = FALSE],
    ranker = raw_congo_model_dt$wealth,
    wt = raw_congo_model_dt$sample_weight,
    correction = TRUE,
    citype = "CI"
  ),
  error = function(e) {
    stop(
      "Rineq lower-level decomposition failed for CI: ",
      conditionMessage(e),
      call. = FALSE
    )
  }
)

print(test_ci_decomp)
```

```{r run-bounded-decompositions}
test_cig_decomp <- tryCatch(
  rineq_decomposition(
    outcome = raw_congo_model_dt$deadu5_num,
    betas = linear_average_marginal_effects[linear_terms],
    mm = linear_x[, linear_terms, drop = FALSE],
    ranker = raw_congo_model_dt$wealth,
    wt = raw_congo_model_dt$sample_weight,
    correction = TRUE,
    citype = "CIg"
  ),
  error = function(e) {
    stop("Rineq lower-level decomposition failed for CIg: ", conditionMessage(e), call. = FALSE)
  }
)

test_cic_decomp <- tryCatch(
  rineq_decomposition(
    outcome = raw_congo_model_dt$deadu5_num,
    betas = linear_average_marginal_effects[linear_terms],
    mm = linear_x[, linear_terms, drop = FALSE],
    ranker = raw_congo_model_dt$wealth,
    wt = raw_congo_model_dt$sample_weight,
    correction = TRUE,
    citype = "CIc"
  ),
  error = function(e) {
    stop("Rineq lower-level decomposition failed for CIc: ", conditionMessage(e), call. = FALSE)
  }
)

test_ciw_decomp <- tryCatch(
  rineq_decomposition(
    outcome = raw_congo_model_dt$deadu5_num,
    betas = linear_average_marginal_effects[linear_terms],
    mm = linear_x[, linear_terms, drop = FALSE],
    ranker = raw_congo_model_dt$wealth,
    wt = raw_congo_model_dt$sample_weight,
    correction = TRUE,
    citype = "CIw"
  ),
  error = function(e) {
    stop("Rineq lower-level decomposition failed for CIw: ", conditionMessage(e), call. = FALSE)
  }
)

print(test_cig_decomp)
print(test_cic_decomp)
print(test_ciw_decomp)
```

# Compare With Old Coefficient-Based Route

```{r compare-with-old-rineq-route}
old_ci_decomp <- rineq::contribution(
  linear_decomp_fit,
  raw_congo_model_dt$wealth,
  correction = TRUE,
  type = "CI"
)

cat("\nOld coefficient-based CI decomposition:\n")
print(old_ci_decomp)

cat("\nNew margins-based CI decomposition:\n")
print(test_ci_decomp)
```

# Diagnostic Summary

```{r diagnostic-summary}
cat("\nTEST SUMMARY\n")
cat("1. svyglm fitted: yes\n")
cat("2. margins::margins() ran on svyglm with design: yes\n")
cat(
  "3. AME terms matched model-matrix terms: ",
  length(missing_from_margins) == 0L,
  "\n"
)
cat("4. Rineq lower-level decomposition ran for CI: ", exists("test_ci_decomp"), "\n")
cat(
  "5. Rineq lower-level decomposition ran for CIg, CIc, CIw: ",
  exists("test_cig_decomp") &&
    exists("test_cic_decomp") &&
    exists("test_ciw_decomp"),
  "\n"
)
```
