---
title: "ggparty Tree Plot Experiments"
author: "Moses Mburu"
format:
  html:
    toc: true
    number-sections: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
knitr::opts_chunk$set(
  fig.width = 12,
  fig.height = 6,
  dpi = 300,
  warning = FALSE,
  message = FALSE
)

library(data.table)
library(ggplot2)
library(ggparty)
library(here)

source(here("R", "report_helpers.R"))

results_object_file <- here("data", "drc_report_results_objects.rda")
report_results <- load_drc_report_results(results_object_file)
drc_results_objects <- report_results$drc_results_objects

congo_model_dt <- drc_results_objects$congo_model_dt
congo_var_labels <- drc_results_objects$congo_var_labels
selected_tree_type <- drc_results_objects$selected_tree_type
selected_tree_fit <- drc_results_objects$selected_tree_fit
selected_forest_surrogate <- drc_results_objects$selected_forest_surrogate
forest_surrogates <- drc_results_objects$forest_surrogates_by_type %||% list()

repel_mode <- if (requireNamespace("ggrepel", quietly = TRUE)) {
  "terminal"
} else {
  "none"
}
font_11 <- 11 / ggplot2::.pt

data.table(
  object_file = results_object_file,
  selected_tree_type = selected_tree_type,
  ggrepel_available = requireNamespace("ggrepel", quietly = TRUE),
  repel_mode_used = repel_mode
)
```

# Purpose

This notebook tests ggplot-based tree rendering from the saved DRC results
object. The existing `ci_report_tree_plot()` uses the package/base plotting
method. The experimental `ci_tree_gg()` helper uses `ggparty`, which means the
tree can be styled with ordinary `ggplot2` layers and can optionally use
`ggrepel::geom_label_repel()` when `ggrepel` is installed.

# Selected CI Tree

## Current Base/partykit Plot

```{r selected-tree-base, fig.width=12, fig.height=6}
ci_report_tree_plot(
  fit = selected_tree_fit,
  data = congo_model_dt,
  outcome_name = "deadu5_num",
  outcome_label = "U5",
  ci_type = selected_tree_type,
  var_labels = congo_var_labels,
  fontsize = 11
)
```

## ggparty Plot

```{r selected-tree-ggparty, fig.width=12, fig.height=6}
ci_tree_gg(
  fit = selected_tree_fit,
  data = congo_model_dt,
  outcome_name = "deadu5_num",
  outcome_label = "U5",
  ci_type = selected_tree_type,
  var_labels = congo_var_labels,
  inner_text_size = 3.2,
  terminal_text_size = 2.8,
  edge_text_size = 2.8,
  split_label_width = 24,
  edge_label_width = 20,
  terminal_label_width = 16,
  use_repel = repel_mode
)
```

## ggparty Wider Labels

```{r selected-tree-ggparty-wide-labels, fig.width=13, fig.height=6.5}
ci_tree_gg(
  fit = selected_tree_fit,
  data = congo_model_dt,
  outcome_name = "deadu5_num",
  outcome_label = "U5",
  ci_type = selected_tree_type,
  var_labels = congo_var_labels,
  inner_text_size = 3.0,
  terminal_text_size = 2.6,
  edge_text_size = 2.6,
  split_label_width = 30,
  edge_label_width = 26,
  terminal_label_width = 20,
  terminal_space = 0.35,
  use_repel = repel_mode
)
```

## ggparty Partykit-Like Layout

This variant uses true node depth from the `partykit` object, so terminal nodes
are not forced into a common bottom band. The selected single tree has leaves at
one common depth, but this setting matters for the forest surrogate trees below.

```{r selected-tree-ggparty-level-layout, fig.width=9.5, fig.height=6.5}
ci_tree_gg(
  fit = selected_tree_fit,
  data = congo_model_dt,
  outcome_name = "deadu5_num",
  outcome_label = "U5",
  ci_type = selected_tree_type,
  var_labels = congo_var_labels,
  inner_text_size = font_11,
  terminal_text_size = font_11,
  edge_text_size = font_11,
  split_label_width = 28,
  edge_label_width = 24,
  terminal_label_width = 18,
  terminal_layout = "level",
  node_box_style = "ci_report_tree",
  edge_label_style = "text",
  use_repel = repel_mode
)
```

# Selected Forest Surrogate

```{r selected-forest-surrogate-ggparty, fig.width=12, fig.height=6}
ci_tree_gg(
  fit = selected_forest_surrogate$fit,
  data = selected_forest_surrogate$data,
  outcome_name = selected_forest_surrogate$prediction_name,
  outcome_label = "U5",
  ci_type = selected_tree_type,
  var_labels = congo_var_labels,
  inner_text_size = 3.1,
  terminal_text_size = 2.7,
  edge_text_size = 2.7,
  split_label_width = 26,
  edge_label_width = 22,
  terminal_label_width = 17,
  terminal_space = 0.35,
  use_repel = repel_mode
)
```

## Selected Forest Surrogate: Partykit-Like Layout

```{r selected-forest-surrogate-ggparty-level-layout, fig.width=9.5, fig.height=6.5}
ci_tree_gg(
  fit = selected_forest_surrogate$fit,
  data = selected_forest_surrogate$data,
  outcome_name = selected_forest_surrogate$prediction_name,
  outcome_label = "U5",
  ci_type = selected_tree_type,
  var_labels = congo_var_labels,
  inner_text_size = font_11,
  terminal_text_size = font_11,
  edge_text_size = font_11,
  split_label_width = 28,
  edge_label_width = 24,
  terminal_label_width = 18,
  terminal_layout = "level",
  node_box_style = "ci_report_tree",
  edge_label_style = "text",
  use_repel = repel_mode
)
```

```{r terminal-depth-diagnostic}
terminal_depths <- function(fit, label) {
  p <- ci_tree_gg(
    fit = fit,
    data = selected_forest_surrogate$data,
    outcome_name = selected_forest_surrogate$prediction_name,
    outcome_label = "U5",
    ci_type = selected_tree_type,
    var_labels = congo_var_labels,
    terminal_layout = "level"
  )
  depth_dt <- as.data.table(p$data)[kids == 0L]
  depth_dt[, plot := label]
  depth_dt[, .N, by = .(plot, true_level)][
    order(plot, true_level)
  ]
}

terminal_depths(
  selected_forest_surrogate$fit,
  "Selected forest surrogate"
)
```

# Criterion-Specific Forest Surrogates

These chunks render each saved surrogate from `forest_surrogates_by_type` using
the same ggparty helper.

```{r forest-surrogate-ci, fig.width=12, fig.height=6}
if ("CI" %in% names(forest_surrogates)) {
  surrogate <- forest_surrogates[["CI"]]
  ci_tree_gg(
    fit = surrogate$fit,
    data = surrogate$data,
    outcome_name = surrogate$prediction_name,
    outcome_label = "U5",
    ci_type = "CI",
    var_labels = congo_var_labels,
    inner_text_size = font_11,
    terminal_text_size = font_11,
    edge_text_size = font_11,
    terminal_layout = "level",
    node_box_style = "ci_report_tree",
    edge_label_style = "text",
    use_repel = repel_mode
  )
}
```

```{r forest-surrogate-cig, fig.width=12, fig.height=6}
if ("CIg" %in% names(forest_surrogates)) {
  surrogate <- forest_surrogates[["CIg"]]
  ci_tree_gg(
    fit = surrogate$fit,
    data = surrogate$data,
    outcome_name = surrogate$prediction_name,
    outcome_label = "U5",
    ci_type = "CIg",
    var_labels = congo_var_labels,
    inner_text_size = font_11,
    terminal_text_size = font_11,
    edge_text_size = font_11,
    terminal_layout = "level",
    node_box_style = "ci_report_tree",
    edge_label_style = "text",
    use_repel = repel_mode
  )
}
```

```{r forest-surrogate-cic, fig.width=12, fig.height=6}
if ("CIc" %in% names(forest_surrogates)) {
  surrogate <- forest_surrogates[["CIc"]]
  ci_tree_gg(
    fit = surrogate$fit,
    data = surrogate$data,
    outcome_name = surrogate$prediction_name,
    outcome_label = "U5",
    ci_type = "CIc",
    var_labels = congo_var_labels,
    inner_text_size = font_11,
    terminal_text_size = font_11,
    edge_text_size = font_11,
    terminal_layout = "level",
    node_box_style = "ci_report_tree",
    edge_label_style = "text",
    use_repel = repel_mode
  )
}
```

```{r forest-surrogate-l, fig.width=12, fig.height=6}
if ("L" %in% names(forest_surrogates)) {
  surrogate <- forest_surrogates[["L"]]
  ci_tree_gg(
    fit = surrogate$fit,
    data = surrogate$data,
    outcome_name = surrogate$prediction_name,
    outcome_label = "U5",
    ci_type = "L",
    var_labels = congo_var_labels,
    inner_text_size = font_11,
    terminal_text_size = font_11,
    edge_text_size = font_11,
    terminal_layout = "level",
    node_box_style = "ci_report_tree",
    edge_label_style = "text",
    use_repel = repel_mode
  )
}
```

# Notes

The ggparty layout uses the tree's node structure but does not promise pixel
identity with the base `partykit` plot. Its advantage is that the returned
object is a normal `ggplot`, so labels, margins, text sizes, fills, and optional
repelling can be tuned in the usual ggplot ecosystem.
